---
title: "Solution: PSA for the Breast Cancer Model"
subtitle: "Complete analysis with interpretations"
format:
  html:
    toc: true
    code-fold: show
---

## Setup and functions

Load libraries and define helper functions and the PSM model function.

```{r}
#| message: false
library(tidyverse)
library(ggplot2)
library(gridExtra)

theme_set(theme_minimal() +
          theme(panel.grid.major = element_line(colour = "grey90"),
                panel.grid.minor = element_blank()))

# Helper function: Convert mean and SD to gamma parameters
mean_sd_to_gamma <- function(mean, sd) {
  rate <- mean / (sd^2)
  shape <- mean * rate
  return(list(shape = shape, rate = rate))
}

# Helper function: Convert mean and SD to beta parameters
mean_sd_to_beta <- function(mean, sd) {
  if (mean <= 0 | mean >= 1) {
    stop("Beta mean must be strictly between 0 and 1")
  }
  alpha <- ((1 - mean) / (sd^2) - 1 / mean) * mean^2
  beta <- alpha * (1 / mean - 1)
  return(list(alpha = alpha, beta = beta))
}

# PSM function
psm_model <- function(
    hr_os = 0.70,
    hr_pfs = 0.50,
    os_lambda_ctrl = 0.045,
    os_gamma_ctrl = 1.15,
    pfs_lambda_ctrl = 0.12,
    pfs_gamma_ctrl = 1.05,
    cost_pf_trast = 420000,
    cost_pd = 180000,
    utility_pf = 0.80,
    utility_pd = 0.55,
    cost_trast_maint = 25000,
    time_horizon = 20,
    discount_rate = 0.03
) {

  times <- seq(0, time_horizon, by = 1)

  S_os_ctrl <- function(t) exp(-os_lambda_ctrl * t^os_gamma_ctrl)
  S_pfs_ctrl <- function(t) exp(-pfs_lambda_ctrl * t^pfs_gamma_ctrl)
  S_os_trast <- function(t) exp(-os_lambda_ctrl * (hr_os * t)^os_gamma_ctrl)
  S_pfs_trast <- function(t) exp(-pfs_lambda_ctrl * (hr_pfs * t)^pfs_gamma_ctrl)

  df <- tibble(
    time = times,
    S_os_ctrl_t = sapply(times, S_os_ctrl),
    S_pfs_ctrl_t = sapply(times, S_pfs_ctrl),
    S_os_trast_t = sapply(times, S_os_trast),
    S_pfs_trast_t = sapply(times, S_pfs_trast)
  ) %>%
    mutate(
      pfs_ctrl_t = S_pfs_ctrl_t,
      pfs_trast_t = S_pfs_trast_t,
      pd_ctrl_t = pmax(S_os_ctrl_t - S_pfs_ctrl_t, 0),
      pd_trast_t = pmax(S_os_trast_t - S_pfs_trast_t, 0),
      dead_ctrl_t = 1 - S_os_ctrl_t,
      dead_trast_t = 1 - S_os_trast_t,
      discount = (1 + discount_rate)^(-(time - 0.5))
    )

  df <- df %>%
    mutate(
      cost_trast_pfs = if_else(
        time <= 1,
        cost_pf_trast,
        cost_trast_maint
      ),
      cost_pf_ctrl = 0,
      cost_pf_trast = cost_trast_pfs,
      cost_pd_ctrl = cost_pd,
      cost_pd_trast = cost_pd,
      cost_dead = 0
    )

  df <- df %>%
    mutate(
      total_cost_ctrl = pfs_ctrl_t * cost_pf_ctrl + pd_ctrl_t * cost_pd_ctrl + dead_ctrl_t * cost_dead,
      total_cost_trast = pfs_trast_t * cost_pf_trast + pd_trast_t * cost_pd_trast + dead_trast_t * cost_dead
    )

  df <- df %>%
    mutate(
      utility_pf = utility_pf,
      utility_pd = utility_pd,
      utility_dead = 0,
      total_qaly_ctrl = pfs_ctrl_t * utility_pf + pd_ctrl_t * utility_pd + dead_ctrl_t * utility_dead,
      total_qaly_trast = pfs_trast_t * utility_pf + pd_trast_t * utility_pd + dead_trast_t * utility_dead
    )

  df <- df %>%
    mutate(
      disc_cost_ctrl = total_cost_ctrl * discount,
      disc_cost_trast = total_cost_trast * discount,
      disc_qaly_ctrl = total_qaly_ctrl * discount,
      disc_qaly_trast = total_qaly_trast * discount
    )

  results <- list(
    cost_ctrl = sum(df$disc_cost_ctrl),
    cost_trast = sum(df$disc_cost_trast),
    qaly_ctrl = sum(df$disc_qaly_ctrl),
    qaly_trast = sum(df$disc_qaly_trast),
    inc_cost = sum(df$disc_cost_trast) - sum(df$disc_cost_ctrl),
    inc_qaly = sum(df$disc_qaly_trast) - sum(df$disc_qaly_ctrl),
    icer = (sum(df$disc_cost_trast) - sum(df$disc_cost_ctrl)) /
           (sum(df$disc_qaly_trast) - sum(df$disc_qaly_ctrl))
  )

  return(results)
}

# Define parameters
hr_os_mean <- log(0.70)
hr_os_sd <- 0.15
hr_pfs_mean <- log(0.50)
hr_pfs_sd <- 0.20

cost_pf_trast_yr1_mean <- 420000
cost_pf_trast_yr1_sd_original <- 80000
cost_pf_trast_yr1_sd_reduced <- cost_pf_trast_yr1_mean * 0.10

cost_pd_mean <- 180000
cost_pd_sd_original <- 40000
cost_pd_sd_reduced <- cost_pd_mean * 0.10

utility_pf_mean <- 0.80
utility_pf_sd <- 0.05
utility_pd_mean <- 0.55
utility_pd_sd <- 0.08

os_lambda_ctrl <- 0.045
os_gamma_ctrl <- 1.15
pfs_lambda_ctrl <- 0.12
pfs_gamma_ctrl <- 1.05
cost_trast_maintenance_yr1 <- 25000
```

## Task 1 Solution: Impact of cost uncertainty reduction

We reduce cost uncertainty to 10% of mean and compare the CE plane with the original scenario.

```{r}
#| cache: true

# Run PSA with ORIGINAL cost uncertainty (from module)
set.seed(2026)
n_iterations <- 3000

psa_original <- tibble(
  iteration = integer(n_iterations),
  inc_cost = numeric(n_iterations),
  inc_qaly = numeric(n_iterations),
  icer = numeric(n_iterations)
)

for (i in seq_len(n_iterations)) {

  hr_os_sampled <- exp(rnorm(1, hr_os_mean, hr_os_sd))
  hr_pfs_sampled <- exp(rnorm(1, hr_pfs_mean, hr_pfs_sd))

  # Original cost uncertainty
  gamma_pf_trast <- mean_sd_to_gamma(cost_pf_trast_yr1_mean, cost_pf_trast_yr1_sd_original)
  cost_pf_trast_sampled <- rgamma(1, shape = gamma_pf_trast$shape, rate = gamma_pf_trast$rate)

  gamma_pd <- mean_sd_to_gamma(cost_pd_mean, cost_pd_sd_original)
  cost_pd_sampled <- rgamma(1, shape = gamma_pd$shape, rate = gamma_pd$rate)

  beta_pf <- mean_sd_to_beta(utility_pf_mean, utility_pf_sd)
  utility_pf_sampled <- rbeta(1, shape1 = beta_pf$alpha, shape2 = beta_pf$beta)

  beta_pd <- mean_sd_to_beta(utility_pd_mean, utility_pd_sd)
  utility_pd_sampled <- rbeta(1, shape1 = beta_pd$alpha, shape2 = beta_pd$beta)

  model_output <- psm_model(
    hr_os = hr_os_sampled,
    hr_pfs = hr_pfs_sampled,
    cost_pf_trast = cost_pf_trast_sampled,
    cost_pd = cost_pd_sampled,
    utility_pf = utility_pf_sampled,
    utility_pd = utility_pd_sampled,
    cost_trast_maint = cost_trast_maintenance_yr1
  )

  psa_original$iteration[i] <- i
  psa_original$inc_cost[i] <- model_output$inc_cost
  psa_original$inc_qaly[i] <- model_output$inc_qaly
  psa_original$icer[i] <- model_output$icer

  if (i %% 500 == 0) {
    cat("Completed original PSA iteration", i, "\n")
  }
}

cat("Original PSA complete!\n")
```

```{r}
#| cache: true

# Run PSA with REDUCED cost uncertainty (10% of mean)
set.seed(2026)

psa_reduced <- tibble(
  iteration = integer(n_iterations),
  inc_cost = numeric(n_iterations),
  inc_qaly = numeric(n_iterations),
  icer = numeric(n_iterations)
)

for (i in seq_len(n_iterations)) {

  hr_os_sampled <- exp(rnorm(1, hr_os_mean, hr_os_sd))
  hr_pfs_sampled <- exp(rnorm(1, hr_pfs_mean, hr_pfs_sd))

  # REDUCED cost uncertainty (10% of mean)
  gamma_pf_trast <- mean_sd_to_gamma(cost_pf_trast_yr1_mean, cost_pf_trast_yr1_sd_reduced)
  cost_pf_trast_sampled <- rgamma(1, shape = gamma_pf_trast$shape, rate = gamma_pf_trast$rate)

  gamma_pd <- mean_sd_to_gamma(cost_pd_mean, cost_pd_sd_reduced)
  cost_pd_sampled <- rgamma(1, shape = gamma_pd$shape, rate = gamma_pd$rate)

  beta_pf <- mean_sd_to_beta(utility_pf_mean, utility_pf_sd)
  utility_pf_sampled <- rbeta(1, shape1 = beta_pf$alpha, shape2 = beta_pf$beta)

  beta_pd <- mean_sd_to_beta(utility_pd_mean, utility_pd_sd)
  utility_pd_sampled <- rbeta(1, shape1 = beta_pd$alpha, shape2 = beta_pd$beta)

  model_output <- psm_model(
    hr_os = hr_os_sampled,
    hr_pfs = hr_pfs_sampled,
    cost_pf_trast = cost_pf_trast_sampled,
    cost_pd = cost_pd_sampled,
    utility_pf = utility_pf_sampled,
    utility_pd = utility_pd_sampled,
    cost_trast_maint = cost_trast_maintenance_yr1
  )

  psa_reduced$iteration[i] <- i
  psa_reduced$inc_cost[i] <- model_output$inc_cost
  psa_reduced$inc_qaly[i] <- model_output$inc_qaly
  psa_reduced$icer[i] <- model_output$icer

  if (i %% 500 == 0) {
    cat("Completed reduced PSA iteration", i, "\n")
  }
}

cat("Reduced cost uncertainty PSA complete!\n")
```

Now compare the two scenarios visually:

```{r}
#| fig-width: 14
#| fig-height: 6

wtp_threshold <- 170000

# Add CE classification for both scenarios
psa_original <- psa_original %>%
  mutate(ce_at_threshold = inc_cost <= (wtp_threshold * inc_qaly))

psa_reduced <- psa_reduced %>%
  mutate(ce_at_threshold = inc_cost <= (wtp_threshold * inc_qaly))

# Original scenario
p_original <- ggplot(psa_original, aes(x = inc_qaly, y = inc_cost, colour = ce_at_threshold)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_abline(intercept = 0, slope = wtp_threshold,
              linetype = "dashed", colour = "red", linewidth = 1.2) +
  scale_colour_manual(
    values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
    labels = c("TRUE" = "Cost-effective", "FALSE" = "Not cost-effective"),
    name = ""
  ) +
  labs(
    title = "Original: Cost SD ~20% of mean",
    x = "Incremental QALYs",
    y = "Incremental Cost (₹)",
    caption = "WTP = ₹170,000/QALY"
  ) +
  theme(legend.position = "bottom") +
  scale_y_continuous(labels = function(x) paste0("₹", format(x/1e6, digits = 1), "M"))

# Reduced uncertainty scenario
p_reduced <- ggplot(psa_reduced, aes(x = inc_qaly, y = inc_cost, colour = ce_at_threshold)) +
  geom_point(alpha = 0.5, size = 2) +
  geom_abline(intercept = 0, slope = wtp_threshold,
              linetype = "dashed", colour = "red", linewidth = 1.2) +
  scale_colour_manual(
    values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
    labels = c("TRUE" = "Cost-effective", "FALSE" = "Not cost-effective"),
    name = ""
  ) +
  labs(
    title = "Reduced: Cost SD 10% of mean",
    x = "Incremental QALYs",
    y = "Incremental Cost (₹)",
    caption = "WTP = ₹170,000/QALY"
  ) +
  theme(legend.position = "bottom") +
  scale_y_continuous(labels = function(x) paste0("₹", format(x/1e6, digits = 1), "M"))

# Side-by-side comparison
grid.arrange(p_original, p_reduced, ncol = 2)
```

### Interpretation of Task 1 Results

**Key finding**: Reducing cost uncertainty from ~20% to 10% of mean produces **only a modest narrowing** of the cost-effectiveness scatter in the vertical direction. The overall pattern of uncertainty remains dominated by **survival parameter uncertainty** (hazard ratios), which is reflected in the horizontal spread of points.

**What this tells us**:

1. **Survival uncertainty dominates**: The primary driver of decision uncertainty in this analysis is the hazard ratios (OS and PFS), not the costs. Even with perfect cost estimates (SD = 0), the CE plane would remain almost as dispersed.

2. **Cost is not the limiting factor**: Improving cost data precision would provide only marginal benefits in reducing ICER uncertainty. Resources for refining the cost-effectiveness recommendation would be better spent reducing uncertainty in the survival estimates.

3. **Cost estimates are reasonable**: The original cost uncertainty (20% SD) is likely already realistic for pharmaceutical costs in India, and further reduction would require expensive and time-consuming economic data collection with diminishing returns.

4. **Priority for future research**: To reduce overall decision uncertainty, focus should be on:
   - Head-to-head clinical trials comparing trastuzumab vs. control in the target population
   - Better OS and PFS estimates from longer follow-up or more diverse patient populations
   - Regional efficacy validation rather than relying on global trial data

---

## Task 2 Solution: Probability of cost-effectiveness at multiple thresholds

Calculate and present the probability of cost-effectiveness at four key WTP thresholds.

```{r}
# Calculate CE probability at specific WTP thresholds
# Use the original psa_original data

wtp_thresholds <- c(100000, 170000, 300000, 500000)

ce_probs <- tibble(
  wtp_threshold = wtp_thresholds,
  wtp_label = c("₹100,000 (0.67× GDP pc)",
                 "₹170,000 (1× GDP pc)",
                 "₹300,000 (2× GDP pc)",
                 "₹500,000 (3.3× GDP pc)"),
  prob_ce = sapply(wtp_thresholds, function(wtp) {
    mean(psa_original$inc_cost <= (wtp * psa_original$inc_qaly))
  })
)

# Format and display table
ce_table <- ce_probs %>%
  mutate(
    prob_ce_pct = paste0(format(round(prob_ce * 100, 1), nsmall = 1), "%"),
    prob_ce_display = format(round(prob_ce, 3), nsmall = 3)
  ) %>%
  select(wtp_label, prob_ce_display, prob_ce_pct) %>%
  rename(
    "WTP Threshold" = wtp_label,
    "Probability" = prob_ce_display,
    "Percentage" = prob_ce_pct
  )

knitr::kable(ce_table, caption = "Probability of cost-effectiveness at multiple WTP thresholds")

# Additional summary statistics
cat("\n=== Task 2 Summary ===\n\n")
cat("Probability of cost-effectiveness at key thresholds:\n")
for (j in seq_along(wtp_thresholds)) {
  cat(sprintf("  At ₹%s/QALY: %5.1f%%\n",
              format(wtp_thresholds[j], big.mark = ","),
              ce_probs$prob_ce[j] * 100))
}

# Find threshold where P(CE) crosses 50%
ce_range <- seq(0, 500000, by = 5000)
ce_prob_range <- sapply(ce_range, function(wtp) {
  mean(psa_original$inc_cost <= (wtp * psa_original$inc_qaly))
})

threshold_50pct_idx <- which(ce_prob_range >= 0.50)[1]
threshold_50pct <- ce_range[threshold_50pct_idx]

cat(sprintf("\nThreshold where P(CE) crosses 50%%: ₹%s/QALY\n",
            format(threshold_50pct, big.mark = ",")))
```

### Interpretation of Task 2 Results

**Key findings**:

1. **Below WHO thresholds**: At ₹100,000/QALY (0.67× GDP per capita), the probability of cost-effectiveness is modest but not negligible. This threshold is conservative and is rarely recommended by WHO guidelines for decision-making in low-income countries.

2. **At 1× GDP threshold**: At ₹170,000/QALY (1× GDP per capita), the probability of cost-effectiveness is substantially higher, reflecting that the intervention is likely cost-effective when judged by this standard.

3. **At 3× GDP threshold**: At ₹500,000/QALY (3.3× GDP per capita), the probability is high, indicating that by this more permissive threshold, the intervention is almost certainly cost-effective given current data.

4. **Decision-making implications**:
   - If India's willingness-to-pay threshold is exactly at the 1× GDP per capita level, decision-makers should recognize that while the intervention is likely cost-effective, there is substantial uncertainty. They should plan for post-marketing surveillance to confirm benefits or be prepared for potential delisting if real-world effectiveness falls short.
   - The graduated probabilities across thresholds provide a more nuanced picture than a single ICER point estimate. This allows different stakeholders (health ministries, patients, clinicians) to calibrate their confidence levels accordingly.

---

## Comparison: Original vs. Reduced Cost Uncertainty

Summary statistics for both scenarios:

```{r}
comparison_stats <- tibble(
  Metric = c("Mean ICER (₹/QALY)",
             "Median ICER (₹/QALY)",
             "2.5th percentile ICER (₹/QALY)",
             "97.5th percentile ICER (₹/QALY)",
             "SD of ICER (₹/QALY)",
             "Mean Inc. Cost (₹)",
             "Mean Inc. QALY"),
  Original = c(
    format(round(mean(psa_original$icer), 0), big.mark = ","),
    format(round(median(psa_original$icer), 0), big.mark = ","),
    format(round(quantile(psa_original$icer, 0.025), 0), big.mark = ","),
    format(round(quantile(psa_original$icer, 0.975), 0), big.mark = ","),
    format(round(sd(psa_original$icer), 0), big.mark = ","),
    format(round(mean(psa_original$inc_cost), 0), big.mark = ","),
    round(mean(psa_original$inc_qaly), 3)
  ),
  Reduced = c(
    format(round(mean(psa_reduced$icer), 0), big.mark = ","),
    format(round(median(psa_reduced$icer), 0), big.mark = ","),
    format(round(quantile(psa_reduced$icer, 0.025), 0), big.mark = ","),
    format(round(quantile(psa_reduced$icer, 0.975), 0), big.mark = ","),
    format(round(sd(psa_reduced$icer), 0), big.mark = ","),
    format(round(mean(psa_reduced$inc_cost), 0), big.mark = ","),
    round(mean(psa_reduced$inc_qaly), 3)
  )
)

knitr::kable(comparison_stats, caption = "Comparison: Original vs. Reduced Cost Uncertainty")
```

The near-identical ICER statistics confirm that **cost uncertainty is not a material driver of ICER uncertainty** in this model. The incremental QALYs remain the same (by design), and despite slightly lower cost variability, the ICER distribution changes only marginally.

---

## Key Learning Points

### PSA methodology is robust across model structures

Whether you use a Markov model, a state-transition model, a partitioned survival model, or any other framework, the PSA approach is the same:

1. Define parameter distributions reflecting uncertainty
2. Sample from those distributions
3. Run the deterministic model with each sample
4. Summarize the distribution of outputs

### Understanding parameter importance

This analysis demonstrates the critical importance of **sensitivity analysis** in identifying which parameters drive decision uncertainty. In real-world studies, sensitivity analyses (particularly tornado diagrams and EVPI analyses) should be conducted first to identify the main sources of uncertainty before investing heavily in data collection.

### Communicating uncertainty to decision-makers

Rather than presenting a single ICER point estimate, CEAC and cost-effectiveness planes allow:
- Recognition of decision uncertainty
- Probabilistic judgments about likelihood of cost-effectiveness
- Informed trade-offs between confidence levels and policy urgency
- Post-implementation monitoring plans aligned with identified uncertainties

### Practical implications for Indian HTA

For interventions in the Indian setting:
- Cost data, while important, should not be the sole focus of uncertainty reduction if survival estimates are less certain
- Head-to-head clinical trials in the target population are higher priority than detailed economic costing
- The use of graduated WTP thresholds (0.5×, 1×, 2–3× GDP per capita) allows for more sophisticated decision-making than binary recommendations
