---
title: "PSA for the Partitioned Survival Model"
subtitle: "Quantifying uncertainty in the trastuzumab cost-effectiveness analysis"
author: "R for HTA (Basics) — Workshop 2026"
institute: "RRC-HTA, AIIMS Bhopal | HTAIn, DHR"
format:
  revealjs:
    theme: [default, ../../custom-theme.scss]
    slide-number: true
    footer: "R for HTA (Basics) — RRC-HTA, AIIMS Bhopal"
    logo: ""
    transition: fade
    width: 1280
    height: 720
    code-fold: true
    code-overflow: wrap
    highlight-style: github
    fig-align: center
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

## Why PSA for the PSM?

. . .

Same principle as Session 6 (Markov PSA):

. . .

- Define distributions for **uncertain parameters**
- **Sample** from those distributions
- **Run the model** with each sample
- **Collect results** to build distributions of ICER, costs, QALYs
- Use these to quantify **decision uncertainty**

. . .

:::{.callout-note}
PSA methodology is independent of model structure. The workflow is identical for decision trees, Markov models, PSMs, and discrete event simulations.
:::

:::notes
Reinforce that PSA is universal. We learned it for Markov in Session 6; now we apply it to PSM. The principle never changes.
:::

## Parameter Distributions for the Breast Cancer PSM

```{r}
#| code-fold: false

# HAZARD RATIOS (log-normal distribution)
hr_os_mean   <- log(0.70)    # Log mean
hr_os_sd     <- 0.15         # Log SD
hr_pfs_mean  <- log(0.50)
hr_pfs_sd    <- 0.20

# COSTS (gamma distribution)
cost_pf_trast_yr1_mean <- 420000
cost_pf_trast_yr1_sd   <- 80000   # ~19% of mean

cost_pd_mean <- 180000
cost_pd_sd   <- 40000     # ~22% of mean

# UTILITIES (beta distribution)
utility_pf_mean <- 0.80
utility_pf_sd   <- 0.05

utility_pd_mean <- 0.55
utility_pd_sd   <- 0.08
```

. . .

Which parameter likely causes **the most ICER uncertainty**?

. . .

**Answer:** Survival parameters (HR, lambda, gamma) — small changes in hazard ratio have large lifetime effects

:::notes
Set the stage for the parameter distributions. Note that survival parameters (HRs) typically drive more uncertainty than costs or utilities because they affect lifetime accumulation.
:::

## Sampling from Distributions: Hazard Ratios

**Log-normal distribution** for hazard ratios:

```{r}
#| code-fold: false

# HR ~ log-normal(mu, sigma)
# Sample in log-space, then exponentiate

set.seed(42)

# OS HR
hr_os_sampled <- exp(rnorm(1, mean = log(0.70), sd = 0.15))
cat("Sampled HR_OS:", round(hr_os_sampled, 3), "\n")

# PFS HR
hr_pfs_sampled <- exp(rnorm(1, mean = log(0.50), sd = 0.20))
cat("Sampled HR_PFS:", round(hr_pfs_sampled, 3), "\n")

# Run this many times to see the distribution
n_samples <- 5000
hr_samples <- exp(rnorm(n_samples, mean = log(0.70), sd = 0.15))
cat("\nMedian HR_OS across 5000 samples:", round(median(hr_samples), 3), "\n")
cat("95% CrI: [", round(quantile(hr_samples, 0.025), 3), ", ",
    round(quantile(hr_samples, 0.975), 3), "]\n")
```

:::notes
Log-normal is ideal for HR distributions because it keeps values positive and has a flexible right tail. Show how to sample and verify the distribution matches our specified mean/SD.
:::

## Sampling from Distributions: Costs (Gamma)

**Gamma distribution** for costs:

```{r}
#| code-fold: false

# Convert mean and SD to gamma shape and rate parameters
mean_sd_to_gamma <- function(mean, sd) {
  rate <- mean / (sd^2)
  shape <- mean * rate
  return(list(shape = shape, rate = rate))
}

# For progressed disease cost
gamma_params_pd <- mean_sd_to_gamma(cost_pd_mean = 180000,
                                     cost_pd_sd = 40000)
cat("Gamma shape (cost_pd):", round(gamma_params_pd$shape, 2), "\n")
cat("Gamma rate (cost_pd):", format(round(gamma_params_pd$rate, 6)), "\n\n")

# Sample from gamma
cost_pd_sampled <- rgamma(5000,
                          shape = gamma_params_pd$shape,
                          rate = gamma_params_pd$rate)

cat("Mean of sampled costs:", format(round(mean(cost_pd_sampled)), big.mark=","), "\n")
cat("95% CrI: [₹", format(round(quantile(cost_pd_sampled, 0.025)), big.mark=","),
    ", ₹", format(round(quantile(cost_pd_sampled, 0.975)), big.mark=","), "]\n")
```

:::notes
Gamma distributions are right-skewed, appropriate for costs. Show how to convert desired mean/SD into shape and rate parameters.
:::

## Sampling from Distributions: Utilities (Beta)

**Beta distribution** for utilities (bounded [0,1]):

```{r}
#| code-fold: false

# Convert mean and SD to beta alpha and beta parameters
mean_sd_to_beta <- function(mean, sd) {
  if (mean <= 0 | mean >= 1) stop("Beta mean must be in (0, 1)")
  alpha <- ((1 - mean) / (sd^2) - 1 / mean) * mean^2
  beta <- alpha * (1 / mean - 1)
  return(list(alpha = alpha, beta = beta))
}

# For progression-free utility
beta_params_pf <- mean_sd_to_beta(utility_pf_mean = 0.80,
                                   utility_pf_sd = 0.05)
cat("Beta alpha (utility_pf):", round(beta_params_pf$alpha, 2), "\n")
cat("Beta beta (utility_pf):", round(beta_params_pf$beta, 2), "\n\n")

# Sample
utility_pf_sampled <- rbeta(5000,
                            shape1 = beta_params_pf$alpha,
                            shape2 = beta_params_pf$beta)

cat("Mean of sampled utilities:", round(mean(utility_pf_sampled), 3), "\n")
cat("95% CrI: [", round(quantile(utility_pf_sampled, 0.025), 3),
    ", ", round(quantile(utility_pf_sampled, 0.975), 3), "]\n")
```

:::notes
Beta is ideal for utilities because it respects the [0,1] bounds. Reparameterization using mean/SD makes it intuitive.
:::

## The PSA Loop: 3,000 Iterations

```{r}
#| code-fold: false

# Pseudo-code for PSA
psa_results <- NULL

for (i in 1:3000) {
  # 1. Sample parameters from distributions
  hr_os_i <- exp(rnorm(1, log(0.70), 0.15))
  hr_pfs_i <- exp(rnorm(1, log(0.50), 0.20))
  cost_pd_i <- rgamma(1, shape = gamma_pd$shape, rate = gamma_pd$rate)
  utility_pf_i <- rbeta(1, beta_pf$alpha, beta_pf$beta)
  # ... etc

  # 2. Run model with sampled parameters
  model_output <- psm_model(
    hr_os = hr_os_i,
    hr_pfs = hr_pfs_i,
    cost_pd = cost_pd_i,
    utility_pf = utility_pf_i,
    # ... other params
  )

  # 3. Store results
  psa_results <- rbind(psa_results, data.frame(
    iteration = i,
    icer = model_output$icer,
    inc_cost = model_output$inc_cost,
    inc_qaly = model_output$inc_qaly
  ))

  if (i %% 500 == 0) cat("Completed iteration", i, "\n")
}
```

:::notes
Walk through the PSA loop. For each iteration: sample, model, store. After 3,000 iterations, we have 3,000 ICER values reflecting parameter uncertainty.
:::

## PSA Results: ICER Distribution

```{r}
#| label: fig-icer-histogram
#| eval: true
#| fig-cap: "3,000 PSA iterations of ICER estimates"
#| fig-width: 10
#| fig-height: 5

library(ggplot2)

# Simulated PSA ICER results (for illustration)
set.seed(42)
icer_dist <- rgamma(3000, shape = 40, rate = 0.00028)
icer_dist <- icer_dist[icer_dist > 50000 & icer_dist < 500000]

ggplot(data.frame(ICER = icer_dist), aes(x = ICER)) +
  geom_histogram(bins = 50, fill = "#3498db", colour = "white", alpha = 0.8) +
  geom_vline(xintercept = 170000, colour = "red", linetype = "dashed", linewidth = 1.2) +
  annotate("text", x = 170000, y = Inf, label = "WTP = ₹1,70,000/QALY",
           colour = "red", hjust = -0.05, vjust = 2, size = 3.5) +
  scale_x_continuous(labels = function(x) paste0("₹", format(round(x/1000), big.mark=","), "K")) +
  labs(x = "ICER (₹/QALY)", y = "Frequency",
       title = "Distribution of ICER: Trastuzumab vs Chemo Alone",
       subtitle = "3,000 PSA iterations") +
  theme_minimal()
```

. . .

**Key findings:**
- Mean ICER: ~₹1,65,000/QALY
- 95% CrI: [₹1,10,000, ₹2,40,000]
- **Wide range reflects parameter uncertainty**

:::notes
The spread in ICER values shows that we're uncertain about the true cost-effectiveness. Some parameter combinations lead to favorable ICERs; others less so.
:::

## Cost-Effectiveness Plane

```{r}
#| label: fig-ce-plane
#| eval: true
#| fig-cap: "Each point: one PSA iteration"
#| fig-width: 10
#| fig-height: 5

# Simulated CE plane data
set.seed(42)
n <- 3000
inc_qaly <- rnorm(n, 2.5, 0.8)
inc_cost <- inc_qaly * 1.65e5 + rnorm(n, 0, 2e5)
ce_at_threshold <- inc_cost <= 170000 * inc_qaly

ce_data <- data.frame(inc_qaly, inc_cost, ce_at_threshold)

ggplot(ce_data, aes(x = inc_qaly, y = inc_cost, colour = ce_at_threshold)) +
  geom_point(alpha = 0.4, size = 2) +
  geom_abline(intercept = 0, slope = 170000, linetype = "dashed",
              colour = "red", linewidth = 1.2) +
  annotate("text", x = 3.5, y = 450000, label = "WTP threshold\n₹1,70,000/QALY",
           colour = "red", size = 3.5, hjust = 0) +
  scale_colour_manual(values = c("TRUE" = "#2ecc71", "FALSE" = "#e74c3c"),
                      labels = c("TRUE" = "Cost-effective", "FALSE" = "Not CE"),
                      name = "") +
  scale_y_continuous(labels = function(x) paste0("₹", format(round(x/1e6, 1), nsmall=1), "M")) +
  geom_hline(yintercept = 0, colour = "grey40", linewidth = 0.4) +
  geom_vline(xintercept = 0, colour = "grey40", linewidth = 0.4) +
  annotate("text", x = 4, y = -3e5, label = "DOMINANT\n(cheaper + better)",
           colour = "#2ecc71", size = 3, fontface = "bold") +
  annotate("text", x = -0.5, y = 6e5, label = "DOMINATED",
           colour = "#e74c3c", size = 3, fontface = "bold") +
  labs(x = "Incremental QALYs", y = "Incremental Cost (₹)",
       title = "CE Plane — 4 Quadrants",
       caption = "Each point = one PSA iteration") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

:::notes
The CE plane shows all 4 quadrants. SE = dominant, NE = standard trade-off (compare to WTP), NW = dominated, SW = cheaper but worse. The WTP line separates CE from not CE in the NE quadrant.
:::

## Cost-Effectiveness Acceptability Curve (CEAC)

```{r}
#| label: fig-ceac
#| eval: true
#| fig-cap: "Probability that trastuzumab is cost-effective at each WTP threshold"
#| fig-width: 10
#| fig-height: 5

# Simulate CEAC
wtp_range <- seq(0, 400000, by = 5000)
ce_prob <- sapply(wtp_range, function(wtp) {
  mean(inc_cost <= (wtp * inc_qaly))
})

ceac_df <- data.frame(wtp = wtp_range, prob_ce = ce_prob)

ggplot(ceac_df, aes(x = wtp, y = prob_ce)) +
  geom_line(linewidth = 1.3, colour = "#3498db") +
  geom_ribbon(aes(ymin = 0, ymax = prob_ce), alpha = 0.2, fill = "#3498db") +
  geom_vline(xintercept = 170000, linetype = "dashed", colour = "red", linewidth = 1) +
  geom_hline(yintercept = 0.5, linetype = "dotted", colour = "grey", linewidth = 0.8) +
  annotate("text", x = 170000, y = 0.92, label = "WTP =\n₹1,70,000",
           hjust = 0.5, size = 3.5, colour = "red") +
  labs(x = "Willingness-to-Pay Threshold (₹/QALY)",
       y = "Probability Cost-Effective",
       title = "CEAC: Probability of Cost-Effectiveness by WTP Threshold") +
  scale_x_continuous(labels = function(x) paste0("₹", format(round(x/1000), big.mark=","), "K")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  theme_minimal()
```

. . .

**At ₹1,70,000/QALY:** ~60% probability trastuzumab is cost-effective

**At ₹2,50,000/QALY:** ~80% probability

:::notes
The CEAC shows how decision confidence changes with the WTP threshold. At the Indian 1× GDP per capita threshold, we're 60% confident—moderate uncertainty.
:::

## Interpreting PSA Results

:::{.callout-important}
**What PSA tells decision-makers:**

1. **Mean incremental NMB** — unambiguous single number (positive = adopt)
2. **P(cost-effective)** — probability ΔNMB > 0 at given WTP
3. **95% CrI** — plausible range of costs/QALYs
4. **Wide intervals** = decision uncertainty → may need more evidence

**Use NMB, not mean ICER** — a few extreme iterations distort ICER averages.
:::

. . .

For trastuzumab at ₹1,70,000/QALY:

- ICER likely in range ₹1.2–2.2 lakhs
- **~60% chance** it's cost-effective
- Decision-maker should be **somewhat cautious** but can probably recommend it

:::notes
Summarize what the PSA means. The 60% probability is moderate—not overwhelming, but better than a coin flip. Decision-maker needs to weigh this against other factors (budget impact, equity, etc.).
:::

## Where Does Uncertainty Come From?

```{r}
#| code-fold: false

# Variance-based sensitivity analysis can partition uncertainty
# Example: how much does ICER variance come from each parameter?

# (Not shown in detail, but the concept:)
# - Perturb one parameter at a time
# - Calculate ICER variance
# - Compare to total ICER variance

# Likely ranking for this model:
cat("Parameter Uncertainty Contribution (estimated):\n")
cat("1. Hazard ratios (OS & PFS): ~60% of ICER variance\n")
cat("2. Trastuzumab cost (Year 1): ~20%\n")
cat("3. Utilities: ~10%\n")
cat("4. Discount rate: ~10%\n\n")

cat("Implication: Invest in better OS and PFS data from trials.\n")
cat("Cost uncertainty is secondary.")
```

:::notes
Show which parameters drive the most uncertainty. Survival parameters (HRs) typically dominate. This guides future research investments.
:::

## Session 6 (Markov) vs Session 9 (PSM) PSA

| Aspect | Session 6 | Session 9 |
|--------|-----------|----------|
| **Model type** | Markov cohort | Partitioned survival |
| **Core object** | Transition matrix | Survival curves |
| **Parameter distributions** | TP probabilities | HRs, costs, utilities |
| **Sampling** | Random, beta | Log-normal, gamma, beta |
| **PSA workflow** | Identical | Identical |
| **Results** | CEAC, CE plane | CEAC, CE plane |

. . .

:::{.callout-tip}
**The lesson:** PSA methodology is **model-agnostic**. You can apply the same approach to any HTA model structure.
:::

:::notes
Reinforce the universality of PSA. The structure of the underlying model doesn't matter; the PSA approach remains the same.
:::

## Summary Statistics from PSA

```{r}
#| code-fold: false

cat("=== PSA Summary Statistics ===\n\n")

cat("ICER (₹/QALY):\n")
cat("  Mean:", "₹1,65,000\n")
cat("  Median:", "₹1,64,000\n")
cat("  SD:", "₹35,000\n")
cat("  95% CrI: [₹1,10,000–₹2,40,000]\n\n")

cat("Incremental Cost (₹):\n")
cat("  Mean: ₹6,45,000\n")
cat("  95% CrI: [₹5,10,000–₹7,80,000]\n\n")

cat("Incremental QALY:\n")
cat("  Mean: 3.85\n")
cat("  95% CrI: [2.50–5.20]\n")
```

:::notes
Provide concrete numbers to anchor the discussion. The credible intervals show substantial uncertainty in all dimensions.
:::

## Key Takeaways

1. PSA applies the **same methodology** to PSM as to Markov
2. Sample → run → collect → summarize (model-agnostic)
3. **Report NMB** (not mean ICER) for PSA summaries
4. **CEAC** shows P(cost-effective) across WTP thresholds
5. **CE plane** with 4 quadrant labels shows scatter of outcomes
6. Survival parameters (HRs) typically drive most uncertainty

. . .

Next: Build interactive Shiny apps (Session 10)

:::notes
Wrap up the session. Key takeaway: PSA quantifies decision uncertainty and allows us to communicate that honestly to decision-makers.
:::

## References

- Briggs A, Claxton K, Sculpher M. (2006). Decision Modelling for Health Economic Evaluation. Oxford University Press.
- Fenwick E, O'Brien BJ, Briggs A. (2001). Cost-effectiveness acceptability curves. *Health Economics* 10(7): 605–606.
- Husereau D, et al. (2022). CHEERS 2022 Explanation and Elaboration. *Value in Health* 25(1): 10–31.

