---
title: "The Case for R in HTA"
subtitle: "Why transition from Excel and TreeAge?"
author: "R for HTA (Basics) — Workshop 2026"
institute: "RRC-HTA, AIIMS Bhopal | HTAIn, DHR"
format:
  revealjs:
    theme: [default, ../../custom-theme.scss]
    slide-number: true
    footer: "R for HTA (Basics) — RRC-HTA, AIIMS Bhopal"
    logo: ""
    transition: fade
    width: 1280
    height: 720
    code-fold: true
    code-overflow: wrap
    highlight-style: github
    fig-align: center
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

## Why Are We Here?

You have been doing HTA work — and doing it well — using Excel and TreeAge.

. . .

This session does not ask you to abandon what works.

. . .

It asks you to consider **what becomes possible** when you add R to your toolkit.

## The Limitations You Already Know

PSA is painful
- Complex macros or VBA required
- Spreadsheet becomes slow and fragile

. . .

Reproducibility is difficult
- Tracing nested cell references tedious

. . .

Version control is manual
- Model_v3_final_FINAL_revised.xlsx?

. . .

Transparency for publication
- Journals and agencies require open-source, auditable models

## What R Makes Possible

**Reproducible**
: Your entire analysis is a script — anyone can re-run it and get the same result

. . .

**Scalable**
: 10,000 PSA iterations in seconds, not hours

. . .

**Transparent**
: A reviewer can read your code and see exactly what you did

. . .

**Version-controlled**
: Every change tracked, documented, reversible

. . .

**Free and extensible**
: No license fees — thousands of packages for health economics

## The HTA Workflow in R

```{r}
#| label: fig-hta-workflow
#| eval: true
#| fig-cap: "The complete modelling workflow"

library(DiagrammeR)

grViz("
digraph hta_workflow {
  graph [rankdir=LR, bgcolor='transparent', fontname='Helvetica']
  node [shape=box, style='filled,rounded', fontname='Helvetica', fontsize=11]

  A [label='Define\nParameters', fillcolor='#4e79a7', fontcolor='white']
  B [label='Build\nModel', fillcolor='#59a14f', fontcolor='white']
  C [label='Run\nAnalysis', fillcolor='#f28e2b', fontcolor='white']
  D [label='Sensitivity\nAnalysis', fillcolor='#e15759', fontcolor='white']
  E [label='Visualise\nResults', fillcolor='#76b7b2', fontcolor='white']
  F [label='Report &\nPublish', fillcolor='#edc948', fontcolor='black']

  A -> B -> C -> D -> E -> F
  D -> B [style=dashed, label='Iterate', fontsize=9]
}
")
```

## Simple ICER: Excel Logic vs R

**The Problem**

New screening: ₹2,500 cost, 0.85 QALYs
Standard care: ₹1,200 cost, 0.78 QALYs
What is the ICER?

. . .

**In Excel** ✓ Works for one calculation
```
| Cell | Formula |
|------|---------|
| B6   | =B2-B3  | (Incremental cost)
| B7   | =B4-B5  | (Incremental QALY)
| B8   | =B6/B7  | (ICER)
```

. . .

**Now what if you want 10,000 simulations?** ✗ Painful

## The R Code for ICER

```{r}
#| label: simple-icer
#| echo: true
#| eval: false

# Define inputs
cost_intervention <- 2500
cost_standard     <- 1200
qaly_intervention <- 0.85
qaly_standard     <- 0.78

# Calculate ICER
incremental_cost  <- cost_intervention - cost_standard
incremental_qaly  <- qaly_intervention - qaly_standard
icer <- incremental_cost / incremental_qaly

cat("ICER: ₹", round(icer, 0), "per QALY\n")
```

So far, not obviously better than Excel...

## The Power: PSA in Three Lines

```{r}
#| label: simple-psa
#| echo: true
#| eval: false

# Sample from distributions and calculate 10,000 ICERs
icer_psa <- (rnorm(10000, 2500, 300) - rnorm(10000, 1200, 150)) /
            (rbeta(10000, 85, 15) - rbeta(10000, 78, 22))

cat("Mean ICER: ₹", round(mean(icer_psa), 0), "per QALY\n")
cat("95% CI: ₹", round(quantile(icer_psa, 0.025), 0),
    "to ₹", round(quantile(icer_psa, 0.975), 0), "\n")
```

**10,000 simulations in a fraction of a second.**

Try doing that in Excel.

## The Cost-Effectiveness Plane

```{r}
#| label: ce-plane-simple
#| echo: true
#| eval: false

library(ggplot2)

ce_data <- data.frame(
  inc_qaly = rbeta(10000, 85, 15) - rbeta(10000, 78, 22),
  inc_cost = rnorm(10000, 1300, 350)
)

ggplot(ce_data, aes(x = inc_qaly, y = inc_cost)) +
  geom_point(alpha = 0.1, colour = "steelblue") +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Incremental QALYs", y = "Incremental Cost (₹)",
       title = "Cost-Effectiveness Plane: 10,000 PSA iterations") +
  theme_minimal()
```

## The Cost-Effectiveness Acceptability Curve

```{r}
#| label: simple-ceac
#| echo: true
#| eval: false

wtp_range <- seq(0, 300000, by = 5000)
prob_ce <- sapply(wtp_range, function(wtp) {
  mean(inc_cost <= wtp * inc_qaly)
})

ggplot(data.frame(WTP = wtp_range, Prob = prob_ce),
       aes(x = WTP, y = Prob)) +
  geom_line(colour = "steelblue", linewidth = 1.2) +
  geom_vline(xintercept = 100000, linetype = "dashed", colour = "red") +
  annotate("text", x = 110000, y = 0.3, label = "WTP = ₹1,00,000") +
  labs(x = "Willingness-to-Pay (₹/QALY)", y = "Probability Cost-Effective",
       title = "Cost-Effectiveness Acceptability Curve") +
  theme_minimal()
```

## Beyond This Simple Example

In this workshop, you will see R applied to **real HTA problems**:

. . .

- **Decision trees** with instant re-runs at different parameters
- **Markov models** with probabilistic transition probabilities
- **Partitioned survival models** with multiple extrapolation methods
- **Shiny apps** for interactive policymaker tools

## Four Model Types We'll Cover

```{r}
#| label: fig-four-models
#| eval: true
#| fig-cap: "HTA model types in this workshop"

grViz("
digraph four_models {
  graph [rankdir=TB, bgcolor='transparent', fontname='Helvetica']
  node [shape=box, style='filled,rounded', fontname='Helvetica', fontsize=11, width=2.5]

  center [label='HTA Models in R', shape=ellipse, fillcolor='#4e79a7', fontcolor='white', fontsize=13]

  dt_diag [label='Diagnostic\nDecision Tree\n(GDM)', fillcolor='#59a14f', fontcolor='white']
  dt_ther [label='Therapeutic\nDecision Tree\n(DES vs BMS)', fillcolor='#f28e2b', fontcolor='white']
  markov [label='Markov Cohort\nModel\n(CKD)', fillcolor='#e15759', fontcolor='white']
  psm [label='Partitioned Survival\nModel\n(Breast Cancer)', fillcolor='#76b7b2', fontcolor='white']

  center -> dt_diag
  center -> dt_ther
  center -> markov
  center -> psm
}
")
```

## What This Workshop Will NOT Do

- Turn you into R programmers in three days
- Ask you to memorise syntax
- Expect you to write complex code from scratch

## What This Workshop WILL Do

- Give you working templates you can adapt
- Build confidence in reading and modifying R code
- Show how to use AI tools to support your workflow
- Help you understand when R is the right tool

## Key Takeaway

R is not replacing your HTA expertise.

. . .

It is giving that expertise **a more powerful medium**.

. . .

The clinical judgement, model structure, parameter selection — that remains yours.

. . .

R simply makes the execution **more reproducible, transparent, and scalable**.

. . .

Let us get started.
