---
title: "Beyond Spreadsheets — Why R for HTA"
subtitle: "Session 1 | R for HTA (Basics) Workshop 2026"
author: "RRC-HTA, AIIMS Bhopal | HTAIn, DHR"
format:
  revealjs:
    theme: [default, ../../custom-theme.scss]
    slide-number: true
    footer: "R for HTA (Basics) — RRC-HTA, AIIMS Bhopal"
    logo: ""
    transition: fade
    width: 1280
    height: 720
    code-fold: true
    code-overflow: wrap
    highlight-style: github
    fig-align: center
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

## The Spreadsheet That Shook Nations {.section-slide}

::: {.notes}
This section sets the stage with a real-world cautionary tale. The Reinhart-Rogoff incident is the most powerful argument for why spreadsheets can be dangerous when the stakes are high.
:::

## 2008: The World is on Fire

The global financial crisis had just hit. Governments were debating the biggest economic question of the decade:

. . .

::: {style="font-size: 1.1em; color: #028090; font-weight: bold; margin: 16px 0;"}
Should we spend our way out of the crisis, or cut spending to reduce debt?
:::

. . .

Trillions of dollars — and the livelihoods of millions — hung on the answer.

## 2010: Harvard Economists Provide the Answer

Carmen Reinhart and Kenneth Rogoff published **"Growth in a Time of Debt"**.

. . .

Their finding: countries with public debt above **90% of GDP** experience **negative economic growth** (−0.1% average).

. . .

::: {style="font-size: 1.2em; font-weight: bold; color: #e74c3c; margin-top: 10px;"}
Cited 3,000+ times. A paper that shaped global policy.
:::

. . .

**Paul Ryan** cited it in the US Budget Committee. The **European Commission** and **UK Treasury** used it to justify austerity. The **IMF** referenced it in country recommendations.

## What Happened Next

Governments across Europe and the US cut public spending — affecting:

. . .

::: {.comparison-grid}
::: {style="background: #FEF2F2; padding: 16px; border-radius: 8px; border-left: 4px solid #e15759;"}
**Healthcare budgets** — hospital funding reduced, waiting lists grew
:::

::: {style="background: #FEF2F2; padding: 16px; border-radius: 8px; border-left: 4px solid #e15759;"}
**Social services** — disability support, unemployment benefits cut
:::

::: {style="background: #FEF2F2; padding: 16px; border-radius: 8px; border-left: 4px solid #e15759;"}
**Education funding** — university fees tripled in England
:::

::: {style="background: #FEF2F2; padding: 16px; border-radius: 8px; border-left: 4px solid #e15759;"}
**Research grants** — scientific funding stalled across Europe
:::
:::

## 2013: A Graduate Student Checks the Numbers

Thomas Herndon, a **28-year-old PhD student** at UMass Amherst, was given a class assignment:

. . .

*Pick a famous economics paper and try to replicate its results.*

. . .

He chose Reinhart-Rogoff. He couldn't replicate the results.

. . .

He emailed the authors. **They sent him the spreadsheet.**

## The Three Errors {.smaller}

::: {style="font-size: 0.95em;"}

**Error 1: Excel Row Exclusion**

The `AVERAGE` formula covered rows 30–44 instead of 30–49. Five countries (Australia, Austria, Belgium, Canada, Denmark) were simply left out. A drag-select error — the kind Excel makes invisible.

. . .

**Error 2: Unconventional Weighting**

New Zealand's single year of high-debt data (−7.6% growth) was given the same weight as the UK's 19 years. This one outlier dragged the entire average into negative territory.

. . .

**Error 3: Selective Data Exclusion**

Post-war data from three countries was excluded without clear justification.

:::

## The Corrected Result

::: {.comparison-grid}

::: {style="text-align: center; padding: 28px; background: #FEF2F2; border-radius: 12px; border: 3px solid #e15759;"}
::: {.stat-callout}
::: {.stat-number}
−0.1%
:::
::: {.stat-label}
ORIGINAL (with errors)
:::
:::
:::

::: {style="text-align: center; padding: 28px; background: #F0FDF4; border-radius: 12px; border: 3px solid #028090;"}
::: {.stat-callout}
::: {.stat-number}
+2.2%
:::
::: {.stat-label}
CORRECTED (Herndon et al.)
:::
:::
:::
:::

. . .

::: {style="text-align: center; font-size: 1.2em; font-weight: bold; color: #e74c3c; margin-top: 16px;"}
Growth was POSITIVE, not negative. One spreadsheet error. Billions in policy impact.
:::

. . .

Nobody could catch the error earlier — because **nobody could audit the spreadsheet**.

## Five Lessons from Spreadsheet-Gate

1. **Errors hide in plain sight** — wrong cell ranges are invisible in Excel
2. **Peer review failed** — reviewers couldn't access or audit the calculations
3. **Replication was impossible** — until the original spreadsheet was shared
4. **Policy moved faster than science** — the paper was cited before it was checked
5. **Transparency is not optional** — it is the only protection

## Could This Happen to YOUR HTA Model?

Your cost-effectiveness analysis informs drug pricing, reimbursement, and treatment guidelines.

. . .

Can a reviewer **trace every calculation** in your Excel model?

. . .

Can **you** be certain there is no hidden error in a nested cell reference?

. . .

::: {style="font-size: 1.3em; color: #028090; font-weight: bold; margin-top: 16px;"}
What if there was a better way?
:::

## Live Demo: Beyond Spreadsheets {.section-slide}

::: {style="font-size: 0.9em; color: rgba(255,255,255,0.8);"}
Let me show you what R makes possible — interactively.
:::

## Open the Shiny App

::: {style="text-align: center; padding: 20px;"}

::: {style="font-size: 1.3em; font-weight: bold; color: #028090; margin-bottom: 20px;"}
🔗 [PLACEHOLDER_SHINY_APP_URL](PLACEHOLDER_SHINY_APP_URL)
:::

We will explore:

1. **The Reinhart-Rogoff story** — interactive error explorer
2. **PSA Demo** — 10,000 simulations, cost-effectiveness plane, one click
3. **Markov Model** — state transitions with live parameter tuning
4. **CEAC Explorer** — acceptability curves at different WTP thresholds
5. **R vs Excel** — side-by-side comparison

:::

::: {.notes}
Open the Shiny app in a browser tab. Walk through each tab live. Spend ~3 min on the R-R error explorer, ~3 min on PSA, ~3 min on Markov, ~2 min on CEAC. Return to slides for the wrap-up.
:::

## What You Just Saw

In the Shiny app, everything was:

. . .

**Reproducible** — same seed, same results, every time, anywhere

. . .

**Auditable** — every simulation row visible, every formula transparent

. . .

**Interactive** — change a parameter, see the result immediately

. . .

**Scalable** — 10,000 PSA iterations in a fraction of a second

. . .

::: {.key-insight}
The entire PSA logic was **3 lines of R**. The entire Markov model was **10 lines**. Try auditing that in Excel.
:::

## The Limitations You Already Know

::: {style="font-size: 0.95em;"}

**PSA is painful in Excel** — complex macros or VBA; spreadsheet becomes slow and fragile

. . .

**Reproducibility is difficult** — tracing nested cell references across multiple sheets

. . .

**Version control is manual** — Model_v3_final_FINAL_revised.xlsx

. . .

**Transparency for publication** — journals and HTA agencies increasingly require open-source models

. . .

**Collaboration is messy** — two people cannot work on the same Excel model simultaneously

:::

## What R Makes Possible

**Reproducible**
: Your entire analysis is a script — anyone can re-run it and get the same result

. . .

**Scalable**
: 10,000 PSA iterations in seconds, not hours

. . .

**Transparent**
: A reviewer can read your code and see exactly what you did

. . .

**Version-controlled**
: Every change tracked, documented, reversible with Git

. . .

**Free and extensible**
: No license fees — thousands of packages for health economics

## The HTA Workflow in R

```{r}
#| label: fig-hta-workflow
#| eval: true
#| echo: false
#| fig-cap: "The complete modelling workflow"

library(DiagrammeR)

grViz("
digraph hta_workflow {
  graph [rankdir=LR, bgcolor='transparent', fontname='Helvetica']
  node [shape=box, style='filled,rounded', fontname='Helvetica', fontsize=11]

  A [label='Define\nParameters', fillcolor='#4e79a7', fontcolor='white']
  B [label='Build\nModel', fillcolor='#59a14f', fontcolor='white']
  C [label='Run\nAnalysis', fillcolor='#f28e2b', fontcolor='white']
  D [label='Sensitivity\nAnalysis', fillcolor='#e15759', fontcolor='white']
  E [label='Visualise\nResults', fillcolor='#76b7b2', fontcolor='white']
  F [label='Report &\nPublish', fillcolor='#edc948', fontcolor='black']

  A -> B -> C -> D -> E -> F
  D -> B [style=dashed, label='Iterate', fontsize=9]
}
")
```

## Four Model Types We'll Cover

```{r}
#| label: fig-four-models
#| eval: true
#| echo: false
#| fig-cap: "HTA model types in this workshop"

grViz("
digraph four_models {
  graph [rankdir=TB, bgcolor='transparent', fontname='Helvetica']
  node [shape=box, style='filled,rounded', fontname='Helvetica', fontsize=11, width=2.5]

  center [label='HTA Models in R', shape=ellipse, fillcolor='#4e79a7', fontcolor='white', fontsize=13]

  dt_diag [label='Diagnostic\nDecision Tree\n(GDM)', fillcolor='#59a14f', fontcolor='white']
  dt_ther [label='Therapeutic\nDecision Tree\n(DES vs BMS)', fillcolor='#f28e2b', fontcolor='white']
  markov [label='Markov Cohort\nModel\n(CKD)', fillcolor='#e15759', fontcolor='white']
  psm [label='Partitioned Survival\nModel\n(Breast Cancer)', fillcolor='#76b7b2', fontcolor='white']

  center -> dt_diag
  center -> dt_ther
  center -> markov
  center -> psm
}
")
```

## By Day 3, You Will Build This

::: {style="text-align: center; margin-bottom: 10px; font-weight: bold; font-size: 1.1em; color: #028090;"}
An interactive Shiny app — change parameters with sliders, see results update live
:::

```{=html}
<iframe src="https://YOUR-USERNAME.shinyapps.io/markov-model/"
        width="1100" height="480"
        style="border: 2px solid #02C39A; border-radius: 8px; background: white;"
        allow="fullscreen">
  <p style="text-align:center; color: #999; padding: 40px;">
    Shiny app will load here — requires deployment to shinyapps.io
  </p>
</iframe>
```

::: {.notes}
Replace the iframe src with your actual shinyapps.io URL once deployed.
:::

## What This Workshop Will NOT Do

- Turn you into R programmers in three days
- Ask you to memorise syntax
- Expect you to write complex code from scratch

## What This Workshop WILL Do

- Give you **working templates** you can adapt for your own studies
- Build confidence in **reading and modifying** R code
- Show how to use **AI tools** to support your workflow
- Help you understand **when R is the right tool**

## Key Takeaway

::: {style="text-align: center; padding-top: 30px; font-size: 1.1em;"}

R is not replacing your HTA expertise.

. . .

It is giving that expertise **a more powerful medium**.

. . .

The clinical judgement, model structure, parameter selection — **that remains yours**.

. . .

R simply makes the execution **more reproducible, transparent, and scalable**.

. . .

::: {style="font-size: 1.3em; color: #028090; font-weight: bold; margin-top: 20px;"}
Let us get started.
:::

:::
