---
title: "Session 1: The Case for R in HTA"
subtitle: "Why transition from Excel and TreeAge?"
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
---

## Why Are We Here?

You have been doing HTA work — and doing it well — using Excel spreadsheets and TreeAge. These are familiar, well-established tools. So why consider R?

This session does not ask you to abandon what works. It asks you to consider **what becomes possible** when you add R to your toolkit.

## The Limitations You Already Know

If you have built cost-effectiveness models in Excel, you have likely encountered some of these challenges:

- **Running PSA is painful.** Setting up 1,000 or 10,000 Monte Carlo iterations in Excel requires complex macros or VBA, and the spreadsheet becomes slow and fragile.
- **Reproducibility is difficult.** When a reviewer asks "how exactly did you calculate this ICER?", tracing through nested cell references across multiple sheets is tedious — for you and for them.
- **Version control is manual.** Model_v3_final_FINAL_revised.xlsx is not an unfamiliar filename.
- **Transparency for publication.** Journals and HTA agencies (including NICE, WHO, and increasingly Indian HTA bodies) are moving toward requiring open-source, auditable models.
- **Collaboration is messy.** Two people cannot easily work on the same Excel model simultaneously without overwriting each other's changes.

TreeAge solves some of these problems but introduces others: proprietary software, licensing costs, limited customisation, and difficulty integrating with modern data analysis workflows.

## What R Makes Possible

R does not replace your HTA knowledge. It gives you a **different medium** to express that knowledge — one that is:

- **Reproducible**: Your entire analysis is a script. Anyone can re-run it and get the same result.
- **Scalable**: Running 10,000 PSA iterations takes seconds, not hours.
- **Transparent**: A reviewer can read your code and see exactly what you did.
- **Version-controlled**: With Git and GitHub, every change is tracked, documented, and reversible.
- **Free and open-source**: No license fees, no vendor lock-in.
- **Extensible**: Thousands of packages built specifically for health economics, survival analysis, Bayesian methods, and visualisation.

```{r}
#| label: fig-hta-workflow
#| echo: true
#| fig-cap: "The HTA modelling workflow in R"

library(DiagrammeR)

grViz("
digraph hta_workflow {
  graph [rankdir=LR, bgcolor='transparent', fontname='Helvetica']
  node [shape=box, style='filled,rounded', fontname='Helvetica', fontsize=11]

  A [label='Define\nParameters', fillcolor='#4e79a7', fontcolor='white']
  B [label='Build\nModel', fillcolor='#59a14f', fontcolor='white']
  C [label='Run\nAnalysis', fillcolor='#f28e2b', fontcolor='white']
  D [label='Sensitivity\nAnalysis', fillcolor='#e15759', fontcolor='white']
  E [label='Visualise\nResults', fillcolor='#76b7b2', fontcolor='white']
  F [label='Report &\nPublish', fillcolor='#edc948', fontcolor='black']

  A -> B -> C -> D -> E -> F
  D -> B [style=dashed, label='Iterate', fontsize=9]
}
")
```

## A Simple Comparison: Excel Logic vs R

Let us take a very simple example — calculating the **incremental cost-effectiveness ratio (ICER)** for two interventions — and see how it looks in both approaches.

### The Problem

A new screening programme for a disease costs ₹2,500 per person and produces 0.85 QALYs on average. The current standard of care costs ₹1,200 per person and produces 0.78 QALYs. What is the ICER?

### In Excel

You would set up cells like this:

| | A | B |
|---|---|---|
| 1 | **Intervention** | **Standard Care** |
| 2 | Cost (₹) | 2500 |
| 3 | | 1200 |
| 4 | QALYs | 0.85 |
| 5 | | 0.78 |
| 6 | Incremental Cost | =B2-B3 |
| 7 | Incremental QALYs | =B4-B5 |
| 8 | ICER | =B6/B7 |

Result: ICER = ₹18,571 per QALY.

This is fine for one calculation. But what if you want to vary the costs and QALYs probabilistically across 10,000 simulations?

### In R

```{r}
#| label: simple-icer
#| echo: true

# Define inputs
cost_intervention <- 2500
cost_standard     <- 1200
qaly_intervention <- 0.85
qaly_standard     <- 0.78

# Calculate ICER
incremental_cost  <- cost_intervention - cost_standard
incremental_qaly  <- qaly_intervention - qaly_standard
icer <- incremental_cost / incremental_qaly

# Display result
cat("Incremental Cost: ₹", incremental_cost, "\n")
cat("Incremental QALYs:", incremental_qaly, "\n")
cat("ICER: ₹", round(icer, 0), "per QALY\n")
```

So far, this is not obviously better than Excel. The advantage shows when we add **probabilistic analysis**:

```{r}
#| label: simple-psa
#| echo: true

# Set number of simulations
n_sim <- 10000
set.seed(42)  # For reproducibility

# Probabilistic inputs — sample from distributions
cost_intervention_psa <- rnorm(n_sim, mean = 2500, sd = 300)
cost_standard_psa     <- rnorm(n_sim, mean = 1200, sd = 150)
qaly_intervention_psa <- rbeta(n_sim, shape1 = 85, shape2 = 15)
qaly_standard_psa     <- rbeta(n_sim, shape1 = 78, shape2 = 22)

# Calculate 10,000 ICERs
inc_cost <- cost_intervention_psa - cost_standard_psa
inc_qaly <- qaly_intervention_psa - qaly_standard_psa
icer_psa <- inc_cost / inc_qaly

# Summary
cat("Mean ICER across", n_sim, "simulations: ₹", round(mean(icer_psa), 0), "per QALY\n")
cat("95% credible interval: ₹", round(quantile(icer_psa, 0.025), 0),
    "to ₹", round(quantile(icer_psa, 0.975), 0), "\n")
```

**10,000 simulations in a fraction of a second.** Try doing that in Excel.

Now let us visualise the results — something that would require significant manual effort in Excel:

```{r}
#| label: ce-plane-simple
#| echo: true
#| fig-cap: "Cost-effectiveness plane from 10,000 PSA iterations"

library(ggplot2)

ce_data <- data.frame(
  inc_cost = inc_cost,
  inc_qaly = inc_qaly
)

ggplot(ce_data, aes(x = inc_qaly, y = inc_cost)) +
  geom_point(alpha = 0.1, colour = "steelblue", size = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    x = "Incremental QALYs",
    y = "Incremental Cost (₹)",
    title = "Cost-Effectiveness Plane",
    subtitle = "10,000 PSA iterations — New Screening vs Standard Care"
  ) +
  theme_minimal()
```

```{r}
#| label: simple-ceac
#| echo: true
#| fig-cap: "Cost-effectiveness acceptability curve — probability of being cost-effective at different WTP thresholds"

wtp_range <- seq(0, 300000, by = 5000)
prob_ce <- sapply(wtp_range, function(wtp) {
  mean(inc_cost <= wtp * inc_qaly)
})

ceac_df <- data.frame(WTP = wtp_range, Probability = prob_ce)

ggplot(ceac_df, aes(x = WTP, y = Probability)) +
  geom_line(colour = "steelblue", linewidth = 1.2) +
  geom_ribbon(aes(ymin = 0, ymax = Probability), fill = "steelblue", alpha = 0.2) +
  geom_vline(xintercept = 100000, linetype = "dashed", colour = "red") +
  annotate("text", x = 110000, y = 0.3, label = "WTP = ₹1,00,000",
           colour = "red", size = 3, hjust = 0) +
  scale_x_continuous(labels = function(x) paste0("₹", format(x/1000, big.mark=","), "K")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  labs(x = "Willingness-to-Pay Threshold (₹/QALY)",
       y = "Probability Cost-Effective",
       title = "Cost-Effectiveness Acceptability Curve",
       subtitle = "From 10,000 PSA iterations") +
  theme_minimal()
```

## Beyond This Simple Example

In this workshop, you will see R applied to real HTA problems:

- **Decision trees** where you can instantly re-run the model with different prevalence estimates or test accuracy values
- **Markov models** where transition probabilities, costs, and utilities can all be varied probabilistically
- **Partitioned survival models** where you can fit and compare multiple survival distributions and see how extrapolation choices affect the ICER
- **Shiny apps** where you can build interactive tools that non-R-users (like policymakers) can use to explore your model

```{r}
#| label: fig-four-models
#| echo: true
#| fig-cap: "Four HTA model types covered in this workshop"

grViz("
digraph four_models {
  graph [rankdir=TB, bgcolor='transparent', fontname='Helvetica']
  node [shape=box, style='filled,rounded', fontname='Helvetica', fontsize=11, width=2.5]

  center [label='HTA Decision\nModels in R', shape=ellipse, fillcolor='#4e79a7', fontcolor='white', fontsize=13]

  dt_diag [label='Diagnostic\nDecision Tree\n(Session 3: GDM)', fillcolor='#59a14f', fontcolor='white']
  dt_ther [label='Therapeutic\nDecision Tree\n(Session 4: DES vs BMS)', fillcolor='#f28e2b', fontcolor='white']
  markov [label='Markov Cohort\nModel\n(Session 5: CKD)', fillcolor='#e15759', fontcolor='white']
  psm [label='Partitioned Survival\nModel\n(Session 8: Breast Cancer)', fillcolor='#76b7b2', fontcolor='white']

  center -> dt_diag
  center -> dt_ther
  center -> markov
  center -> psm
}
")
```

## What This Workshop Will NOT Do

We will not:

- Turn you into R programmers in three days
- Ask you to memorise syntax
- Expect you to write complex code from scratch

We **will**:

- Give you working templates you can adapt
- Build your confidence in reading and modifying R code
- Show you how to use available tools (including AI) to support your workflow
- Help you understand when R is the right tool for your HTA work

## Key Takeaway

R is not replacing your HTA expertise — it is giving that expertise a more powerful medium. The clinical judgement, the model structuring, the parameter selection — that remains yours. R simply makes the execution more reproducible, transparent, and scalable.

Let us get started.
