---
title: "Wrap-Up and Next Steps"
subtitle: "Where to go from here with R for HTA"
author: "R for HTA (Basics) — Workshop 2026"
institute: "RRC-HTA, AIIMS Bhopal | HTAIn, DHR"
format:
  revealjs:
    theme: [default, ../../custom-theme.scss]
    slide-number: true
    footer: "R for HTA (Basics) — RRC-HTA, AIIMS Bhopal"
    logo: ""
    transition: fade
    width: 1280
    height: 720
    code-fold: true
    code-overflow: wrap
    highlight-style: github
    fig-align: center
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

## What You Accomplished in 3 Days

**Day 1: Foundations + Decision Trees**
- R fundamentals (syntax, data types, visualization)
- GDM diagnostic decision tree
- TES therapeutic decision tree

. . .

**Day 2: Markov Models + PSA + GenAI**
- CKD Markov cohort model (4 states)
- Probabilistic sensitivity analysis
- How to use AI tools to help write code

. . .

**Day 3: PSM + Shiny Apps**
- Breast cancer partitioned survival model
- PSA for PSM (parameter distributions, CEAC, CE plane)
- Interactive Shiny app templates

:::notes
Recap the journey. In three days, participants went from R basics to building complete HTA models with interactive apps.
:::

## Four Model Types You've Mastered

```{r}
#| label: fig-model-types
#| eval: true
#| fig-cap: "Four model structures for different HTA questions"
#| fig-width: 10
#| fig-height: 6

library(DiagrammeR)

grViz("
digraph models {
  graph [rankdir=TB, bgcolor='transparent', fontname='Helvetica']
  node [fontname='Helvetica', fontsize=10, style='filled,rounded', shape=box, width=2.5]

  dt_diag [label='Diagnostic\nDecision Tree\n(GDM)', fillcolor='#59a14f', fontcolor='white']
  dt_ther [label='Therapeutic\nDecision Tree\n(Stents)', fillcolor='#f28e2b', fontcolor='white']
  markov [label='Markov\nCohort Model\n(CKD)', fillcolor='#e15759', fontcolor='white']
  psm [label='Partitioned\nSurvival Model\n(Breast Ca)', fillcolor='#76b7b2', fontcolor='white']

  workshop [label='This Workshop', shape=ellipse, fillcolor='#4e79a7', fontcolor='white', width=2]

  workshop -> dt_diag
  workshop -> dt_ther
  workshop -> markov
  workshop -> psm
}
")
```

. . .

**You can now:**
- Identify which model fits a given decision problem
- Build and validate each type
- Run sensitivity analyses and PSA
- Communicate results to stakeholders

:::notes
Summarize the scope. Four distinct model types, each useful for different clinical questions.
:::

## Key R Packages for HTA

| Package | Purpose |
|---------|---------|
| **heemod** | Markov models (easy syntax, automatic discounting) |
| **dampack** | Decision trees, sensitivity analyses (tornado plots, frontiers) |
| **BCEA** | Bayesian cost-effectiveness (CEAC, EVPI, credible regions) |
| **hesim** | Microsimulation & DES (complex state-transition models) |
| **flexsurv** | Parametric survival fitting (Weibull, log-logistic, etc.) |
| **survHE** | Survival analysis for HE (fit, compare, integrate) |

. . .

**You've already used the core concepts** in this workshop. These packages automate and streamline them.

:::notes
Introduce the major packages in the R-HTA ecosystem. Participants have done the manual version; now they know what these packages do.
:::

## The DARTH Group: Your Community

**DARTH** = Decision Analysis in R for Technologies in Health

. . .

They publish:
- **Tutorial papers** (Lancet, MDM, other journals)
- **GitHub repositories** with fully worked examples
- **Coding standards** for reproducible HTA

. . .

Find them: GitHub, DARTH website, MDM journal

. . .

**R-HTA Consortium:** Connect with other analysts, webinars, code sharing

:::notes
Point to existing communities and resources. Participants aren't alone; there's a growing community.
:::

## Your Learning Path Forward

```{r}
#| label: fig-learning-path
#| eval: true
#| fig-cap: "Progression from this workshop to mastery"
#| fig-width: 10
#| fig-height: 5

grViz("
digraph path {
  graph [rankdir=TB, bgcolor='transparent', fontname='Helvetica']
  node [fontname='Helvetica', fontsize=10, style='filled,rounded', shape=box, width=3]

  now [label='This Workshop\n(Read, modify, run R code)', fillcolor='#4e79a7', fontcolor='white']
  next1 [label='Next 1-2 weeks\n(Adapt templates for your problem)', fillcolor='#59a14f', fontcolor='white']
  next2 [label='Weeks 3-4\n(Add features: PSA, new outcomes)', fillcolor='#f28e2b', fontcolor='white']
  next3 [label='Months 2-3\n(Share Shiny app, deploy)', fillcolor='#e15759', fontcolor='white']
  community [label='Ongoing\n(DARTH, R-HTA Consortium,\npublish methods)', fillcolor='#76b7b2', fontcolor='white']

  now -> next1 -> next2 -> next3 -> community
}
")
```

:::notes
Give concrete milestones. Learning is a journey; outline realistic next steps.
:::

## Phase 1: Consolidate (Next 1-2 Weeks)

1. Take one workshop template
2. Modify it for a problem you care about
3. Change parameter values, time horizon, strategies
4. Run the model several times to understand assumptions
5. Get feedback from colleagues

. . .

**Goal:** Build confidence with the templates. Understand how each assumption drives results.

:::notes
Actionable first steps. Keep it concrete and achievable.
:::

## Phase 2: Expand (Weeks 3-4)

1. Identify a feature you want to add:
   - PSA (if starting from a deterministic model)
   - New outcome measure
   - Sensitivity analysis tornado plot

2. Use AI tools (ChatGPT, Claude) to help:
   - Show them the workshop code
   - Ask for modifications
   - They generate scaffolding; you refine

3. Read the relevant package vignette

4. Test and validate

:::notes
Practical guidance on using AI. Participants shouldn't be intimidated by code they don't yet understand.
:::

## Phase 3: Share (Months 2-3)

1. Deploy a Shiny app to shinyapps.io (free tier)
2. Share your Quarto report with stakeholders
3. Invite colleagues to explore and give feedback
4. Iterate on the model

. . .

**This is where impact happens.** Your models inform real decisions.

:::notes
Sharing is crucial. The model in a drawer helps no one.
:::

## Phase 4: Contribute (Ongoing)

- Join R-HTA Consortium or local R user group
- Attend webinars and workshops
- Help others by sharing your code
- Consider publishing methods in open-access journal

:::notes
The virtuous cycle. Learn, build, share, teach.
:::

## Four Model Types: A Decision Guide

**Your decision problem:...**

. . .

→ **Is it about diagnostic accuracy?** → Use **Diagnostic Decision Tree**
  (sensitivity, specificity, prevalence)

. . .

→ **Is it about a one-time treatment choice?** → Use **Therapeutic Decision Tree**
  (event probabilities, immediate costs/benefits)

. . .

→ **Does the disease progress over time?** → Use **Markov Cohort Model**
  (states, transition probabilities, cycles)

. . .

→ **Do you have survival curves from a trial?** → Use **Partitioned Survival Model**
  (OS and PFS, extrapolation)

:::notes
Practical decision tree for choosing model types. Participants should reference this.
:::

## Key Principles to Remember

**Transparency first.** Every assumption visible and adjustable. Avoid hiding logic.

. . .

**Start simple.** Simplest model that answers the question. Add complexity later.

. . .

**Validate early.** Check that your model reproduces expected results before publishing.

. . .

**Document as you go.** Clear comments explaining each modeling choice.

. . .

**Test your code.** A bug in a transition probability propagates through the analysis.

. . .

**Share your work.** HTA community improves when we share methods and code.

:::notes
Core principles that apply regardless of model type or package.
:::

## Publishing with R: Quarto for Reproducible Reports

Quarto weaves **R code and narrative** into a single reproducible document.

. . .

A typical HTA report:
- Executive summary
- Model structure explanation
- Embedded code chunks (transparent)
- Plots and tables generated directly from R
- Sensitivity analyses
- Conclusion

. . .

**Output:** HTML, PDF, Word, or website — write once, export everywhere.

. . .

**Benefit:** Reviewers can see every line of code that produced every number.

:::notes
Quarto is powerful for reproducibility. Many journals now encourage or require it.
:::

## You Don't Need to Be a Programmer

You need to be **an HTA analyst who has learned to use R as a tool**.

. . .

Just as you might use Excel, but better.

. . .

Your models will be:
- Reproducible (everyone can run the code)
- Transparent (all assumptions visible)
- Shareable (interactive apps, GitHub)

. . .

:::{.callout-important}
You've invested three days in learning these skills. That investment pays dividends.
:::

:::notes
Final encouragement. Participants shouldn't feel like they need to be programmers. They're analysts who happen to use R.
:::

## The R Community is Welcoming

When you have questions: **Ask them.**

When you get stuck: **Others have been stuck there too.**

When you solve a problem: **Share the solution.**

. . .

Resources:
- Stack Overflow (tag: r, health-economics)
- R-HTA Consortium forums
- GitHub Discussions
- DARTH tutorials
- Local R user groups

:::notes
Community support is real. Reduce isolation. Participants are joining a helpful ecosystem.
:::

## Resources to Dive Deeper

**Books:**
- *R for Data Science* by Hadley Wickham
- *Mastering Shiny* by Hadley Wickham
- *Decision Modelling for HE* by Briggs, Claxton, Sculpher

. . .

**Online:**
- DARTH tutorials (GitHub)
- R-HTA Consortium webinars
- Shiny official documentation

. . .

**Packages to explore:**
- `heemod` vignette
- `dampack` vignette
- `BCEA` vignette

:::notes
Provide a reading list for self-directed learning.
:::

## ParCC: Your Tool from RRC-HTA

[Information about ParCC — the platform or tool your organization offers]

. . .

Check the workshop GitHub repository for:
- All slide decks (reveal.js)
- All R code and Quarto documents
- All Shiny app templates
- Sample data and results

:::notes
Reference any institutional tools or resources. Direct participants back to the workshop repository.
:::

## Your Next Steps (Concrete)

**This week:**
1. Run one of the workshop apps locally
2. Modify a parameter and observe the change
3. Understand how the code works (read comments, not every line)

. . .

**Next week:**
1. Adapt a template for a problem at your organization
2. Change parameter values to match your data
3. Share results with a colleague

. . .

**Next month:**
1. Add a feature (plot, sensitivity analysis, PSA)
2. Deploy a Shiny app
3. Get feedback

:::notes
Concrete, achievable milestones. Reduce overwhelm. One step at a time.
:::

## Thank You!

You've worked hard over three days. You now have the skills to:
- **Build** health technology assessment models in R
- **Validate** and **test** them rigorously
- **Communicate** results interactively to stakeholders
- **Contribute** to reproducible, transparent decision-making in health

. . .

**Go build something great.**

. . .

Questions? Contact:
- **RRC-HTA, AIIMS Bhopal** [email/contact]
- **HTAIn, Department of Health Research** [email/contact]

:::notes
Final thanks and inspiration. End on a high note.
:::

## Workshop Materials

All materials (slides, code, templates, data) available at:

**GitHub:** [repository URL]

**Local path:** `/day3/`

. . .

**How to use the repository:**
1. Fork it (create your own copy)
2. Clone to your computer
3. Explore and modify the templates
4. Share improvements via pull request

. . .

Happy coding!

:::notes
Direct participants to the resource repository. Make it easy to access and use materials after the workshop.
:::

