---
title: "Interactive Shiny App Templates"
subtitle: "Turning your HTA models into tools anyone can use"
author: "R for HTA (Basics) — Workshop 2026"
institute: "RRC-HTA, AIIMS Bhopal | HTAIn, DHR"
format:
  revealjs:
    theme: [default, ../../custom-theme.scss]
    slide-number: true
    footer: "R for HTA (Basics) — RRC-HTA, AIIMS Bhopal"
    logo: ""
    transition: fade
    width: 1280
    height: 720
    code-fold: true
    code-overflow: wrap
    highlight-style: github
    fig-align: center
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

## What is Shiny?

Shiny is an R package that builds **interactive web applications** directly from R code.

. . .

**No HTML, CSS, or JavaScript required.**

. . .

Your HTA model (thousands of lines of R) → wrapped in a friendly UI → anyone can use in a browser.

. . .

:::{.callout-tip}
Instead of sharing Excel with hidden formulas, you create an app where stakeholders adjust sliders and see results update instantly.
:::

:::notes
Introduce Shiny as the bridge between complex R models and non-technical users. Emphasize no web dev skills needed.
:::

## Why Shiny for HTA?

Health technology assessment is fundamentally about **"What if?" questions**.

. . .

- What if treatment cost goes down?
- What if efficacy improves?
- What if we target a high-risk subgroup?

. . .

**Shiny apps let decision-makers explore these scenarios in real time.**

. . .

**Bonus:** Transparency and reproducibility built in. Every calculation is real R code, not hidden Excel cells.

:::notes
Emphasize the value: interactivity + transparency. Stakeholders see exactly how you arrived at each number.
:::

## The 4 Shiny Apps in This Workshop

1. **GDM Diagnostic Tree** — 3 strategies, NMB ranking, real tornado DSA
2. **DES vs BMS** — NPPA 2025 pricing, cost breakdown, break-even analysis
3. **CKD Markov Model** — Trace plots, tornado DSA, **PSA with CE plane + CEAC**
4. **Breast Cancer PSM** — Weibull survival, distribution comparison, NMB

. . .

Each is **complete, ready-to-run, and easy to modify.**

## Try the Live Apps Now

| App | Link |
|-----|------|
| GDM Diagnostic Tree | [Open](https://019cf9b2-a07c-c07e-5962-1551244e5a72.share.connect.posit.cloud/) |
| DES vs BMS | [Open](https://019cf9b5-46ac-590e-4233-0f24f96d7257.share.connect.posit.cloud/) |
| CKD Markov (with PSA) | [Open](https://019cf9b3-3d70-a837-f435-c349bbc8cb98.share.connect.posit.cloud/) |
| Breast Cancer PSM | [Open](https://019cf9b6-102e-3d69-c6d2-15e4a916bfb4.share.connect.posit.cloud/) |

. . .

**Try:** Adjust sliders → watch ICER/NMB update. In the Markov app, click "Run PSA" for CE plane + CEAC.

## Anatomy of a Shiny App

Every Shiny app has **three core parts:**

```{r}
#| label: fig-shiny-anatomy
#| eval: true
#| fig-cap: "How Shiny connects the user to your HTA model"
#| fig-width: 10
#| fig-height: 5

library(DiagrammeR)

grViz("
digraph shiny_anatomy {
  graph [rankdir=LR, bgcolor='transparent', fontname='Helvetica']
  node [fontname='Helvetica', fontsize=11, style='filled,rounded', shape=box]

  user [label='User\n(Browser)', fillcolor='#76b7b2', fontcolor='white',
        shape=ellipse, width=1.5]

  subgraph cluster_ui {
    label='UI (Frontend)'
    style=filled
    fillcolor='#d4e6f1'

    inputs [label='Input Widgets\n(sliders, dropdowns,\ntext boxes)',
            fillcolor='#4e79a7', fontcolor='white']
    outputs [label='Output Displays\n(plots, tables,\nresults)',
             fillcolor='#4e79a7', fontcolor='white']
  }

  subgraph cluster_server {
    label='Server (R Logic)'
    style=filled
    fillcolor='#d5f5e3'

    model [label='HTA Model\n(tree, Markov,\nPSM, etc)',
           fillcolor='#59a14f', fontcolor='white']
    render [label='Render Outputs\n(create plots,\ncompute ICER)',
            fillcolor='#59a14f', fontcolor='white']
  }

  user -> inputs [label='Adjusts\nsliders']
  inputs -> model [label='Send\nparameters']
  model -> render [label='Calculate']
  render -> outputs [label='Update']
  outputs -> user [label='Display\nresults']
}
")
```

:::notes
Visual walkthrough of the Shiny architecture. User → UI inputs → Server logic → HTA model → Server rendering → UI outputs → User sees results.
:::

## The Three Parts: UI, Server, shinyApp()

```{r}
#| code-fold: false

library(shiny)

# 1. DEFINE USER INTERFACE
ui <- fluidPage(
  titlePanel("My First HTA App"),
  sidebarLayout(
    sidebarPanel(
      sliderInput("cohort_size", "Cohort Size:",
                  min = 100, max = 10000, value = 1000)
    ),
    mainPanel(
      textOutput("result")
    )
  )
)

# 2. DEFINE SERVER LOGIC
server <- function(input, output) {
  output$result <- renderText({
    total_cost <- input$cohort_size * 50000
    paste("Total cost for cohort: ₹", format(total_cost, big.mark=","))
  })
}

# 3. COMBINE UI AND SERVER
# shinyApp(ui = ui, server = server)
# (Don't actually run here, just show the pattern)
```

. . .

**Key idea:** When user moves slider, `input$cohort_size` updates → server re-runs → `output$result` updates → browser shows new result.

:::notes
Walk through the three-part structure. Emphasize the reactivity: changing an input automatically triggers recalculation.
:::

## Running a Shiny App Locally

Open an app's `app.R` file in RStudio:

. . .

**Option 1:** Click the **"Run App"** button in the script pane

. . .

**Option 2:** Type in console:
```r
shiny::runApp("path/to/app.R")
```

. . .

The app launches in your default browser. Press Escape or close the browser to stop.

:::notes
Simple instructions. Getting started with Shiny is frictionless.
:::

## Running a Shiny App From This Workshop

```bash
# Navigate to the app directory
cd day3/session10-shiny-apps/gdm-decision-tree-app

# In RStudio, open app.R and click "Run App"
# OR in R console:
shiny::runApp()
```

. . .

Explore:
- Adjust sliders (probabilities, costs)
- See plots and tables update in real time
- Understand how each parameter affects the ICER

:::notes
Practical guidance. Make it concrete by referring to the workshop apps.
:::

## Key UI Widgets (Inputs)

```{r}
#| code-fold: false

# SLIDER: continuous parameter
sliderInput("discount_rate", "Discount Rate (%):",
            min = 0, max = 10, value = 3, step = 0.1)

# NUMERIC INPUT: single value
numericInput("cohort_size", "Cohort Size:",
             value = 1000, min = 1)

# DROPDOWN: choose one option
selectInput("distribution", "Survival Distribution:",
            choices = c("Weibull", "Exponential", "Log-logistic"),
            selected = "Weibull")

# RADIO BUTTONS: choose one (visible)
radioButtons("currency", "Currency:",
             choices = c("INR" = "inr", "USD" = "usd"),
             selected = "inr")

# CHECKBOX: toggle on/off
checkboxInput("include_psa", "Include PSA?", value = FALSE)
```

:::notes
Show common input widgets. Each has `inputId` (how you reference it in server) and appearance options.
:::

## Key Server Functions (Outputs)

```{r}
#| code-fold: false

# RENDER TEXT
output$summary <- renderText({
  paste("ICER: ₹", format(round(icer()), big.mark=","), "/QALY")
})

# RENDER PLOT
output$ce_plane <- renderPlot({
  ggplot(ce_data, aes(x = inc_qaly, y = inc_cost)) +
    geom_point(alpha = 0.5) +
    geom_abline(intercept = 0, slope = input$wtp_threshold)
    # Note: input$wtp_threshold updates automatically
})

# RENDER TABLE
output$results_table <- renderTable({
  data.frame(
    Strategy = c("Control", "Intervention"),
    Cost = c(cost_control, cost_intervention),
    QALYs = c(qaly_control, qaly_intervention)
  )
})
```

. . .

**Reactivity:** When `input$wtp_threshold` changes, `renderPlot()` re-runs automatically.

:::notes
Show how server functions work. The key is reactivity: when inputs change, outputs recalculate.
:::

## Example: Simple CKD Markov App

```{r}
#| code-fold: false

# Sidebar inputs
sidebarPanel(
  titlePanel("CKD Markov Model"),
  sliderInput("baseline_cost", "Annual cost (Stage 1):",
              min = 5000, max = 50000, value = 20000),
  sliderInput("treatment_effect", "Treatment effect (%):",
              min = 0, max = 50, value = 20)
)

# Main panel outputs
mainPanel(
  tabsetPanel(
    tabPanel("Results",
             textOutput("icer_summary"),
             tableOutput("cost_table")),
    tabPanel("Sensitivity Analysis",
             plotOutput("tornado_plot")),
    tabPanel("Cohort Trace",
             plotOutput("state_occupancy"))
  )
)
```

. . .

This structure is identical across all four workshop apps. **Only the HTA model logic changes.**

:::notes
Show a typical app layout. Sidebar for inputs, main panel with tabs for different outputs. This layout is standard.
:::

## Modifying the Apps

The apps are **designed to be adapted**:

. . .

**Want to add a parameter?**
1. Add a `sliderInput()` in the UI
2. Use `input$parameter_name` in the server calculation
3. Re-run the app

. . .

**Want to change a plot theme?**
→ Edit the `ggplot()` code directly

. . .

**Want to add a new output?**
→ Add a `renderPlot()` or `renderTable()` in the server, then reference it in the UI

. . .

:::{.callout-note}
You don't need to understand all the Shiny internals. Just focus on: inputs, model logic, outputs.
:::

:::notes
Empower participants. The apps are templates, not black boxes. They can modify them confidently.
:::

## Sharing Shiny Apps

**Option 1: shinyapps.io (Free)**
- Cloud hosting (Posit/RStudio account)
- Deploy with one click: `rsconnect::deployApp()`
- Others access via URL in browser

. . .

**Option 2: Shiny Server (Self-hosted)**
- Host on your organization's server
- Control who has access
- More control, more setup

. . .

**Option 3: Share the app.R file**
- Email or GitHub
- Others run locally with `shiny::runApp()`
- Simplest; no server needed

. . .

For this workshop: Start with **Option 3** (share app.R file).

:::notes
Give practical guidance on deployment. Option 3 is easiest for getting started; others are for wider sharing.
:::

## Modifying Apps for Your Organization

Change **parameter defaults and ranges** to match your data:

```{r}
#| code-fold: false

# BEFORE: Generic defaults
sliderInput("treatment_cost", "Annual treatment cost:",
            min = 10000, max = 500000, value = 100000)

# AFTER: Your organization's estimates
sliderInput("treatment_cost", "Annual trastuzumab cost (INR):",
            min = 350000, max = 500000,
            value = 420000,  # Your base case
            step = 10000)
```

. . .

This is **your model**. Own it. Change assumptions to reflect your beliefs.

:::notes
Emphasize ownership. The apps are starting points; participants should customize them for their context.
:::

## Don't Build from Scratch

:::{.callout-important}
**This session is NOT about learning to code Shiny from the ground up.**

You are learning to:
1. Understand how provided apps work
2. Identify which template matches your problem
3. Modify parameters and ranges
4. Add/remove outputs (plots, tables)
:::

. . .

If you want to dive deeper later: Read *Mastering Shiny* by Hadley Wickham.

For now: **Use the templates.**

:::notes
Set expectations. Focus is on using and customizing existing apps, not building from first principles.
:::

## The Power of Interactive Models

Without Shiny:
- Share a PDF report
- "If you want to change an assumption, email me"
- Decision-makers are passive

. . .

With Shiny:
- Deploy an interactive app
- Decision-makers explore scenarios themselves
- Trust increases; decisions are informed

. . .

:::{.callout-tip}
By publishing your Shiny app, you are **democratizing health technology assessment**.
:::

:::notes
Show the transformative potential. Interactive models engage stakeholders and increase buy-in.
:::

## Summary

✓ Shiny wraps your HTA model in a **user-friendly web interface**

✓ **Three-part structure:** UI (inputs) → Server (logic) → shinyApp()

✓ **No HTML/CSS/JavaScript needed** — it's all R

✓ Easy to modify: **change sliders, add plots, adjust assumptions**

✓ Easy to share: **email app.R or deploy to cloud**

. . .

Next: Wrap-up everything you've learned in 3 days (Session 11)

:::notes
Quick summary. Reinforce that Shiny is accessible and powerful.
:::

## References

- Wickham H. (2021). *Mastering Shiny*. O'Reilly Media. [Free online at mastering-shiny.org]
- Shiny Official Documentation: https://shiny.rstudio.com/
- DARTH Group Shiny examples on GitHub

