---
title: "Setup Guide"
subtitle: "Installing R, RStudio, and required packages"
format:
  html:
    toc: true
---

## Step 1: Install R

Download and install R from [CRAN](https://cran.r-project.org/):

- **Windows:** Click "Download R for Windows" → "base" → download the `.exe` file
- **Mac:** Click "Download R for macOS" → download the `.pkg` file
- **Linux:** Follow instructions for your distribution

**Recommended version:** R 4.3.0 or later.

## Step 2: Install RStudio

Download RStudio Desktop (free) from [Posit](https://posit.co/download/rstudio-desktop/).

Install and open RStudio. You should see four panels as described in Session 2.

## Step 3: Install Required Packages

Open RStudio and run the following code in the Console (bottom-left panel):

```{r}
#| label: install-packages
#| echo: true
#| eval: false

# Core packages for the workshop
install.packages(c(
  # Data manipulation and visualization
  "tidyverse",
  "ggplot2",
  "knitr",
  "DT",
  "DiagrammeR",

  # Health economics modelling
  "heemod",
  "dampack",
  "BCEA",

  # Survival analysis
  "survival",
  "survHE",
  "flexsurv",

  # Interactive apps
  "shiny",
  "shinydashboard",

  # Document rendering
  "quarto",
  "rmarkdown"
))
```

This may take 10-15 minutes depending on your internet connection.

## Step 4: Verify Installation

Run this code to check that all packages installed correctly:

```{r}
#| label: verify
#| echo: true
#| eval: false

packages <- c("tidyverse", "ggplot2", "knitr", "DT", "DiagrammeR", "heemod",
               "dampack", "BCEA", "survival", "survHE", "flexsurv",
               "shiny", "shinydashboard", "quarto", "rmarkdown")

for (pkg in packages) {
  if (require(pkg, character.only = TRUE, quietly = TRUE)) {
    cat("[OK]", pkg, "installed successfully\n")
  } else {
    cat("[FAIL]", pkg, "-- try reinstalling with install.packages('", pkg, "')\n")
  }
}
```

## Step 5: Open the Workshop Project

1. Download or clone this repository
2. In RStudio, go to **File → Open Project**
3. Navigate to the workshop folder and open `HTA-R-Workshop-2026.Rproj`
4. You are ready to begin!

## Troubleshooting

**"Package not available" error:** Your R version may be too old. Update R to version 4.3 or later.

**Installation hangs:** Try installing packages one at a time instead of all at once.

**Behind a proxy/firewall:** Ask your IT department for proxy settings, then run:
```r
Sys.setenv(http_proxy = "http://your-proxy:port")
Sys.setenv(https_proxy = "http://your-proxy:port")
```
