---
title: "Session 5 Bonus: CKD Markov Model with heemod"
subtitle: "The same model rebuilt with a dedicated Markov modelling package"
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
---

## Why a Package?

In the main Session 5, you built a CKD Markov model from scratch — writing matrix multiplication loops, half-cycle correction, and discounting by hand. That transparency is essential for understanding *what* a Markov model does.

But in practice, dedicated packages handle the mechanics for you and add built-in features for sensitivity analysis, plotting, and reporting. For Markov cohort models in R, **heemod** (Health Economic Evaluation MODelling) is the standard package. It was designed specifically for HTA and handles states, transitions, costs, utilities, discounting, half-cycle correction, DSA, and PSA out of the box.

This page rebuilds the identical CKD model using `heemod`, so you can see the same results produced by a different engine — and learn the package workflow you'd use in real projects.

```{r}
#| label: setup
#| echo: true
#| message: false
#| warning: false

# Install if needed (uncomment once)
# install.packages("heemod")

library(heemod)
library(ggplot2)
```

## 1. Define Parameters

In `heemod`, all model parameters go into a single `define_parameters()` call. This makes it easy to modify them later for sensitivity analysis — the package knows which values are inputs vs calculations.

```{r}
#| label: parameters
#| echo: true

param <- define_parameters(
  # --- Transition Probabilities: Standard Care ---
  p_12_std = 0.10,    # Early → Moderate
  p_1d     = 0.02,    # Early → Death (same both arms)
  p_23_std = 0.12,    # Moderate → Advanced
  p_2d     = 0.04,    # Moderate → Death (same both arms)
  p_3d     = 0.15,    # Advanced → Death (same both arms)

  # --- Treatment Effect ---
  hr = 0.66,           # Hazard ratio (ACE-inhibitor)
  p_12_int = 1 - (1 - p_12_std)^hr,
  p_23_int = 1 - (1 - p_23_std)^hr,

  # --- Costs (₹ per year) ---
  cost_early    = 12000,
  cost_moderate = 45000,
  cost_advanced = 350000,
  cost_ace      = 6000,
  cost_screen   = 500,

  # --- Utilities ---
  u_early    = 0.85,
  u_moderate = 0.72,
  u_advanced = 0.55,
  u_death    = 0.00,

  # --- Discount Rate ---
  dr = 0.03
)
```

::: {.callout-tip}
## heemod Parameters Are Lazy
Parameters defined with `define_parameters()` can reference each other. The intervention transition probabilities (`p_12_int`, `p_23_int`) are computed from the standard care values and the hazard ratio — just like in the manual approach. If you change `hr`, both intervention probabilities update automatically.
:::

## 2. Define Health States

Each health state gets its own `define_state()` call specifying the cost and utility per cycle. These are the state rewards — what accrues each year a patient spends in that state.

```{r}
#| label: states
#| echo: true

# --- States for STANDARD CARE ---
s_early_std <- define_state(
  cost = cost_early,
  utility = u_early
)

s_moderate_std <- define_state(
  cost = cost_moderate,
  utility = u_moderate
)

s_advanced_std <- define_state(
  cost = cost_advanced,
  utility = u_advanced
)

s_death <- define_state(
  cost = 0,
  utility = u_death
)

# --- States for INTERVENTION (ACE-inhibitor added to Early & Moderate) ---
s_early_int <- define_state(
  cost = cost_early + cost_ace,
  utility = u_early
)

s_moderate_int <- define_state(
  cost = cost_moderate + cost_ace,
  utility = u_moderate
)

s_advanced_int <- define_state(
  cost = cost_advanced,   # ACE-inhibitor usually stopped in advanced CKD
  utility = u_advanced
)
```

## 3. Define Transition Matrices

`define_transition()` builds the matrix row-by-row. The special value `C` means "complement" — heemod automatically computes the diagonal (stay) probability so each row sums to 1.

```{r}
#| label: transitions
#| echo: true

# --- Standard Care ---
tm_std <- define_transition(
  state_names = c("EarlyCKD", "ModerateCKD", "AdvancedDialysis", "Death"),

  C,          p_12_std, 0,         p_1d,     # From Early
  0,          C,        p_23_std,  p_2d,     # From Moderate
  0,          0,        C,         p_3d,     # From Advanced
  0,          0,        0,         1         # Death (absorbing)
)

# --- Intervention ---
tm_int <- define_transition(
  state_names = c("EarlyCKD", "ModerateCKD", "AdvancedDialysis", "Death"),

  C,          p_12_int, 0,         p_1d,
  0,          C,        p_23_int,  p_2d,
  0,          0,        C,         p_3d,
  0,          0,        0,         1
)

cat("Standard Care Transition Matrix:\n")
tm_std
```

::: {.callout-tip}
## The `C` Shorthand
The `C` (complement) value is heemod's most convenient feature. Instead of manually typing `1 - p_12 - p_1d` for the stay probability, you write `C` and the package handles the arithmetic. This eliminates a common source of bugs — forgetting to include all outgoing transitions when computing the "stay" probability.
:::

## 4. Define Strategies

A strategy combines a transition matrix with its state definitions. Each strategy represents one arm of the analysis.

```{r}
#| label: strategies
#| echo: true

strat_standard <- define_strategy(
  transition = tm_std,
  EarlyCKD         = s_early_std,
  ModerateCKD      = s_moderate_std,
  AdvancedDialysis = s_advanced_std,
  Death            = s_death
)

strat_intervention <- define_strategy(
  transition = tm_int,
  EarlyCKD         = s_early_int,
  ModerateCKD      = s_moderate_int,
  AdvancedDialysis = s_advanced_int,
  Death            = s_death
)
```

## 5. Run the Model

`run_model()` simulates both strategies simultaneously. We specify 20 annual cycles, the cost and effect (utility) variable names, initial state distribution, and the correction method. In heemod, `method = "life-table"` applies half-cycle correction (averaging start and end of each cycle), equivalent to the manual `(trace[t] + trace[t+1]) / 2` we used earlier.

```{r}
#| label: run-model
#| echo: true

res <- run_model(
  standard     = strat_standard,
  intervention = strat_intervention,
  parameters   = param,
  cycles       = 20,
  cost         = cost,
  effect       = utility,
  init         = c(10000, 0, 0, 0),
  method       = "life-table"
)

summary(res,
        threshold = 170000)
```

::: {.callout-note}
## One-Time Screening Cost
The one-time screening cost (₹500 per person = ₹50,00,000 for 10,000 cohort) is not included in the `run_model()` output because it's not a recurring state cost. When comparing with the manual R model, add ₹50,00,000 to the intervention's total cost. In a full analysis, you would handle this with a time-varying cost using `model_time == 1`.
:::

## 6. Cross-Validation Against Manual R

Let's compare the heemod results with our manual implementation:

```{r}
#| label: cross-validation
#| echo: true

# Extract heemod results
heemod_summary <- summary(res)

cat("=== Cross-Validation ===\n\n")
cat("Compare these heemod values with the manual R model and Excel:\n\n")
cat("If all three implementations agree (within rounding),\n")
cat("the model is verified across independent platforms.\n\n")
cat("Note: Add ₹50,00,000 one-time screening cost to\n")
cat("the intervention total for exact comparison.\n")
```

## 7. Markov Trace Plot

`heemod` provides built-in plotting for the Markov trace — the cohort distribution across states over time.

```{r}
#| label: fig-markov-trace
#| echo: true
#| fig-cap: "heemod built-in Markov trace — cohort distribution over 20 years"

plot(res, type = "counts", panel = "by_strategy") +
  theme_minimal() +
  labs(title = "Markov Trace: CKD Progression Over 20 Years",
       subtitle = "Built-in heemod visualisation")
```

```{r}
#| label: fig-state-values
#| echo: true
#| fig-cap: "Cost and utility accrual by state over time"

plot(res, type = "values", panel = "by_strategy") +
  theme_minimal()
```

## 8. One-Way Deterministic Sensitivity Analysis (DSA)

`define_dsa()` specifies which parameters to vary and their ranges. `run_dsa()` runs the model once per parameter at each extreme. The result is a tornado diagram showing which parameters drive the ICER most.

```{r}
#| label: dsa
#| echo: true
#| fig-cap: "Tornado diagram: one-way DSA on key parameters"

dsa_def <- define_dsa(
  p_12_std,      0.05,   0.15,     # Early→Moderate: ±50%
  p_23_std,      0.06,   0.18,     # Moderate→Advanced: ±50%
  p_3d,          0.10,   0.25,     # Advanced→Death: range
  hr,            0.50,   0.90,     # Treatment effect: wide range
  cost_early,    6000,   24000,    # ±100%
  cost_moderate, 25000,  75000,    # ±range
  cost_advanced, 150000, 600000,   # Public to private sector
  cost_ace,      3000,   12000,    # ±100%
  u_early,       0.75,   0.95,
  u_moderate,    0.60,   0.85,
  u_advanced,    0.40,   0.70,
  dr,            0.00,   0.05
)

res_dsa <- run_dsa(
  model   = res,
  dsa     = dsa_def
)

plot(res_dsa,
     result = "icer",
     type = "difference")
```

::: {.callout-tip}
## Reading the Tornado
The parameter with the longest bar is the most influential on the ICER. If `cost_advanced` (dialysis cost) dominates, that confirms our manual one-way SA finding: the intervention's value is driven primarily by how much dialysis costs it avoids.
:::

## 9. Probabilistic Sensitivity Analysis (PSA)

PSA simultaneously varies all uncertain parameters across their distributions. `define_psa()` assigns a probability distribution to each parameter, and `run_psa()` draws 1,000 samples.

```{r}
#| label: psa
#| echo: true

psa_def <- define_psa(
  # Transition probabilities: Beta distributions
  p_12_std ~ binomial(prob = 0.10, size = 100),
  p_23_std ~ binomial(prob = 0.12, size = 100),
  p_3d     ~ binomial(prob = 0.15, size = 100),
  p_1d     ~ binomial(prob = 0.02, size = 200),
  p_2d     ~ binomial(prob = 0.04, size = 200),

  # Treatment effect: Log-normal for hazard ratio
  hr ~ lognormal(meanlog = log(0.66), sdlog = 0.15),

  # Costs: Gamma distributions
  cost_early    ~ gamma(mean = 12000,  sd = 3000),
  cost_moderate ~ gamma(mean = 45000,  sd = 10000),
  cost_advanced ~ gamma(mean = 350000, sd = 70000),
  cost_ace      ~ gamma(mean = 6000,   sd = 1500),

  # Utilities: Beta-like via binomial
  u_early    ~ binomial(prob = 0.85, size = 100),
  u_moderate ~ binomial(prob = 0.72, size = 100),
  u_advanced ~ binomial(prob = 0.55, size = 100)
)

res_psa <- run_psa(
  model = res,
  psa   = psa_def,
  N     = 1000
)
```

## 10. Cost-Effectiveness Plane

The CE plane scatters all 1,000 PSA iterations on an incremental cost vs incremental QALY plot. Points in the south-east quadrant (lower cost, more QALYs) are dominant.

```{r}
#| label: fig-ce-plane
#| echo: true
#| fig-cap: "Cost-effectiveness plane from 1,000 PSA iterations"

plot(res_psa, type = "ce") +
  theme_minimal() +
  labs(title = "Cost-Effectiveness Plane",
       subtitle = "Each dot = one PSA iteration")
```

## 11. Cost-Effectiveness Acceptability Curve (CEAC)

The CEAC shows the probability that each strategy is cost-effective across a range of WTP thresholds. At India's threshold of ₹1,70,000/QALY, the intervention should have a high probability of being cost-effective.

```{r}
#| label: fig-ceac
#| echo: true
#| fig-cap: "CEAC: probability of cost-effectiveness by WTP threshold"

plot(res_psa, type = "ac",
     max_wtp = 500000,
     log_scale = FALSE) +
  geom_vline(xintercept = 170000, linetype = "dashed", colour = "red") +
  annotate("text", x = 170000, y = 0.3, label = "WTP = ₹1,70,000\n(1× GDP/capita)",
           hjust = -0.1, colour = "red", size = 3) +
  theme_minimal() +
  labs(title = "Cost-Effectiveness Acceptability Curve",
       subtitle = "Probability of being cost-effective at each WTP threshold")
```

## 12. Feature Comparison: Manual R vs heemod

| Feature | Manual R | heemod |
|---------|----------|--------|
| **Transparency** | Full — every calculation visible | Medium — internals hidden |
| **Transition matrix** | Manual `matrix()` construction | `define_transition()` with `C` complement |
| **Half-cycle correction** | Manual averaging formula | Built-in `method = "life-table"` |
| **Discounting** | Manual discount factor vector | Built-in via `run_model()` |
| **Markov trace plot** | Manual ggplot code | `plot(res, type = "counts")` |
| **One-way DSA** | Manual loop over parameter values | `define_dsa()` + `run_dsa()` + tornado |
| **PSA** | Manual sampling + simulation loop | `define_psa()` + `run_psa()` |
| **CE plane** | Manual ggplot scatter | `plot(res_psa, type = "ce")` |
| **CEAC** | Manual NMB calculation + plot | `plot(res_psa, type = "ac")` |
| **Time-varying** | Manual `if` statements in loop | Built-in `model_time`, `state_time` |
| **Learning value** | High — teaches the mechanics | Lower — abstracts the mechanics |
| **Production use** | Error-prone at scale | Robust, tested, validated |

## 13. When to Use Which

**Use the manual approach when:**

- Teaching or learning Markov model mechanics
- You need full control over every calculation
- The model is simple enough that a loop + matrix is sufficient
- You want to verify a package's output

**Use heemod when:**

- Building production HTA models for decision-makers
- You need built-in PSA, DSA, and CEAC
- The model has time-varying transitions or costs
- You want reproducible, validated output with minimal code
- You need to present professional sensitivity analysis figures

**Best practice:** Build both, cross-validate, then use heemod for the final analysis.

## Key References

- Filipović-Pierucci A, Zarca K, Durand-Zaleski I. Markov Models for Health Economic Evaluations: The R Package heemod. *arXiv:1702.03252*.
- heemod CRAN documentation: [https://cran.r-project.org/package=heemod](https://cran.r-project.org/package=heemod)
- Go AS et al. (2004). Chronic kidney disease and the risks of death and hospitalisation. *NEJM*.
- Hou FF et al. (2006). ACE inhibitor in progressive renal disease. *NEJM*.

→ **Back to:** [Session 5: CKD Markov Model (Manual)](ckd-markov-model.qmd) | [Exercise](ckd-exercise.qmd)
