---
title: "Exercise: CKD Markov Model"
subtitle: "Modify parameters and interpret Markov model results"
format:
  html:
    toc: true
    code-fold: false
---

```{r setup, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggplot2)

# Reusable function to run the full CKD Markov model
run_markov_full <- function(
    # Transition probabilities
    p_ec_arm1, p_cm_arm1, p_ad_arm1, p_d_arm1,
    p_ec_arm2, p_cm_arm2, p_ad_arm2, p_d_arm2,
    # Costs per state (annual)
    cost_early = 12000,
    cost_moderate = 45000,
    cost_advanced = 350000,
    cost_dialysis = 350000,
    # Utilities per state
    util_early = 0.95,
    util_moderate = 0.80,
    util_advanced = 0.50,
    util_dialysis = 0.40,
    # Model parameters
    n_cohort = 10000,
    n_cycles = 20,
    discount_rate = 0.03,
    cost_screening = 500,
    cost_ace_annual = 6000,
    treatment_effect_hr = 0.66) {

  # Define discount factors
  discount_factors <- 1 / (1 + discount_rate) ^ (0:n_cycles)

  # ==== ARM 1: Standard care (no ACE-inhibitor) ====

  # Transition matrix for arm 1
  m_P_arm1 <- matrix(
    c(1 - p_ec_arm1,  p_ec_arm1,        0,              0,
      0,               1 - p_cm_arm1,   p_cm_arm1,      0,
      0,               0,                1 - p_ad_arm1,  p_ad_arm1,
      0,               0,                0,              1),
    nrow = 4, ncol = 4, byrow = TRUE,
    dimnames = list(
      c("Early", "Moderate", "Advanced", "Dialysis"),
      c("Early", "Moderate", "Advanced", "Dialysis")
    )
  )

  # Initialize cohort
  v_s_arm1 <- c(1, 0, 0, 0)
  m_trace_arm1 <- matrix(NA, nrow = n_cycles + 1, ncol = 4)
  m_trace_arm1[1,] <- v_s_arm1

  # Run Markov simulation
  for (t in 2:(n_cycles + 1)) {
    m_trace_arm1[t,] <- m_trace_arm1[t-1,] %*% m_P_arm1
  }

  # Costs per cycle (arm 1: no ACE-inhibitor)
  m_cost_arm1 <- matrix(NA, nrow = n_cycles + 1, ncol = 4)
  m_cost_arm1[,1] <- cost_screening + cost_early
  m_cost_arm1[,2] <- cost_screening + cost_moderate
  m_cost_arm1[,3] <- cost_screening + cost_advanced
  m_cost_arm1[,4] <- cost_screening + cost_dialysis

  # QALYs per cycle (arm 1)
  m_qaly_arm1 <- matrix(NA, nrow = n_cycles + 1, ncol = 4)
  m_qaly_arm1[,1] <- util_early
  m_qaly_arm1[,2] <- util_moderate
  m_qaly_arm1[,3] <- util_advanced
  m_qaly_arm1[,4] <- util_dialysis

  # Calculate total costs and QALYs with half-cycle correction
  v_costs_arm1 <- colSums(m_trace_arm1[1:n_cycles,] * m_cost_arm1[1:n_cycles,] * discount_factors[1:n_cycles]) +
                  0.5 * colSums(m_trace_arm1[n_cycles + 1,] * m_cost_arm1[n_cycles + 1,] * discount_factors[n_cycles + 1])

  v_qaly_arm1 <- colSums(m_trace_arm1[1:n_cycles,] * m_qaly_arm1[1:n_cycles,] * discount_factors[1:n_cycles]) +
                 0.5 * colSums(m_trace_arm1[n_cycles + 1,] * m_qaly_arm1[n_cycles + 1,] * discount_factors[n_cycles + 1])

  total_cost_arm1 <- sum(v_costs_arm1) * n_cohort
  total_qaly_arm1 <- sum(v_qaly_arm1) * n_cohort

  # ==== ARM 2: ACE-inhibitor treatment ====

  # Transition matrix for arm 2
  m_P_arm2 <- matrix(
    c(1 - p_ec_arm2,  p_ec_arm2,        0,              0,
      0,               1 - p_cm_arm2,   p_cm_arm2,      0,
      0,               0,                1 - p_ad_arm2,  p_ad_arm2,
      0,               0,                0,              1),
    nrow = 4, ncol = 4, byrow = TRUE,
    dimnames = list(
      c("Early", "Moderate", "Advanced", "Dialysis"),
      c("Early", "Moderate", "Advanced", "Dialysis")
    )
  )

  # Initialize cohort
  v_s_arm2 <- c(1, 0, 0, 0)
  m_trace_arm2 <- matrix(NA, nrow = n_cycles + 1, ncol = 4)
  m_trace_arm2[1,] <- v_s_arm2

  # Run Markov simulation
  for (t in 2:(n_cycles + 1)) {
    m_trace_arm2[t,] <- m_trace_arm2[t-1,] %*% m_P_arm2
  }

  # Costs per cycle (arm 2: with ACE-inhibitor)
  m_cost_arm2 <- matrix(NA, nrow = n_cycles + 1, ncol = 4)
  m_cost_arm2[,1] <- cost_screening + cost_early + cost_ace_annual
  m_cost_arm2[,2] <- cost_screening + cost_moderate + cost_ace_annual
  m_cost_arm2[,3] <- cost_screening + cost_advanced + cost_ace_annual
  m_cost_arm2[,4] <- cost_screening + cost_dialysis + cost_ace_annual

  # QALYs per cycle (arm 2)
  m_qaly_arm2 <- matrix(NA, nrow = n_cycles + 1, ncol = 4)
  m_qaly_arm2[,1] <- util_early
  m_qaly_arm2[,2] <- util_moderate
  m_qaly_arm2[,3] <- util_advanced
  m_qaly_arm2[,4] <- util_dialysis

  # Calculate total costs and QALYs with half-cycle correction
  v_costs_arm2 <- colSums(m_trace_arm2[1:n_cycles,] * m_cost_arm2[1:n_cycles,] * discount_factors[1:n_cycles]) +
                  0.5 * colSums(m_trace_arm2[n_cycles + 1,] * m_cost_arm2[n_cycles + 1,] * discount_factors[n_cycles + 1])

  v_qaly_arm2 <- colSums(m_trace_arm2[1:n_cycles,] * m_qaly_arm2[1:n_cycles,] * discount_factors[1:n_cycles]) +
                 0.5 * colSums(m_trace_arm2[n_cycles + 1,] * m_qaly_arm2[n_cycles + 1,] * discount_factors[n_cycles + 1])

  total_cost_arm2 <- sum(v_costs_arm2) * n_cohort
  total_qaly_arm2 <- sum(v_qaly_arm2) * n_cohort

  # ==== Calculate ICER ====

  delta_cost <- total_cost_arm2 - total_cost_arm1
  delta_qaly <- total_qaly_arm2 - total_qaly_arm1

  icer <- delta_cost / delta_qaly

  # Return results
  list(
    costs = data.frame(
      arm1 = total_cost_arm1,
      arm2 = total_cost_arm2,
      delta = delta_cost
    ),
    qalys = data.frame(
      arm1 = total_qaly_arm1,
      arm2 = total_qaly_arm2,
      delta = delta_qaly
    ),
    icer = icer,
    trace_arm1 = as.data.frame(m_trace_arm1),
    trace_arm2 = as.data.frame(m_trace_arm2)
  )
}
```

## Task 1: Vary Time Horizon

**Objective:** Understand how the length of the analysis period affects the ICER.

The standard CKD model uses a 20-year time horizon. Change the time horizon to 10 years and run the model.

```{r task1, eval=FALSE}
# Base case parameters: transition probabilities
p_ec <- 0.08        # Early to Moderate
p_cm <- 0.12        # Moderate to Advanced
p_ad <- 0.15        # Advanced to Dialysis

# Apply treatment effect
p_ec_arm2 <- p_ec * 0.66
p_cm_arm2 <- p_cm * 0.66
p_ad_arm2 <- p_ad * 0.66

# Run with 10-year horizon
result_10y <- run_markov_full(
  p_ec_arm1 = p_ec,
  p_cm_arm1 = p_cm,
  p_ad_arm1 = p_ad,
  p_d_arm1 = 0.02,
  p_ec_arm2 = p_ec_arm2,
  p_cm_arm2 = p_cm_arm2,
  p_ad_arm2 = p_ad_arm2,
  p_d_arm2 = 0.02,
  n_cycles = ___
)

# Display results
cat("10-year time horizon:\n")
cat("Cost (Arm 1):", format(round(result_10y$costs$arm1), big.mark=","), "₹\n")
cat("Cost (Arm 2):", format(round(result_10y$costs$arm2), big.mark=","), "₹\n")
cat("QALYs (Arm 1):", round(result_10y$qalys$arm1, 2), "\n")
cat("QALYs (Arm 2):", round(result_10y$qalys$arm2, 2), "\n")
cat("ICER:", format(round(result_10y$icer), big.mark=","), "₹/QALY\n")
```

**Interpretation:**

*[Write here: How does the ICER compare to a 20-year horizon? Why do you think the ICER changes in this direction?]*

*[Write here: What does this suggest about the importance of long-term modeling in CKD interventions?]*

---

## Task 2: Vary Treatment Effectiveness

**Objective:** Assess how the magnitude of the treatment effect influences cost-effectiveness.

The ACE-inhibitor has a hazard ratio (HR) of 0.66, meaning it reduces disease progression by 34%. What if the drug was less effective with HR = 0.80 (20% reduction)?

```{r task2, eval=FALSE}
# Base case parameters
p_ec <- 0.08
p_cm <- 0.12
p_ad <- 0.15

# Weaker treatment effect (HR = 0.80)
p_ec_arm2_weak <- p_ec * ___
p_cm_arm2_weak <- p_cm * ___
p_ad_arm2_weak <- p_ad * ___

result_weak <- run_markov_full(
  p_ec_arm1 = p_ec,
  p_cm_arm1 = p_cm,
  p_ad_arm1 = p_ad,
  p_d_arm1 = 0.02,
  p_ec_arm2 = p_ec_arm2_weak,
  p_cm_arm2 = p_cm_arm2_weak,
  p_ad_arm2 = p_ad_arm2_weak,
  p_d_arm2 = 0.02,
  n_cycles = 20,
  treatment_effect_hr = ___
)

# Display results
cat("Weaker treatment effect (HR = 0.80):\n")
cat("Cost (Arm 1):", format(round(result_weak$costs$arm1), big.mark=","), "₹\n")
cat("Cost (Arm 2):", format(round(result_weak$costs$arm2), big.mark=","), "₹\n")
cat("QALYs (Arm 1):", round(result_weak$qalys$arm1, 2), "\n")
cat("QALYs (Arm 2):", round(result_weak$qalys$arm2, 2), "\n")
cat("ICER:", format(round(result_weak$icer), big.mark=","), "₹/QALY\n")
```

**Interpretation:**

*[Write here: How does a weaker treatment effect (HR = 0.80 vs 0.66) affect the ICER? Is the intervention still likely cost-effective?]*

*[Write here: What does this tell you about the value of conducting clinical trials to establish treatment efficacy?]*

---

## Task 3: Vary Dialysis Cost

**Objective:** Explore the impact of dialysis cost on cost-effectiveness, considering India's public vs. private sector variation.

In India, dialysis costs vary dramatically:
- Public sector (subsidised): ₹1,50,000/year
- Private sector: ₹5,00,000/year

Explore how the ICER changes across this range.

```{r task3, eval=FALSE}
# Base case parameters
p_ec <- 0.08
p_cm <- 0.12
p_ad <- 0.15
p_ec_arm2 <- p_ec * 0.66
p_cm_arm2 <- p_cm * 0.66
p_ad_arm2 <- p_ad * 0.66

# Test different dialysis costs
dialysis_costs <- c(150000, 250000, 350000, 500000)
results_dialysis <- list()

for (cost in dialysis_costs) {
  results_dialysis[[as.character(cost)]] <- run_markov_full(
    p_ec_arm1 = p_ec,
    p_cm_arm1 = p_cm,
    p_ad_arm1 = p_ad,
    p_d_arm1 = 0.02,
    p_ec_arm2 = p_ec_arm2,
    p_cm_arm2 = p_cm_arm2,
    p_ad_arm2 = p_ad_arm2,
    p_d_arm2 = 0.02,
    cost_dialysis = ___,
    n_cycles = 20
  )
}

# Summarize results
df_dialysis <- data.frame(
  dialysis_cost = dialysis_costs,
  icer = sapply(results_dialysis, function(x) x$icer)
)

print(df_dialysis)
```

**Interpretation:**

*[Write here: How sensitive is the ICER to dialysis cost? What is the threshold dialysis cost at which the intervention becomes dominant (cost-saving)?]*

*[Write here: What are the policy implications for recommending ACE-inhibitors in public vs. private sector settings in India?]*

---

## Task 4: Two-Way Sensitivity Analysis

**Objective:** Simultaneously vary two key parameters to understand their combined impact.

Below is complete code for a two-way sensitivity analysis varying treatment effect (HR: 0.5 to 0.9) and dialysis cost (₹1,50,000 to ₹5,00,000). Run this code and interpret the results.

```{r task4}
# Base case parameters
p_ec <- 0.08
p_cm <- 0.12
p_ad <- 0.15

# Parameter ranges for sensitivity analysis
treatment_hrs <- seq(0.5, 0.9, by = 0.05)
dialysis_costs <- seq(150000, 500000, by = 50000)

# Create grid
df_sa <- expand.grid(
  treatment_hr = treatment_hrs,
  dialysis_cost = dialysis_costs
)

# Run model for each combination
df_sa$icer <- NA

for (i in 1:nrow(df_sa)) {
  hr <- df_sa$treatment_hr[i]
  cost_dial <- df_sa$dialysis_cost[i]

  p_ec_arm2 <- p_ec * hr
  p_cm_arm2 <- p_cm * hr
  p_ad_arm2 <- p_ad * hr

  result <- run_markov_full(
    p_ec_arm1 = p_ec,
    p_cm_arm1 = p_cm,
    p_ad_arm1 = p_ad,
    p_d_arm1 = 0.02,
    p_ec_arm2 = p_ec_arm2,
    p_cm_arm2 = p_cm_arm2,
    p_ad_arm2 = p_ad_arm2,
    p_d_arm2 = 0.02,
    cost_dialysis = cost_dial,
    n_cycles = 20,
    treatment_effect_hr = hr
  )

  df_sa$icer[i] <- result$icer
}

# Create heatmap
p_heatmap <- ggplot(df_sa, aes(x = treatment_hr, y = dialysis_cost, fill = icer)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "darkgreen",
    mid = "white",
    high = "darkred",
    midpoint = 100000,
    name = "ICER (₹/QALY)"
  ) +
  scale_y_continuous(
    labels = function(x) format(x / 1000, big.mark=","),
    name = "Dialysis Cost (₹000s)"
  ) +
  scale_x_continuous(
    name = "Treatment Effect (Hazard Ratio)"
  ) +
  theme_minimal() +
  theme(
    axis.text = element_text(size = 10),
    plot.title = element_text(size = 12, face = "bold")
  ) +
  ggtitle("Sensitivity Analysis: ICER across Treatment Effect and Dialysis Cost")

print(p_heatmap)
```

**Interpretation:**

*[Write here: Which parameter (treatment effect or dialysis cost) appears to have a larger impact on the ICER?]*

*[Write here: In what region of the parameter space does the intervention become dominant (cost-saving)?]*

*[Write here: What does this analysis suggest about where the ACE-inhibitor program should be prioritized within India's health system?]*

---

## Next Steps

→ See the **[solution file](ckd-solution.qmd)** for completed answers and detailed interpretations.
