---
title: "4-State Markov Model — Chronic Kidney Disease"
subtitle: "Early screening and ACE-inhibitor intervention vs standard care"
author: "R for HTA (Basics) — Workshop 2026"
institute: "RRC-HTA, AIIMS Bhopal | HTAIn, DHR"
format:
  revealjs:
    theme: [default, ../../custom-theme.scss]
    slide-number: true
    footer: "R for HTA (Basics) — RRC-HTA, AIIMS Bhopal"
    logo: ""
    transition: fade
    width: 1280
    height: 720
    code-fold: true
    code-overflow: wrap
    highlight-style: github
    fig-align: center
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

## The Clinical Question

CKD affects ~**17% of India's population**, with ~6% at stage 3 or worse.

. . .

**The problem:** Over 50% of Indian patients present late — when eGFR < 15 ml/min (stage 5), requiring expensive dialysis or transplantation.

. . .

**The HTA question:**

Is **early population-based screening** followed by **ACE-inhibitor therapy** cost-effective compared to standard care (where CKD is detected only when symptomatic)?

. . .

*This is a **Markov cohort model** — the most common type for chronic diseases that progress through stages.*

## What Is a Markov Model?

A Markov model tracks a **cohort** of patients through defined **health states** over multiple **time cycles** (usually years).

. . .

At each cycle, patients can:

- Stay in their current state
- Move to a worse state (progression)
- Move to a better state (rare in CKD)
- **Die**

. . .

The movement between states is governed by **transition probabilities** — the heart of any Markov model.

## Our 4-State Model Structure

```{r}
#| label: fig-ckd-states
#| echo: false
#| eval: true
#| fig-width: 12
#| fig-height: 5

library(DiagrammeR)

grViz("
digraph ckd_markov {
  graph [rankdir=LR, bgcolor='transparent', fontname='Helvetica', nodesep=0.8]
  node [fontname='Helvetica', fontsize=11, style='filled,rounded', shape=box]

  E [label='Early CKD\n(Stage 1-2)\neGFR ≥ 60', fillcolor='#59a14f', fontcolor='white', width=1.8]
  M [label='Moderate CKD\n(Stage 3)\neGFR 30-59', fillcolor='#f28e2b', fontcolor='white', width=1.8]
  A [label='Advanced CKD\n/ Dialysis\n(Stage 4-5)', fillcolor='#e15759', fontcolor='white', width=1.8]
  D [label='Death', fillcolor='#76b7b2', fontcolor='white', shape=doublecircle, width=1.2]

  E -> M [label='p_12', fontsize=10]
  M -> A [label='p_23', fontsize=10]
  E -> D [label='p_1d', fontsize=10]
  M -> D [label='p_2d', fontsize=10]
  A -> D [label='p_3d', fontsize=10]

  E -> E [label='stay', style=dashed, fontsize=9]
  M -> M [label='stay', style=dashed, fontsize=9]
  A -> A [label='stay', style=dashed, fontsize=9]
}
")
```

## Health States Explained

::::: {.columns}

:::: {.column width="50%"}

**Early CKD** (Stage 1-2)
- eGFR ≥ 60
- Often asymptomatic
- Managed with lifestyle + renoprotective drugs

**Moderate CKD** (Stage 3)
- eGFR 30-59
- Requires specialist referral
- Medication optimization

::::

:::: {.column width="50%"}

**Advanced CKD / Dialysis** (Stage 4-5)
- eGFR < 30
- Requires dialysis or transplantation
- Most expensive state: **₹3.5 lakhs/year**

**Death**
- Absorbing state
- Patients cannot leave

::::

:::::

## Key Model Parameters

```{r}
#| label: parameters-display
#| echo: false

cat("TRANSITION PROBABILITIES (Annual):\n")
cat("Early → Moderate:     10% (standard), 7% (with ACE-inhibitor)\n")
cat("Moderate → Advanced:  12% (standard), 8% (with ACE-inhibitor)\n")
cat("Early → Death:        2% (both arms)\n")
cat("Moderate → Death:     4% (both arms)\n")
cat("Advanced → Death:     15% (both arms)\n\n")

cat("COSTS (₹ per year):\n")
cat("Early CKD:            ₹12,000\n")
cat("Moderate CKD:         ₹45,000\n")
cat("Advanced/Dialysis:    ₹3,50,000\n")
cat("ACE-inhibitor:        ₹6,000 additional\n")
cat("Screening (one-time): ₹500\n\n")

cat("UTILITIES (Quality-adjusted life expectancy):\n")
cat("Early:                0.85\n")
cat("Moderate:             0.72\n")
cat("Advanced:             0.55\n")
```

## Building the Transition Matrix

A **transition matrix** is a table where:
- Each **row** = current state
- Each **column** = destination state
- All rows sum to **1.0** (everyone goes somewhere)

. . .

```r
# Standard Care Transition Matrix (example)
#                Early  Moderate  Advanced  Death
# Early         0.880   0.100     0.000    0.020
# Moderate      0.000   0.840     0.120    0.040
# Advanced      0.000   0.000     0.850    0.150
# Death         0.000   0.000     0.000    1.000
```

. . .

**Key insight:** The intervention matrix differs only in progression probabilities. Treatment **slows progression** but doesn't change mortality directly.

## How the Simulation Works

At each **cycle**, patient distribution is updated:

```r
trace[cycle + 1, ] <- trace[cycle, ] %*% tm
```

. . .

This single line of **matrix multiplication** handles all transitions simultaneously:
- How many move from Early → Moderate?
- How many stay in Moderate?
- How many die?

. . .

All answered by multiplying the current state vector by the transition matrix.

## Half-Cycle Correction

For accuracy, we assume patients spend **half the cycle** in their original state and **half in the new state**.

```r
trace_hcc <- (trace[1:n_cycles, ] + trace[2:(n_cycles + 1), ]) / 2
```

. . .

This adjustment leads to more realistic cost and QALY calculations, especially important for short time horizons.

## Results: Costs and QALYs per Patient

| Strategy | Cost/Patient (₹) | QALYs/Patient |
|----------|------------------|---------------|
| Standard Care | ₹8,06,572 | 8.47 |
| Screening + ACE-inhibitor | ₹6,47,370 | 9.24 |
| **Incremental** | **₹−1,59,202** | **0.76** |

. . .

**Key finding:** The intervention is **cheaper AND more effective** — it saves ₹1.6 lakhs per patient by keeping them out of expensive dialysis, while gaining 0.76 QALYs.

## ICER — But Check the Signs First!

```r
inc_cost <- total_cost_int - total_cost_std   # Negative (saves money)
inc_qaly <- total_qaly_int - total_qaly_std   # Positive (gains health)
icer <- inc_cost / inc_qaly                   # Negative ICER
```

. . .

A negative ICER can mean **DOMINANT** or **DOMINATED** — always check:

| | ΔQALYs > 0 | ΔQALYs < 0 |
|---|---|---|
| **ΔCost < 0** | **DOMINANT** ✓ | Trade-off |
| **ΔCost > 0** | Compare to WTP | **DOMINATED** |

. . .

Here: ΔCost < 0 AND ΔQALYs > 0 → **DOMINANT** (ICER = ₹−2,08,132)

## Net Monetary Benefit (NMB)

NMB removes the ICER ambiguity. **Positive ΔNMB = adopt.**

$$\text{NMB} = \text{WTP} \times \text{QALYs} - \text{Cost}$$

. . .

| Strategy | NMB/Patient |
|----------|-------------|
| Standard Care | ₹6,33,454 |
| Intervention | ₹9,22,691 |
| **ΔNMB** | **₹2,89,237** → ADOPT |

. . .

Use NMB (not ICER) in PSA — you can't average ICERs but you can average NMBs.

## Discounting — Why and How

**Question:** Why do we discount future costs and benefits?

. . .

**Answer:** Society prefers benefits today over benefits tomorrow (time preference). A QALY today is worth more than a QALY in year 10.

. . .

**Standard rate:** **3% annual discounting** in India

. . .

```r
discount_factors <- 1 / (1 + 0.03)^(year - 1)
# Year 1: factor = 1.000
# Year 5: factor = 0.863
# Year 10: factor = 0.744
```

. . .

Both **costs AND utilities** must be discounted.

## Key Insight: The Markov Trace

The **Markov trace** visualizes the cohort's journey:

. . .

Early CKD (green) shrinks over time as patients progress → Moderate (orange) → Advanced (red) → Death (blue).

. . .

**With treatment:** Patients stay in green longer = **less time in expensive dialysis** = lower cost despite higher upfront drug spending.

. . .

This is the visual signature of a drug that works: it **buys time in better health states**.

## What You Learned About Markov Models

1. **Health states** — distinct conditions patients inhabit
2. **Transition matrix** — rows sum to 1, `C` shorthand for stay probability
3. **Matrix multiplication** — `trace[t+1, ] <- trace[t, ] %*% tm`
4. **Markov trace** — cohort distribution over time
5. **Half-cycle correction** — average start/end of each cycle
6. **Discounting** — 3% per year on costs AND QALYs
7. **ICER** — with proper 4-quadrant interpretation
8. **NMB** — unambiguous decision metric

. . .

**In Excel:** Complex array formulas. **In R:** One line of matrix multiplication.

## Bonus Materials

- **heemod package** — same model with `define_parameters`, `define_state`, `define_transition`, built-in DSA + PSA
- **Excel companion** — 5-sheet workbook for cross-validation
- **Pen-and-paper worksheet** — handshake tables for matrix multiplication

. . .

→ See the **Session 5 Bonus** and **Downloads** pages on the workshop website.

---

**Next:** Session 6 — Probabilistic Sensitivity Analysis (uncertainty in all parameters)
