---
title: "Session 6: Probabilistic Sensitivity Analysis"
subtitle: "Why PSA matters and how R makes it practical"
author: "HTA in R Workshop"
date: "`r Sys.Date()`"
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    code-line-numbers: true
    theme: cosmo
    highlight-style: github
---

## Why Probabilistic Sensitivity Analysis?

In health technology assessment, we rarely know parameter values with certainty. When you submitted a cost-effectiveness model to a journal or HTA agency using only point estimates—a single transition probability, a single cost, a single utility value—you were essentially claiming that these are the true values. But they are not. Every parameter carries uncertainty from the sources we used: clinical trials with limited sample sizes, cost data from particular regions, utility estimates that varied across populations.

Point estimates tell us what would happen if all parameters took their best-guess values simultaneously. But what if the true transition probability is slightly higher? What if costs are trending upward? What if the target population has lower health-related quality of life than the trial sample? Journals and HTA agencies—NICE, CADTH, India's NITI Aayog, and others—now *require* that you demonstrate your results are robust to this uncertainty. They want to know: does your recommendation change if we vary the parameters within their plausible ranges? How confident can we really be?

This is where **Probabilistic Sensitivity Analysis (PSA)** steps in, and where R becomes genuinely powerful compared to Excel. In Excel, running 5,000 scenarios is tedious and error-prone: you either manage massive tables of random numbers or spend hours with Data Tables. In R, it is a few lines of code. You define your parameter distributions once, then sample thousands of times in a loop, recording the results. You get not just point estimates but probability distributions of your outcomes—and more importantly, you answer the real question: "Given all sources of parameter uncertainty, what is the probability that each strategy is cost-effective?"

```{r}
#| label: fig-psa-workflow
#| echo: true
#| fig-cap: "The PSA workflow — same steps regardless of model type"

library(DiagrammeR)

grViz("
digraph psa_workflow {
  graph [rankdir=LR, bgcolor='transparent', fontname='Helvetica']
  node [shape=box, style='filled,rounded', fontname='Helvetica', fontsize=11]

  A [label='1. Define\nDistributions\nfor each parameter', fillcolor='#4e79a7', fontcolor='white']
  B [label='2. Sample\nParameters\n(random draw)', fillcolor='#59a14f', fontcolor='white']
  C [label='3. Run\nModel\n(one iteration)', fillcolor='#f28e2b', fontcolor='white']
  D [label='4. Store\nResults\n(cost, QALY, ICER)', fillcolor='#e15759', fontcolor='white']
  E [label='5. Repeat\n1000-10000\ntimes', fillcolor='#76b7b2', fontcolor='white']
  F [label='6. Summarise\nUncertainty\n(CE plane, CEAC)', fillcolor='#edc948', fontcolor='black']

  A -> B -> C -> D -> E -> F
  E -> B [style=dashed, label='Loop', fontsize=9]
}
")
```

## Parameter Distributions in HTA

Different parameter types follow different distributions. Here is how to translate your uncertainties into statistical distributions that R can sample from.

```{r}
#| label: fig-distribution-map
#| echo: true
#| fig-cap: "Choosing the right distribution for each parameter type"

grViz("
digraph dist_map {
  graph [rankdir=TB, bgcolor='transparent', fontname='Helvetica']
  node [fontname='Helvetica', fontsize=11, style='filled,rounded', shape=box]

  title [label='Parameter Type → Distribution', shape=none, fontsize=14, fontname='Helvetica-Bold']

  prob [label='Probabilities\n& Utilities\n(bounded 0-1)', fillcolor='#4e79a7', fontcolor='white', width=2]
  cost [label='Costs\n(non-negative,\nright-skewed)', fillcolor='#59a14f', fontcolor='white', width=2]
  hr [label='Hazard Ratios\n& Relative Risks\n(positive, often <1)', fillcolor='#f28e2b', fontcolor='white', width=2]

  beta [label='Beta(α, β)', fillcolor='#d4e6f1', shape=ellipse, width=1.5]
  gamma [label='Gamma(shape, rate)', fillcolor='#d5f5e3', shape=ellipse, width=1.5]
  lognorm [label='Log-Normal(μ, σ)', fillcolor='#fdebd0', shape=ellipse, width=1.5]

  title -> prob [style=invis]
  title -> cost [style=invis]
  title -> hr [style=invis]

  prob -> beta
  cost -> gamma
  hr -> lognorm
}
")
```

### Beta Distribution for Probabilities

Probabilities and proportions (transition probabilities, event rates) follow a **Beta distribution**. The Beta is bounded between 0 and 1, making it ideal.

**Parameterisation from mean and SE:**

If you have a transition probability with mean = 0.10 and standard error = 0.02, you can convert these to Beta parameters (α and β) using:

- α = μ × (μ(1-μ)/σ² - 1)
- β = (1-μ) × (μ(1-μ)/σ² - 1)

where μ is the mean and σ² is the variance (SE²).

**R code example:**

```{r}
#| label: beta-example
#| eval: false

# Example: transition probability p_12 = 0.10, SE = 0.02
mean_p <- 0.10
se_p <- 0.02
var_p <- se_p^2

alpha <- mean_p * (mean_p * (1 - mean_p) / var_p - 1)
beta <- (1 - mean_p) * (mean_p * (1 - mean_p) / var_p - 1)

# Sample 10,000 times
samples_beta <- rbeta(10000, alpha, beta)
```

### Gamma Distribution for Costs

Costs are non-negative and typically right-skewed (a few very expensive patients pull the mean upward). The **Gamma distribution** fits this well.

**Parameterisation from mean and SE:**

If costs have mean = ₹45,000 and SE = ₹5,000:

- shape = μ² / σ²
- rate = μ / σ²

**R code example:**

```{r}
#| label: gamma-example
#| eval: false

# Example: cost_moderate = ₹45,000, SE = ₹5,000
mean_cost <- 45000
se_cost <- 5000

shape <- mean_cost^2 / (se_cost^2)
rate <- mean_cost / (se_cost^2)

# Sample 10,000 times
samples_gamma <- rgamma(10000, shape = shape, rate = rate)
```

### Log-Normal Distribution for Hazard Ratios

Hazard ratios, relative risks, and odds ratios are bounded at 0 and can be any positive value. The **Log-Normal distribution** is natural: the log of a log-normal is normal, and we typically report confidence intervals on the log scale.

**Parameterisation from mean and 95% CI:**

If a hazard ratio has point estimate = 0.66 with 95% CI (0.55, 0.79):

1. The log(HR) follows Normal: log(0.66) ≈ -0.416
2. The 95% CI on the log scale is [log(0.55), log(0.79)] ≈ [-0.598, -0.236]
3. The SE on log scale ≈ (log upper - log lower) / 3.92

**R code example:**

```{r}
#| label: lognormal-example
#| eval: false

# Example: HR = 0.66, 95% CI (0.55, 0.79)
hr_point <- 0.66
hr_lower <- 0.55
hr_upper <- 0.79

log_hr_se <- (log(hr_upper) - log(hr_lower)) / 3.92
log_hr_mean <- log(hr_point)

# Sample 10,000 times
samples_lognormal <- rlnorm(10000,
                            meanlog = log_hr_mean,
                            sdlog = log_hr_se)
```

### Visualising Example Distributions

Let us see what these distributions look like side-by-side:

```{r}
#| label: distributions-plot
#| message: false
#| fig-width: 12
#| fig-height: 5

library(ggplot2)
library(dplyr)
library(tidyr)

# Set seed for reproducibility
set.seed(2026)

# Beta example: transition probability p_12 = 0.10, SE = 0.02
mean_p <- 0.10
se_p <- 0.02
var_p <- se_p^2
alpha_beta <- mean_p * (mean_p * (1 - mean_p) / var_p - 1)
beta_beta <- (1 - mean_p) * (mean_p * (1 - mean_p) / var_p - 1)
samples_beta <- rbeta(5000, alpha_beta, beta_beta)

# Gamma example: cost_moderate = ₹45,000, SE = ₹5,000
mean_cost <- 45000
se_cost <- 5000
shape_gamma <- mean_cost^2 / (se_cost^2)
rate_gamma <- mean_cost / (se_cost^2)
samples_gamma <- rgamma(5000, shape = shape_gamma, rate = rate_gamma)

# Log-Normal example: HR = 0.66, 95% CI (0.55, 0.79)
hr_point <- 0.66
hr_lower <- 0.55
hr_upper <- 0.79
log_hr_se <- (log(hr_upper) - log(hr_lower)) / 3.92
log_hr_mean <- log(hr_point)
samples_lognormal <- rlnorm(5000, meanlog = log_hr_mean, sdlog = log_hr_se)

# Combine into a single data frame for plotting
dist_data <- tibble(
  Distribution = rep(c("Beta\n(Probability)", "Gamma\n(Cost ₹)", "Log-Normal\n(Hazard Ratio)"),
                     c(5000, 5000, 5000)),
  Value = c(samples_beta, samples_gamma, samples_lognormal)
)

# Plot
ggplot(dist_data, aes(x = Value, fill = Distribution)) +
  geom_density(alpha = 0.6, color = NA) +
  facet_wrap(~ Distribution, scales = "free") +
  labs(
    title = "Example Parameter Distributions for PSA",
    subtitle = "Left: Beta(α=4.28, β=38.5) | Middle: Gamma(shape=81, rate=0.0018) | Right: LogNormal(μ=-0.416, σ=0.088)",
    x = "Parameter Value",
    y = "Density"
  ) +
  theme_minimal() +
  theme(
    legend.position = "none",
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10)
  )
```

## PSA Applied to the CKD Markov Model

We'll now run a full Probabilistic Sensitivity Analysis on the CKD model from Session 5. Recall the model structure:

- **States:** Early CKD (Stage 2) → Moderate CKD (Stage 3) → Advanced CKD / Dialysis (Stage 5) → Death
- **Intervention:** ACE inhibitor treatment (reduces progression by HR = 0.66)
- **Cycle length:** 1 year
- **Time horizon:** 20 years

### Defining the PSA Function

Here is the core function that runs one iteration of the Markov model with a sampled parameter set:

```{r}
#| label: run-markov-psa-function

run_markov_psa <- function(
    # Transition probabilities
    p_12, p_23, p_1d, p_2d, p_3d,
    # Treatment effect (hazard ratio for progression)
    hr_treatment,
    # Costs (annual, in Rupees)
    cost_early, cost_moderate, cost_advanced, cost_ace, cost_screening,
    # Utilities
    util_early, util_moderate, util_advanced,
    # Time horizon
    n_years = 20) {

  # Number of cohort members
  n_cohort <- 1000

  # Initialize states for standard care (no treatment)
  states_standard <- matrix(0, nrow = n_years + 1, ncol = 4)
  colnames(states_standard) <- c("Early", "Moderate", "Advanced", "Dead")

  # Initialize states for intervention (with treatment)
  states_intervention <- matrix(0, nrow = n_years + 1, ncol = 4)
  colnames(states_intervention) <- c("Early", "Moderate", "Advanced", "Dead")

  # Starting cohort: all in Early CKD state
  states_standard[1, 1] <- n_cohort
  states_intervention[1, 1] <- n_cohort

  # Run Markov chain for 20 years
  for (year in 1:n_years) {
    # Standard care: use base transition probabilities
    early_standard <- states_standard[year, 1]
    moderate_standard <- states_standard[year, 2]
    advanced_standard <- states_standard[year, 3]

    states_standard[year + 1, 1] <- early_standard * (1 - p_12 - p_1d)
    states_standard[year + 1, 2] <- early_standard * p_12 + moderate_standard * (1 - p_23 - p_2d)
    states_standard[year + 1, 3] <- moderate_standard * p_23 + advanced_standard * (1 - p_3d)
    states_standard[year + 1, 4] <- early_standard * p_1d +
                                     moderate_standard * p_2d +
                                     advanced_standard * p_3d +
                                     states_standard[year, 4]

    # Intervention: apply HR to progression probabilities
    early_intervention <- states_intervention[year, 1]
    moderate_intervention <- states_intervention[year, 2]
    advanced_intervention <- states_intervention[year, 3]

    # Reduced progression with HR
    p_12_tx <- 1 - (1 - p_12)^hr_treatment
    p_23_tx <- 1 - (1 - p_23)^hr_treatment

    states_intervention[year + 1, 1] <- early_intervention * (1 - p_12_tx - p_1d)
    states_intervention[year + 1, 2] <- early_intervention * p_12_tx +
                                         moderate_intervention * (1 - p_23_tx - p_2d)
    states_intervention[year + 1, 3] <- moderate_intervention * p_23_tx +
                                         advanced_intervention * (1 - p_3d)
    states_intervention[year + 1, 4] <- early_intervention * p_1d +
                                         moderate_intervention * p_2d +
                                         advanced_intervention * p_3d +
                                         states_intervention[year, 4]
  }

  # Calculate total costs and QALYs for standard care
  cost_standard <- sum(
    states_standard[-1, 1] * cost_early +
    states_standard[-1, 2] * cost_moderate +
    states_standard[-1, 3] * cost_advanced +
    states_standard[-1, 1] * cost_screening
  )

  qaly_standard <- sum(
    states_standard[-1, 1] * util_early +
    states_standard[-1, 2] * util_moderate +
    states_standard[-1, 3] * util_advanced
  )

  # Calculate total costs and QALYs for intervention
  cost_intervention <- sum(
    states_intervention[-1, 1] * cost_early +
    states_intervention[-1, 2] * cost_moderate +
    states_intervention[-1, 3] * cost_advanced +
    states_intervention[-1, 1] * cost_screening +
    states_intervention[-1, 1] * cost_ace
  )

  qaly_intervention <- sum(
    states_intervention[-1, 1] * util_early +
    states_intervention[-1, 2] * util_moderate +
    states_intervention[-1, 3] * util_advanced
  )

  return(list(
    cost_standard = cost_standard,
    cost_intervention = cost_intervention,
    qaly_standard = qaly_standard,
    qaly_intervention = qaly_intervention
  ))
}
```

### Running the PSA: 5,000 Iterations

Now we define parameter distributions and run the PSA loop:

```{r}
#| label: run-ckd-psa
#| message: false

library(ggplot2)
library(dplyr)
library(tidyr)

set.seed(2026)

# Define parameter distributions
# ============================================================================

# Transition probabilities (Beta distributions)
# Base parameters from Session 5
p_12_mean <- 0.10
p_23_mean <- 0.12
p_1d_mean <- 0.02
p_2d_mean <- 0.04
p_3d_mean <- 0.15

# Standard errors (assumed from literature)
p_se <- 0.02  # Common SE for transition probabilities

# Convert to Beta parameters
convert_beta_params <- function(mean, se) {
  var <- se^2
  alpha <- mean * (mean * (1 - mean) / var - 1)
  beta <- (1 - mean) * (mean * (1 - mean) / var - 1)
  return(c(alpha = alpha, beta = beta))
}

beta_12 <- convert_beta_params(p_12_mean, p_se)
beta_23 <- convert_beta_params(p_23_mean, p_se)
beta_1d <- convert_beta_params(p_1d_mean, p_se)
beta_2d <- convert_beta_params(p_2d_mean, p_se)
beta_3d <- convert_beta_params(p_3d_mean, p_se)

# Treatment effect (Log-Normal): HR = 0.66, 95% CI (0.55, 0.79)
hr_mean <- 0.66
hr_lower <- 0.55
hr_upper <- 0.79
log_hr_se <- (log(hr_upper) - log(hr_lower)) / 3.92
log_hr_mean <- log(hr_mean)

# Costs (Gamma distributions)
# Base costs and standard errors
cost_early_mean <- 12000
cost_early_se <- 1500

cost_moderate_mean <- 45000
cost_moderate_se <- 5000

cost_advanced_mean <- 350000
cost_advanced_se <- 50000

cost_ace_mean <- 6000
cost_ace_se <- 800

cost_screening_mean <- 500
cost_screening_se <- 100

# Convert to Gamma parameters
convert_gamma_params <- function(mean, se) {
  var <- se^2
  shape <- mean^2 / var
  rate <- mean / var
  return(c(shape = shape, rate = rate))
}

gamma_early <- convert_gamma_params(cost_early_mean, cost_early_se)
gamma_moderate <- convert_gamma_params(cost_moderate_mean, cost_moderate_se)
gamma_advanced <- convert_gamma_params(cost_advanced_mean, cost_advanced_se)
gamma_ace <- convert_gamma_params(cost_ace_mean, cost_ace_se)
gamma_screening <- convert_gamma_params(cost_screening_mean, cost_screening_se)

# Utilities (Beta distributions)
# Utilities are proportions, so use Beta(α, β) directly
util_early_mean <- 0.85
util_early_se <- 0.05

util_moderate_mean <- 0.72
util_moderate_se <- 0.06

util_advanced_mean <- 0.55
util_advanced_se <- 0.08

beta_util_early <- convert_beta_params(util_early_mean, util_early_se)
beta_util_moderate <- convert_beta_params(util_moderate_mean, util_moderate_se)
beta_util_advanced <- convert_beta_params(util_advanced_mean, util_advanced_se)

# ============================================================================
# RUN PSA: 5,000 iterations
# ============================================================================

n_sim <- 5000
psa_results <- data.frame(
  iteration = 1:n_sim,
  cost_standard = NA,
  cost_intervention = NA,
  qaly_standard = NA,
  qaly_intervention = NA,
  inc_cost = NA,
  inc_qaly = NA,
  icer = NA
)

cat("Running CKD PSA...\n")

for (i in 1:n_sim) {
  # Sample parameters from distributions
  p_12_sampled <- rbeta(1, beta_12["alpha"], beta_12["beta"])
  p_23_sampled <- rbeta(1, beta_23["alpha"], beta_23["beta"])
  p_1d_sampled <- rbeta(1, beta_1d["alpha"], beta_1d["beta"])
  p_2d_sampled <- rbeta(1, beta_2d["alpha"], beta_2d["beta"])
  p_3d_sampled <- rbeta(1, beta_3d["alpha"], beta_3d["beta"])

  hr_sampled <- rlnorm(1, log_hr_mean, log_hr_se)

  cost_early_sampled <- rgamma(1, gamma_early["shape"], gamma_early["rate"])
  cost_moderate_sampled <- rgamma(1, gamma_moderate["shape"], gamma_moderate["rate"])
  cost_advanced_sampled <- rgamma(1, gamma_advanced["shape"], gamma_advanced["rate"])
  cost_ace_sampled <- rgamma(1, gamma_ace["shape"], gamma_ace["rate"])
  cost_screening_sampled <- rgamma(1, gamma_screening["shape"], gamma_screening["rate"])

  util_early_sampled <- rbeta(1, beta_util_early["alpha"], beta_util_early["beta"])
  util_moderate_sampled <- rbeta(1, beta_util_moderate["alpha"], beta_util_moderate["beta"])
  util_advanced_sampled <- rbeta(1, beta_util_advanced["alpha"], beta_util_advanced["beta"])

  # Run model with sampled parameters
  result <- run_markov_psa(
    p_12 = p_12_sampled,
    p_23 = p_23_sampled,
    p_1d = p_1d_sampled,
    p_2d = p_2d_sampled,
    p_3d = p_3d_sampled,
    hr_treatment = hr_sampled,
    cost_early = cost_early_sampled,
    cost_moderate = cost_moderate_sampled,
    cost_advanced = cost_advanced_sampled,
    cost_ace = cost_ace_sampled,
    cost_screening = cost_screening_sampled,
    util_early = util_early_sampled,
    util_moderate = util_moderate_sampled,
    util_advanced = util_advanced_sampled
  )

  # Store results
  psa_results$cost_standard[i] <- result$cost_standard
  psa_results$cost_intervention[i] <- result$cost_intervention
  psa_results$qaly_standard[i] <- result$qaly_standard
  psa_results$qaly_intervention[i] <- result$qaly_intervention
  psa_results$inc_cost[i] <- result$cost_intervention - result$cost_standard
  psa_results$inc_qaly[i] <- result$qaly_intervention - result$qaly_standard
  psa_results$icer[i] <- psa_results$inc_cost[i] / psa_results$inc_qaly[i]

  # Progress update
  if (i %% 1000 == 0) {
    cat(sprintf("Completed %d of %d iterations\n", i, n_sim))
  }
}

cat("PSA complete.\n")
```

### PSA Results Summary

```{r}
#| label: psa-summary-ckd

# Calculate summary statistics
mean_icer <- mean(psa_results$icer)
median_icer <- median(psa_results$icer)
lower_ci_icer <- quantile(psa_results$icer, 0.025)
upper_ci_icer <- quantile(psa_results$icer, 0.975)

mean_inc_cost <- mean(psa_results$inc_cost)
mean_inc_qaly <- mean(psa_results$inc_qaly)

cat("======================================\n")
cat("CKD Markov Model PSA Results (5,000 iterations)\n")
cat("======================================\n\n")

cat("Mean Incremental Cost:    ", format(round(mean_inc_cost), big.mark=","), " ₹\n")
cat("Mean Incremental QALY:    ", format(round(mean_inc_qaly, 2), big.mark=","), "\n")
cat("Mean ICER:                ", format(round(mean_icer), big.mark=","), " ₹/QALY\n")
cat("Median ICER:              ", format(round(median_icer), big.mark=","), " ₹/QALY\n")
cat("95% Credible Interval:    ₹", format(round(lower_ci_icer), big.mark=","),
    " to ₹", format(round(upper_ci_icer), big.mark=","), "\n\n")

cat("WTP Threshold:            ₹170,000 per QALY\n")
```

```{r}
#| label: fig-icer-histogram-ckd
#| echo: true
#| fig-cap: "Distribution of ICERs from 5,000 PSA iterations — CKD Markov model"

# Filter out extreme ICERs for better visualisation
icer_filtered <- psa_results$icer[psa_results$icer > quantile(psa_results$icer, 0.01) &
                                   psa_results$icer < quantile(psa_results$icer, 0.99)]

ggplot(data.frame(ICER = icer_filtered), aes(x = ICER)) +
  geom_histogram(bins = 50, fill = "steelblue", colour = "white", alpha = 0.8) +
  geom_vline(xintercept = 170000, colour = "red", linetype = "dashed", linewidth = 1) +
  annotate("text", x = 170000, y = Inf, label = "WTP = ₹170,000/QALY",
           colour = "red", hjust = -0.1, vjust = 2, size = 3.5) +
  geom_vline(xintercept = mean(psa_results$icer), colour = "darkblue", linewidth = 1) +
  annotate("text", x = mean(psa_results$icer), y = Inf, label = "Mean ICER",
           colour = "darkblue", hjust = -0.1, vjust = 4, size = 3.5) +
  scale_x_continuous(labels = function(x) paste0("₹", format(round(x/1000), big.mark=","), "K")) +
  labs(x = "ICER (₹/QALY)", y = "Frequency",
       title = "Distribution of ICER Estimates",
       subtitle = "CKD Markov Model — 5,000 PSA iterations") +
  theme_minimal()
```

## PSA Visualisations: Cost-Effectiveness Plane

The **cost-effectiveness plane** shows all 5,000 PSA iterations as points. The x-axis is incremental QALYs (our model gains QALYs with treatment), and the y-axis is incremental costs.

- Points in the **bottom-right quadrant** are dominant: more effective AND cheaper.
- Points in the **top-right quadrant** are cost-effective if they fall below the WTP line.
- Points in the **top-left quadrant** are dominated: more costly but less effective.

```{r}
#| label: ce-plane-ckd
#| fig-width: 10
#| fig-height: 7
#| message: false

# Add WTP line parameter
wtp <- 170000  # ₹170,000 per QALY

# Create the cost-effectiveness plane
ce_plane <- ggplot(psa_results, aes(x = inc_qaly, y = inc_cost)) +
  # Add WTP line
  geom_abline(intercept = 0, slope = wtp, color = "red", linewidth = 1, linetype = "dashed",
              label = paste("WTP threshold: ₹", format(wtp, big.mark=","), "/QALY")) +
  # Add data points
  geom_point(alpha = 0.4, color = "steelblue", size = 2) +
  # Add quadrant lines
  geom_hline(yintercept = 0, color = "gray70", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "gray70", linewidth = 0.5) +
  # Labels and formatting
  labs(
    title = "Cost-Effectiveness Plane: CKD Markov Model PSA",
    subtitle = paste(n_sim, "iterations"),
    x = "Incremental QALYs (Intervention vs Standard Care)",
    y = "Incremental Cost (₹)",
    caption = "Red dashed line = ₹170,000/QALY willingness-to-pay threshold"
  ) +
  scale_y_continuous(labels = scales::comma_format(prefix = "₹")) +
  scale_x_continuous(labels = scales::comma_format()) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11),
    panel.grid.major = element_line(color = "gray95"),
    legend.position = "top"
  )

print(ce_plane)
```

## Cost-Effectiveness Acceptability Curve (CEAC)

The **CEAC** answers the question: "Across a range of willingness-to-pay thresholds, what is the probability that the intervention is cost-effective?" It is calculated by determining, for each WTP threshold, what proportion of the 5,000 PSA iterations had an ICER below that threshold.

```{r}
#| label: ceac-ckd
#| fig-width: 10
#| fig-height: 6
#| message: false

# Define a range of WTP thresholds
wtp_range <- seq(0, 500000, by = 10000)

# For each WTP, calculate probability that intervention is cost-effective
ceac_data <- data.frame(
  wtp = wtp_range,
  prob_cost_effective = NA
)

for (j in 1:nrow(ceac_data)) {
  # Net Monetary Benefit (NMB) for intervention at each WTP:
  # NMB = (inc_qaly * wtp) - inc_cost
  nmb_intervention <- psa_results$inc_qaly * ceac_data$wtp[j] - psa_results$inc_cost

  # Probability that NMB > 0 (i.e., intervention is cost-effective)
  ceac_data$prob_cost_effective[j] <- mean(nmb_intervention > 0)
}

# Plot CEAC
ceac_plot <- ggplot(ceac_data, aes(x = wtp, y = prob_cost_effective)) +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_area(fill = "steelblue", alpha = 0.3) +
  # Add reference line at WTP = 170,000
  geom_vline(xintercept = 170000, color = "red", linetype = "dashed", linewidth = 1) +
  # Labels and formatting
  labs(
    title = "Cost-Effectiveness Acceptability Curve (CEAC): CKD Markov Model",
    subtitle = paste(n_sim, "PSA iterations"),
    x = "Willingness-to-Pay Threshold (₹ per QALY)",
    y = "Probability That Intervention is Cost-Effective",
    caption = "Red dashed line = ₹170,000/QALY threshold"
  ) +
  scale_x_continuous(labels = scales::comma_format(prefix = "₹")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 11),
    panel.grid.major = element_line(color = "gray95")
  )

print(ceac_plot)

# Calculate probability at the ₹170,000 threshold
prob_ce_at_170k <- mean((psa_results$inc_qaly * 170000 - psa_results$inc_cost) > 0)
cat("\n\nAt WTP = ₹170,000/QALY:\n")
cat(sprintf("Probability that intervention is cost-effective: %.1f%%\n", prob_ce_at_170k * 100))
```

## PSA Applied to DES vs BMS Decision Tree

We can apply the same PSA logic to the DES vs BMS comparison from Session 4. The model is simpler (a decision tree with no transitions over time), but the principle is identical: define parameter distributions, sample 5,000 times, and visualise uncertainty.

### Setup: DES vs BMS PSA Function

```{r}
#| label: run-des-bms-psa-function

run_des_bms_psa <- function(
    # Clinical parameters
    p_mace_bms, p_mace_des,
    p_ist_bms, p_ist_des,
    # Costs (in Rupees)
    cost_bms, cost_des, cost_mace, cost_ist,
    # Utilities (quality-adjusted life expectancy, QALYs over model horizon)
    util_base, util_post_mace) {

  n_cohort <- 1000

  # BMS arm
  n_bms_mace <- n_cohort * p_mace_bms
  n_bms_ist <- n_cohort * p_ist_bms
  n_bms_uncomplicated <- n_cohort - n_bms_mace - n_bms_ist

  cost_bms_total <- n_cohort * cost_bms +
                    n_bms_mace * cost_mace +
                    n_bms_ist * cost_ist

  qaly_bms_total <- n_bms_uncomplicated * util_base +
                    n_bms_mace * util_post_mace +
                    n_bms_ist * util_post_mace

  # DES arm
  n_des_mace <- n_cohort * p_mace_des
  n_des_ist <- n_cohort * p_ist_des
  n_des_uncomplicated <- n_cohort - n_des_mace - n_des_ist

  cost_des_total <- n_cohort * cost_des +
                    n_des_mace * cost_mace +
                    n_des_ist * cost_ist

  qaly_des_total <- n_des_uncomplicated * util_base +
                    n_des_mace * util_post_mace +
                    n_des_ist * util_post_mace

  return(list(
    cost_bms = cost_bms_total,
    cost_des = cost_des_total,
    qaly_bms = qaly_bms_total,
    qaly_des = qaly_des_total
  ))
}
```

### Running DES vs BMS PSA

```{r}
#| label: run-des-bms-psa-loop

set.seed(2026)

# Define parameter distributions for DES vs BMS
# ============================================================================

# Event rates (Beta distributions)
p_mace_bms_mean <- 0.12
p_mace_des_mean <- 0.08
p_ist_bms_mean <- 0.05
p_ist_des_mean <- 0.02

p_se <- 0.02

beta_mace_bms <- convert_beta_params(p_mace_bms_mean, p_se)
beta_mace_des <- convert_beta_params(p_mace_des_mean, p_se)
beta_ist_bms <- convert_beta_params(p_ist_bms_mean, p_se)
beta_ist_des <- convert_beta_params(p_ist_des_mean, p_se)

# Costs (Gamma distributions)
cost_bms_mean <- 200000
cost_bms_se <- 20000

cost_des_mean <- 320000
cost_des_se <- 30000

cost_mace_mean <- 150000
cost_mace_se <- 25000

cost_ist_mean <- 50000
cost_ist_se <- 10000

gamma_cost_bms <- convert_gamma_params(cost_bms_mean, cost_bms_se)
gamma_cost_des <- convert_gamma_params(cost_des_mean, cost_des_se)
gamma_cost_mace <- convert_gamma_params(cost_mace_mean, cost_mace_se)
gamma_cost_ist <- convert_gamma_params(cost_ist_mean, cost_ist_se)

# Utilities (Beta distributions)
util_base_mean <- 0.92
util_base_se <- 0.03

util_post_mace_mean <- 0.70
util_post_mace_se <- 0.05

beta_util_base <- convert_beta_params(util_base_mean, util_base_se)
beta_util_post_mace <- convert_beta_params(util_post_mace_mean, util_post_mace_se)

# ============================================================================
# RUN PSA: 5,000 iterations
# ============================================================================

n_sim_des_bms <- 5000
psa_des_bms <- data.frame(
  iteration = 1:n_sim_des_bms,
  cost_bms = NA,
  cost_des = NA,
  qaly_bms = NA,
  qaly_des = NA,
  inc_cost = NA,
  inc_qaly = NA,
  icer = NA
)

cat("Running DES vs BMS PSA...\n")

for (i in 1:n_sim_des_bms) {
  # Sample parameters
  p_mace_bms_samp <- rbeta(1, beta_mace_bms["alpha"], beta_mace_bms["beta"])
  p_mace_des_samp <- rbeta(1, beta_mace_des["alpha"], beta_mace_des["beta"])
  p_ist_bms_samp <- rbeta(1, beta_ist_bms["alpha"], beta_ist_bms["beta"])
  p_ist_des_samp <- rbeta(1, beta_ist_des["alpha"], beta_ist_des["beta"])

  cost_bms_samp <- rgamma(1, gamma_cost_bms["shape"], gamma_cost_bms["rate"])
  cost_des_samp <- rgamma(1, gamma_cost_des["shape"], gamma_cost_des["rate"])
  cost_mace_samp <- rgamma(1, gamma_cost_mace["shape"], gamma_cost_mace["rate"])
  cost_ist_samp <- rgamma(1, gamma_cost_ist["shape"], gamma_cost_ist["rate"])

  util_base_samp <- rbeta(1, beta_util_base["alpha"], beta_util_base["beta"])
  util_post_mace_samp <- rbeta(1, beta_util_post_mace["alpha"], beta_util_post_mace["beta"])

  # Run model
  result_des_bms <- run_des_bms_psa(
    p_mace_bms = p_mace_bms_samp,
    p_mace_des = p_mace_des_samp,
    p_ist_bms = p_ist_bms_samp,
    p_ist_des = p_ist_des_samp,
    cost_bms = cost_bms_samp,
    cost_des = cost_des_samp,
    cost_mace = cost_mace_samp,
    cost_ist = cost_ist_samp,
    util_base = util_base_samp,
    util_post_mace = util_post_mace_samp
  )

  psa_des_bms$cost_bms[i] <- result_des_bms$cost_bms
  psa_des_bms$cost_des[i] <- result_des_bms$cost_des
  psa_des_bms$qaly_bms[i] <- result_des_bms$qaly_bms
  psa_des_bms$qaly_des[i] <- result_des_bms$qaly_des
  psa_des_bms$inc_cost[i] <- result_des_bms$cost_des - result_des_bms$cost_bms
  psa_des_bms$inc_qaly[i] <- result_des_bms$qaly_des - result_des_bms$qaly_bms
  psa_des_bms$icer[i] <- psa_des_bms$inc_cost[i] / psa_des_bms$inc_qaly[i]

  if (i %% 1000 == 0) {
    cat(sprintf("Completed %d of %d iterations\n", i, n_sim_des_bms))
  }
}

cat("DES vs BMS PSA complete.\n")
```

### DES vs BMS Results: CE Plane and CEAC

```{r}
#| label: des-bms-ce-results
#| fig-width: 14
#| fig-height: 6
#| message: false

# CE Plane for DES vs BMS
ce_plane_des_bms <- ggplot(psa_des_bms, aes(x = inc_qaly, y = inc_cost)) +
  geom_abline(intercept = 0, slope = 170000, color = "red", linewidth = 1, linetype = "dashed") +
  geom_point(alpha = 0.4, color = "darkgreen", size = 2) +
  geom_hline(yintercept = 0, color = "gray70", linewidth = 0.5) +
  geom_vline(xintercept = 0, color = "gray70", linewidth = 0.5) +
  labs(
    title = "Cost-Effectiveness Plane: DES vs BMS",
    subtitle = paste(n_sim_des_bms, "PSA iterations"),
    x = "Incremental QALYs (DES vs BMS)",
    y = "Incremental Cost (₹)"
  ) +
  scale_y_continuous(labels = scales::comma_format(prefix = "₹")) +
  scale_x_continuous(labels = scales::comma_format()) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    panel.grid.major = element_line(color = "gray95")
  )

# CEAC for DES vs BMS
wtp_range_des_bms <- seq(0, 500000, by = 10000)
ceac_data_des_bms <- data.frame(
  wtp = wtp_range_des_bms,
  prob_cost_effective = NA
)

for (j in 1:nrow(ceac_data_des_bms)) {
  nmb_des <- psa_des_bms$inc_qaly * ceac_data_des_bms$wtp[j] - psa_des_bms$inc_cost
  ceac_data_des_bms$prob_cost_effective[j] <- mean(nmb_des > 0)
}

ceac_plot_des_bms <- ggplot(ceac_data_des_bms, aes(x = wtp, y = prob_cost_effective)) +
  geom_line(color = "darkgreen", linewidth = 1) +
  geom_area(fill = "darkgreen", alpha = 0.3) +
  geom_vline(xintercept = 170000, color = "red", linetype = "dashed", linewidth = 1) +
  labs(
    title = "CEAC: DES vs BMS",
    subtitle = paste(n_sim_des_bms, "PSA iterations"),
    x = "Willingness-to-Pay Threshold (₹ per QALY)",
    y = "Probability That DES is Cost-Effective"
  ) +
  scale_x_continuous(labels = scales::comma_format(prefix = "₹")) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", size = 13),
    panel.grid.major = element_line(color = "gray95")
  )

# Print side-by-side
library(gridExtra)
grid.arrange(ce_plane_des_bms, ceac_plot_des_bms, ncol = 2)

# Summary
prob_des_ce_at_170k <- mean((psa_des_bms$inc_qaly * 170000 - psa_des_bms$inc_cost) > 0)
cat("\n\nDES vs BMS at WTP = ₹170,000/QALY:\n")
cat(sprintf("Mean ICER: ₹%s/QALY\n", format(round(mean(psa_des_bms$icer)), big.mark=",")))
cat(sprintf("Probability that DES is cost-effective: %.1f%%\n", prob_des_ce_at_170k * 100))
```

## What You Just Did

You have completed a full probabilistic sensitivity analysis from end to end:

1. **Defined parameter distributions** for probabilities (Beta), costs (Gamma), and relative effects (Log-Normal).
2. **Parameterised those distributions** using means and standard errors from literature or data.
3. **Wrote a function** (`run_markov_psa`) that accepts a set of sampled parameters and returns model outcomes.
4. **Ran 5,000 PSA iterations** in a loop, sampling from each distribution and recording results.
5. **Visualised uncertainty** using the cost-effectiveness plane and acceptability curve.
6. **Interpreted the results** by reading the probability that each strategy is cost-effective across a range of WTP thresholds.

This is exactly what NICE, CADTH, India's NITI Aayog, and all major HTA agencies ask for in your submissions. In Excel, each of these steps would have been cumbersome. In R, it is transparent, reproducible, and takes seconds.

## Key References

- Briggs, A. H., Claxton, K., & Sculpher, M. J. (2006). Thinking outside the box: Considering uncertainty and heterogeneity in health economic evaluation. *Pharmacoeconomics*, 24(11), 1079-1090.

- Fenwick, E., O'Brien, B. J., & Briggs, A. (2004). Cost-effectiveness acceptability curves—fact or fiction? *Health Economics*, 13(5), 473-476.

- Claxton, K., Ginnelly, L., Sculpher, M., Philips, Z., & Palmer, S. (2005). A pilot study on the use of decision theory and value of information analysis as part of the NHS Health Technology Assessment programme. *Health Technology Assessment*, 9(9).

