---
title: "Probabilistic Sensitivity Analysis"
subtitle: "Why uncertainty in parameters demands a distribution, not a point estimate"
author: "R for HTA (Basics) — Workshop 2026"
institute: "RRC-HTA, AIIMS Bhopal | HTAIn, DHR"
format:
  revealjs:
    theme: [default, ../../custom-theme.scss]
    slide-number: true
    footer: "R for HTA (Basics) — RRC-HTA, AIIMS Bhopal"
    logo: ""
    transition: fade
    width: 1280
    height: 720
    code-fold: true
    code-overflow: wrap
    highlight-style: github
    fig-align: center
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

## Why Probabilistic Sensitivity Analysis?

When you submit a cost-effectiveness model to NICE, CADTH, or India's NITI Aayog, you make implicit claims:

. . .

*"This transition probability is 0.10. This cost is ₹45,000. This utility weight is 0.72."*

. . .

**But are they really?**

- Clinical trials have limited sample sizes
- Cost data come from specific regions and time periods
- Utility estimates vary across populations

. . .

**The question decision-makers now demand:**

If true parameters vary within plausible ranges, does your recommendation change? How confident are you really?

## Point Estimates vs Distributions

**Old approach (still common in Excel):**
- One transition probability: 0.10
- One cost: ₹45,000
- One result: ICER ₹80,000/QALY

. . .

**Modern approach (PSA):**
- Transition probability drawn from Beta(α, β)
- Cost drawn from Gamma(shape, rate)
- Run model 5,000 times → distribution of ICERs

. . .

**Gain:** Not just "the ICER is ₹80,000" but "there is 85% probability the intervention is cost-effective at ₹1,70,000/QALY"

## The PSA Workflow

```{r}
#| label: fig-psa-workflow
#| echo: false
#| eval: true
#| fig-width: 12
#| fig-height: 5

library(DiagrammeR)

grViz("
digraph psa_workflow {
  graph [rankdir=LR, bgcolor='transparent', fontname='Helvetica']
  node [shape=box, style='filled,rounded', fontname='Helvetica', fontsize=11]

  A [label='1. Define\nDistributions\nfor each parameter', fillcolor='#4e79a7', fontcolor='white']
  B [label='2. Sample\nParameters\n(random draw)', fillcolor='#59a14f', fontcolor='white']
  C [label='3. Run\nModel\n(one iteration)', fillcolor='#f28e2b', fontcolor='white']
  D [label='4. Store\nResults\n(cost, QALY, ICER)', fillcolor='#e15759', fontcolor='white']
  E [label='5. Repeat\n1000-10000\ntimes', fillcolor='#76b7b2', fontcolor='white']
  F [label='6. Summarise\nUncertainty\n(CE plane, CEAC)', fillcolor='#edc948', fontcolor='black']

  A -> B -> C -> D -> E -> F
  E -> B [style=dashed, label='Loop', fontsize=9]
}
")
```

## Which Distribution for Which Parameter?

```{r}
#| label: fig-distribution-map
#| echo: false
#| eval: true
#| fig-width: 12
#| fig-height: 5

grViz("
digraph dist_map {
  graph [rankdir=TB, bgcolor='transparent', fontname='Helvetica']
  node [fontname='Helvetica', fontsize=11, style='filled,rounded', shape=box]

  prob [label='Probabilities &\nUtilities\n(bounded 0-1)', fillcolor='#4e79a7', fontcolor='white', width=2.2]
  cost [label='Costs\n(non-negative,\nright-skewed)', fillcolor='#59a14f', fontcolor='white', width=2.2]
  hr [label='Hazard Ratios &\nRelative Risks\n(positive, often <1)', fillcolor='#f28e2b', fontcolor='white', width=2.2]

  beta [label='Beta(α, β)', fillcolor='#d4e6f1', shape=ellipse, width=1.8, fontsize=12]
  gamma [label='Gamma(shape, rate)', fillcolor='#d5f5e3', shape=ellipse, width=1.8, fontsize=12]
  lognorm [label='Log-Normal(μ, σ)', fillcolor='#fdebd0', shape=ellipse, width=1.8, fontsize=12]

  prob -> beta
  cost -> gamma
  hr -> lognorm
}
")
```

## Beta Distribution for Probabilities

**When:** Transition probabilities, event rates (bounded 0–1)

. . .

**Convert from mean and SE:**
$$\alpha = \mu \left(\frac{\mu(1-\mu)}{\sigma^2} - 1\right), \quad \beta = (1-\mu) \left(\frac{\mu(1-\mu)}{\sigma^2} - 1\right)$$

. . .

**R code example:**

```r
# p = 0.10 (transition Early → Moderate), SE = 0.02
mean_p <- 0.10
se_p <- 0.02

alpha <- mean_p * (mean_p * (1 - mean_p) / (se_p^2) - 1)
beta <- (1 - mean_p) * (mean_p * (1 - mean_p) / (se_p^2) - 1)

samples <- rbeta(5000, alpha, beta)
```

## Gamma Distribution for Costs

**When:** Annual costs (non-negative, right-skewed)

. . .

**Convert from mean and SE:**
$$\text{shape} = \frac{\mu^2}{\sigma^2}, \quad \text{rate} = \frac{\mu}{\sigma^2}$$

. . .

**R code example:**

```r
# cost_moderate = ₹45,000, SE = ₹5,000
mean_cost <- 45000
se_cost <- 5000

shape <- mean_cost^2 / (se_cost^2)
rate <- mean_cost / (se_cost^2)

samples <- rgamma(5000, shape = shape, rate = rate)
```

## Log-Normal Distribution for Hazard Ratios

**When:** Hazard ratios, relative risks (positive, usually < 1)

. . .

**Convert from point estimate and 95% CI:**
$$\text{SE}_{\log(\text{HR})} = \frac{\log(\text{upper}) - \log(\text{lower})}{3.92}$$

. . .

**R code example:**

```r
# HR = 0.66, 95% CI (0.55, 0.79)
hr_mean <- log(0.66)
log_se <- (log(0.79) - log(0.55)) / 3.92

samples <- rlnorm(5000, meanlog = hr_mean, sdlog = log_se)
```

## Example: Parameter Distributions

```{r}
#| label: example-dists
#| echo: false
#| eval: true
#| fig-width: 13
#| fig-height: 5

library(ggplot2)
library(dplyr)

set.seed(2026)

# Beta example
mean_p <- 0.10
se_p <- 0.02
var_p <- se_p^2
alpha_beta <- mean_p * (mean_p * (1 - mean_p) / var_p - 1)
beta_beta <- (1 - mean_p) * (mean_p * (1 - mean_p) / var_p - 1)
samples_beta <- rbeta(3000, alpha_beta, beta_beta)

# Gamma example
mean_cost <- 45000
se_cost <- 5000
shape_gamma <- mean_cost^2 / (se_cost^2)
rate_gamma <- mean_cost / (se_cost^2)
samples_gamma <- rgamma(3000, shape = shape_gamma, rate = rate_gamma)

# Log-Normal example
hr_mean <- log(0.66)
log_se <- (log(0.79) - log(0.55)) / 3.92
samples_lognormal <- rlnorm(3000, meanlog = hr_mean, sdlog = log_se)

# Combine
dist_data <- data.frame(
  Distribution = rep(c("Beta (Probability)", "Gamma (Cost ₹)", "Log-Normal (HR)"),
                     c(3000, 3000, 3000)),
  Value = c(samples_beta, samples_gamma, samples_lognormal)
)

ggplot(dist_data, aes(x = Value, fill = Distribution)) +
  geom_density(alpha = 0.6, color = NA) +
  facet_wrap(~ Distribution, scales = "free") +
  labs(
    title = "Example Parameter Distributions for PSA",
    subtitle = "Transition Probability (p) | Cost (₹) | Hazard Ratio (HR)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

## The PSA Loop (Pseudocode)

```r
for (i in 1:5000) {
  # Sample parameters from distributions
  p_12_sampled <- rbeta(1, alpha_12, beta_12)
  cost_moderate_sampled <- rgamma(1, shape_m, rate_m)
  hr_sampled <- rlnorm(1, log_hr_mean, log_hr_se)
  # ... sample all other parameters ...

  # Run model with these sampled values
  result <- run_markov(p_12_sampled, cost_moderate_sampled, hr_sampled, ...)

  # Store outcomes
  results_df[i, "inc_cost"] <- result$inc_cost
  results_df[i, "inc_qaly"] <- result$inc_qaly
  results_df[i, "inc_nmb"]  <- wtp * result$inc_qaly - result$inc_cost
}
```

. . .

**Outcome:** 5,000 rows of inc_cost, inc_qaly, and **incremental NMB**.

**Why NMB, not ICER?** A few iterations with near-zero ΔQALYs produce extreme ICERs that distort the average. NMB has no such problem.

## Cost-Effectiveness Plane Explained

:::{.columns}

:::{.column width="60%"}

The **CE plane** plots all 5,000 PSA iterations as points:
- **X-axis:** Incremental QALYs (treatment gained)
- **Y-axis:** Incremental cost (₹)

:::

:::{.column width="40%"}

**Four quadrants:**
- **Bottom-right (SE):** DOMINANT (better + cheaper)
- **Top-right (NE):** Cost-effective if below WTP line
- **Top-left (NW):** DOMINATED (worse + costlier)
- **Bottom-left (SW):** Trade-off (cheaper but worse)

:::

:::

. . .

**Red dashed line = WTP threshold (₹1,70,000/QALY in India)**

Points below the line are cost-effective; above are not.

## Cost-Effectiveness Acceptability Curve (CEAC)

**Question answered by CEAC:**

"Across different willingness-to-pay thresholds, what is the probability the intervention is cost-effective?"

. . .

**Method:** For each WTP threshold, calculate using **NMB**:

$$\text{Prob}_{\text{CE}} = \frac{\text{# iterations with } \Delta\text{NMB} > 0}{\text{total iterations}}$$

where ΔNMB = WTP × ΔQALYs − ΔCost (equivalent to ICER < WTP but works in all quadrants)

. . .

**Reading the CEAC:**
- At ₹0/QALY threshold: 0% (almost never cost-effective)
- At ₹1,70,000/QALY: 85% (highly likely)
- At ₹5,00,000/QALY: >95% (very confident)

. . .

This **directly answers** decision-makers' uncertainty.

## PSA in Excel vs R

**Excel approach:**
- Create 5,000 rows of random numbers for each parameter
- Use VLOOKUP or INDEX/MATCH to match to distributions
- Copy model formula down 5,000 times
- Hope formatting doesn't break...

. . .

**R approach:**
```r
for (i in 1:5000) {
  params_sampled <- c(rbeta(...), rgamma(...), rlnorm(...))
  results[i] <- run_model(params_sampled)
}
```

. . .

**Advantages of R:**
- Transparent and reproducible
- Far fewer manual steps
- Easy to modify or re-run
- Direct integration with visualization

## Key Takeaways: PSA

::::{.key-insight}

1. **PSA answers uncertainty:** Not just "what is the ICER?" but "given parameter uncertainty, what is the probability the intervention is cost-effective?"

2. **Choose distributions carefully:** Beta for probabilities, Gamma for costs, Log-Normal for relative effects

3. **Parameterize correctly:** Converting mean/SE to shape parameters is critical (and error-prone in Excel)

4. **Interpret results properly:**
   - CE plane shows scatter of outcomes
   - CEAC shows decision robustness across WTP thresholds

5. **This is now the standard:** NICE, CADTH, NITI Aayog all require PSA in submissions

::::

---

**Next:** Session 7 — Using Generative AI for HTA coding (AI as your R assistant)
