---
title: "GDM Diagnostic Decision Tree"
subtitle: "Universal vs risk-based screening for gestational diabetes in India"
author: "R for HTA (Basics) — Workshop 2026"
institute: "RRC-HTA, AIIMS Bhopal | HTAIn, DHR"
format:
  revealjs:
    theme: [default, ../../custom-theme.scss]
    slide-number: true
    footer: "R for HTA (Basics) — RRC-HTA, AIIMS Bhopal"
    logo: ""
    transition: fade
    width: 1280
    height: 720
    code-fold: true
    code-overflow: wrap
    highlight-style: github
    fig-align: center
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

## The Clinical Question

Gestational diabetes mellitus (GDM) affects 10–14% of pregnancies in India.

. . .

**Much higher than Western countries.**

. . .

Undetected GDM increases risks of:
- Macrosomia and birth injuries
- Neonatal hypoglycaemia
- Pre-eclampsia
- Future type 2 diabetes in mother

## The HTA Question

**Should India implement universal screening for GDM using a fasting 75g OGTT, or is risk-based screening more cost-effective?**

. . .

This is a classic **diagnostic decision tree** problem.

The kind of analysis you might currently do in Excel or TreeAge — but now in R.

## Three Strategies Compared

1. **Universal screening**
   - All pregnant women get OGTT at 24–28 weeks

. . .

2. **Risk-based screening**
   - Only high-risk women (age >25, BMI >25, family history) get OGTT

. . .

3. **No formal screening**
   - GDM detected only when symptomatic (status quo in many settings)

## Key Parameters

```{r}
#| label: fig-params-table
#| echo: false
#| eval: true

library(knitr)
params <- data.frame(
  Parameter = c("GDM prevalence", "High-risk women", "OGTT sensitivity", "OGTT specificity",
                 "Screening cost", "Treatment cost", "Adverse outcome (if treated)",
                 "Adverse outcome (if missed)"),
  Value = c("12%", "60%", "90%", "85%", "₹250", "₹8,000", "10%", "35%")
)
kable(params, col.names = c("Parameter", "Value"))
```

All sourced from **Indian studies** where available.

## The Decision Tree Structure

```{r}
#| label: fig-gdm-tree
#| eval: true
#| fig-cap: "GDM screening decision tree"

library(DiagrammeR)

grViz("
digraph gdm_tree {
  graph [rankdir=LR, bgcolor='transparent', fontname='Helvetica', nodesep=0.5]
  node [fontname='Helvetica', fontsize=10]

  D [label='Screening\nStrategy', shape=square, style=filled, fillcolor='#4e79a7', fontcolor='white', width=1.2]

  S1 [label='Universal\nScreening', shape=box, style='filled,rounded', fillcolor='#59a14f', fontcolor='white']
  S2 [label='Risk-Based\nScreening', shape=box, style='filled,rounded', fillcolor='#f28e2b', fontcolor='white']
  S3 [label='No Formal\nScreening', shape=box, style='filled,rounded', fillcolor='#e15759', fontcolor='white']

  C1 [label='GDM+\n(12%)', shape=circle, style=filled, fillcolor='#d4e6f1', width=0.8, fontsize=9]
  C2 [label='GDM-\n(88%)', shape=circle, style=filled, fillcolor='#d5f5e3', width=0.8, fontsize=9]

  TP [label='Test +\nTreat', shape=box, style='filled,rounded', fillcolor='#aed6f1', fontsize=9]
  FN [label='Test -\nMissed', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=9]

  D -> S1
  D -> S2
  D -> S3

  S1 -> C1
  S1 -> C2
  C1 -> TP
  C1 -> FN
}
")
```

## Strategy 1: Universal Screening

Every woman gets an OGTT. Those who test positive receive treatment.

. . .

**Test results in 10,000 women:**

- **True positives** (GDM detected): 12% × 90% = 10.8% of cohort
- **False negatives** (GDM missed): 12% × 10% = 1.2% of cohort
- **False positives** (wrongly flagged): 88% × 15% = 13.2% of cohort
- **True negatives** (correctly cleared): 88% × 85% = 74.8% of cohort

## Cost Calculation: Universal

```{r}
#| label: universal-costs
#| echo: true
#| eval: false

# Screening cost: everyone gets OGTT
cost_screening_universal <- n_cohort * cost_ogtt

# Treatment costs: true positives + false positives
cost_treatment_universal <- (true_positive + false_positive) * cost_gdm_treatment

# Adverse outcomes
adverse_outcomes_universal <- (true_positive * p_adverse_gdm_treated) +
                              (false_negative * p_adverse_gdm_untreated) +
                              (false_positive * p_adverse_no_gdm) +
                              (true_negative * p_adverse_no_gdm)

cost_outcomes_universal <- (adverse_outcomes_universal * cost_adverse) +
                           ((n_cohort - adverse_outcomes_universal) * cost_no_adverse)

total_cost_universal <- cost_screening_universal + cost_treatment_universal +
                        cost_outcomes_universal
```

## QALY Calculation: Universal

```{r}
#| label: universal-qalys
#| echo: true
#| eval: false

# QALYs by outcome group
qaly_tp <- true_positive * ((1 - p_adverse_gdm_treated) * utility_gdm_treated +
                             p_adverse_gdm_treated * utility_adverse)

qaly_fn <- false_negative * ((1 - p_adverse_gdm_untreated) * utility_no_adverse +
                              p_adverse_gdm_untreated * utility_adverse)

qaly_fp <- false_positive * ((1 - p_adverse_no_gdm) * utility_gdm_treated +
                              p_adverse_no_gdm * utility_adverse)

qaly_tn <- true_negative * ((1 - p_adverse_no_gdm) * utility_no_adverse +
                             p_adverse_no_gdm * utility_adverse)

total_qaly_universal <- qaly_tp + qaly_fn + qaly_fp + qaly_tn
```

## Strategy 2: Risk-Based Screening

Only high-risk women are offered OGTT.
Low-risk women are NOT screened → GDM undetected in ~60% of cases.

. . .

```{r}
#| label: risk-based-logic
#| echo: true
#| eval: false

# High-risk (60% of cohort)
n_high_risk <- n_cohort * prop_high_risk
# Among high-risk: 18% have GDM (higher prevalence)
# Among low-risk: 3% have GDM (lower prevalence)

# Screened (high-risk only)
tp_hr <- n_high_risk * prev_gdm_high_risk * sens_ogtt
fn_hr <- n_high_risk * prev_gdm_high_risk * (1 - sens_ogtt)
fp_hr <- n_high_risk * (1 - prev_gdm_high_risk) * (1 - spec_ogtt)

# NOT screened (low-risk) → all GDM missed
n_gdm_lr <- n_cohort * prop_low_risk * prev_gdm_low_risk  # All undetected
```

## Strategy 3: No Formal Screening

GDM detected only through clinical symptoms (~25% detection rate).

No screening cost, but many cases missed.

. . .

**Result:** Lowest upfront cost, but highest adverse outcome rate due to missed cases.

## Results Comparison

All three strategies produce:
- **Costs** (screening, treatment, outcome management)
- **QALYs** (quality of life impact)
- **Adverse outcomes** (births with complications)

. . .

Then we calculate **ICERs** to compare cost-effectiveness.

## ICER Calculations

Using **No Screening as the reference**:

. . .

```{r}
#| label: icers
#| echo: true
#| eval: false

# Risk-Based vs No Screening
icer_risk_vs_none <- (total_cost_risk - total_cost_none) /
                     (total_qaly_risk - total_qaly_none)

# Universal vs No Screening
icer_univ_vs_none <- (total_cost_universal - total_cost_none) /
                     (total_qaly_universal - total_qaly_none)

# Universal vs Risk-Based
icer_univ_vs_risk <- (total_cost_universal - total_cost_risk) /
                     (total_qaly_universal - total_qaly_risk)
```

## Interpreting the ICER — 4 Quadrants

A negative ICER can mean **two opposite things**:

| | ΔQALYs > 0 (better) | ΔQALYs < 0 (worse) |
|---|---|---|
| **ΔCost > 0** | NE: compare to WTP | **DOMINATED** |
| **ΔCost < 0** | **DOMINANT** | Trade-off |

. . .

**Always check the signs of ΔCost and ΔQALYs separately!**

## Net Monetary Benefit (NMB)

NMB removes the ICER ambiguity. Positive ΔNMB = adopt.

$$\text{NMB} = \text{WTP} \times \text{QALYs} - \text{Cost}$$

. . .

**Highest NMB = optimal strategy** — works for any number of comparators.

No more pairwise ICER confusion with 3+ strategies.

## Key Insight: Trade-offs

**Universal screening:**
- Highest cost (everyone screened)
- Lowest adverse outcome rate (most cases detected)
- BUT some false positives (unnecessary treatment)

. . .

**Risk-based screening:**
- Lower cost (fewer screened)
- Intermediate outcome rate (misses low-risk cases)

. . .

**No screening:**
- Lowest cost
- Highest adverse outcome rate (clinical detection only 25%)

## What You Just Did

1. ✓ Defined clinical question using Indian epidemiological data
2. ✓ Structured decision tree with three screening strategies
3. ✓ Populated it with parameters from published Indian literature
4. ✓ Calculated costs, QALYs, ICERs with 4-quadrant interpretation
5. ✓ Computed NMB ranking across all strategies
6. ✓ Visualised results on cost-effectiveness plane

## The Power of R

**In Excel:** Changing GDM prevalence requires tracing through multiple cells across sheets.

. . .

**In R:** Change one number at the top and re-run the entire analysis.

. . .

This is where R shines for HTA.

## Key Takeaway

A diagnostic decision tree is a **reusable template**:

. . .

- **Disease screening** (GDM, diabetes, cancer)
- **Diagnostic testing** (sensitivity, specificity, costs)
- **Treatment decisions** (treat if positive, manage outcomes)

. . .

Every diagnostic HTA follows this structure — just with different parameters.

## Bonus: rdecision Package

The same GDM model rebuilt with the `rdecision` package:

- **Object-oriented tree** — DecisionNode, ChanceNode, LeafNode
- **Built-in PSA** — Monte Carlo simulation with 1,000 iterations
- **CE plane + CEAC** — decision uncertainty visualised

. . .

→ See the **Session 3 Bonus** page on the workshop website.
