---
title: "Session 2: R Orientation"
subtitle: "Getting comfortable with R and RStudio — just enough to follow along"
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
---

## What This Session Is (and Isn't)

This session is **not** an R programming course. You do not need to memorise anything here. The goal is to build enough familiarity that when you see R code in the following sessions, you can:

- Recognise what it is doing
- Know where to change an input value
- Run it yourself and see the output
- Not be intimidated by it

Think of this as learning to read a new language well enough to follow a conversation — not to write poetry.

## The RStudio Interface

When you open RStudio, you see four panels:

- **Source (top-left):** Where you write and edit code files (.R, .qmd)
- **Console (bottom-left):** Where code actually runs and output appears
- **Environment (top-right):** Shows variables and data currently in memory
- **Files/Plots/Help (bottom-right):** File browser, plot viewer, package help

```{r}
#| label: fig-rstudio-layout
#| echo: true
#| fig-cap: "The four panels of RStudio"

library(DiagrammeR)

grViz("
digraph rstudio {
  graph [bgcolor='transparent', fontname='Helvetica']
  node [shape=box, style='filled', fontname='Helvetica', fontsize=11]

  subgraph cluster_top {
    label='Top Row'
    style=invis
    source [label='SOURCE EDITOR\n(Write & edit code)\n.R and .qmd files', fillcolor='#4e79a7', fontcolor='white', width=3]
    env [label='ENVIRONMENT\n(Variables in memory)\nData you have created', fillcolor='#59a14f', fontcolor='white', width=3]
  }

  subgraph cluster_bottom {
    label='Bottom Row'
    style=invis
    console [label='CONSOLE\n(Run code here)\nSee text output', fillcolor='#f28e2b', fontcolor='white', width=3]
    output [label='FILES / PLOTS / HELP\n(See visualisations)\nBrowse files', fillcolor='#e15759', fontcolor='white', width=3]
  }

  source -> console [style=invis]
  env -> output [style=invis]
}
")
```

For this workshop, you will mostly:

1. Open a `.qmd` file in the Source panel
2. Run code chunks by clicking the green "play" button
3. See results in the Console or Plots panel

## Running Your First Code

Let us make sure everything works. Run this code chunk:

```{r}
#| label: hello-hta
#| echo: true

# This is a comment — R ignores everything after #
# Comments explain what the code does

message("Welcome to R for HTA!")
```

If you see the message, your R installation is working.

## The Basics: Variables and Arithmetic

In R, you store values using the assignment operator `<-` :

```{r}
#| label: variables
#| echo: true

# Store some values
drug_cost <- 5000       # Cost of the drug in ₹
admin_cost <- 800       # Administration cost in ₹
total_cost <- drug_cost + admin_cost

# Print the result
total_cost
```

When you see code like this in later sessions, you will know: a value is being stored under a name, and you can change that value by editing the number.

### Quick reference of common operations

```{r}
#| label: arithmetic
#| echo: true

10 + 3       # Addition
10 - 3       # Subtraction
10 * 3       # Multiplication
10 / 3       # Division
10 ^ 2       # Exponentiation (10 squared)
sqrt(100)    # Square root
round(3.14159, 2)  # Round to 2 decimal places
```

## Vectors: Multiple Values at Once

A vector is simply a collection of values. This is how R handles data efficiently:

```{r}
#| label: vectors
#| echo: true

# Cost of five different treatments
treatment_costs <- c(1200, 3400, 5600, 2300, 4100)

# Basic operations on vectors
mean(treatment_costs)    # Average
sum(treatment_costs)     # Total
min(treatment_costs)     # Minimum
max(treatment_costs)     # Maximum
length(treatment_costs)  # How many values
```

In HTA modelling, vectors are everywhere — transition probabilities, costs per cycle, utility values across health states.

## Data Frames: Tables in R

If you are used to Excel spreadsheets, a **data frame** is the R equivalent of a table:

```{r}
#| label: dataframe
#| echo: true

# Create a simple data frame — like an Excel table
interventions <- data.frame(
  strategy    = c("Standard Care", "New Screening", "Enhanced Treatment"),
  cost        = c(1200, 2500, 4800),
  qalys       = c(0.78, 0.85, 0.91),
  source      = c("Sharma et al. 2023", "Patel et al. 2024", "Gupta et al. 2022")
)

# View it
interventions
```

You can access specific columns:

```{r}
#| label: access-columns
#| echo: true

# Access the cost column
interventions$cost

# Calculate mean cost
mean(interventions$cost)
```

```{r}
#| label: plot-interventions
#| echo: true
#| fig-cap: "Visualising the interventions data frame"

library(ggplot2)

ggplot(interventions, aes(x = qalys, y = cost, label = strategy)) +
  geom_point(size = 4, colour = "#4e79a7") +
  geom_text(vjust = -1.2, size = 3.5) +
  labs(x = "QALYs", y = "Cost (₹)",
       title = "Cost vs QALYs for Three Interventions",
       subtitle = "Your first cost-effectiveness plot in R") +
  expand_limits(y = c(0, 6000)) +
  theme_minimal()
```

## Installing and Loading Packages

R packages are like add-ons that extend what R can do. For HTA, key packages include `heemod`, `ggplot2`, `BCEA`, and others.

```{r}
#| label: packages
#| echo: true
#| eval: false

# Install a package (only need to do this once)
install.packages("ggplot2")

# Load a package (need to do this each session)
library(ggplot2)
```

**Analogy:** Installing is like downloading an app. Loading is like opening it.

In this workshop, we will tell you which packages to load at the start of each session. You do not need to find or choose packages yourself.

## Reading Code: A Practical Guide

You will spend more time **reading** R code than writing it in this workshop. Here are patterns you will see repeatedly:

### Pattern 1: Defining model parameters

```{r}
#| label: pattern-parameters
#| echo: true

# Transition probabilities (from Agarwal et al. 2021)
p_mild_to_moderate <- 0.15    # Annual probability of progression
p_moderate_to_severe <- 0.10
p_severe_to_death <- 0.25
p_background_mortality <- 0.02

# Costs in ₹ (from PMJAY 2023 schedule)
c_mild <- 5000      # Annual cost of managing mild disease
c_moderate <- 25000
c_severe <- 150000   # Includes dialysis
```

When you see this, you know: **these are the model inputs, and you can change these numbers.**

### Pattern 2: A loop that runs the model over cycles

```{r}
#| label: pattern-loop
#| echo: true

# Simple example: accumulating costs over 10 cycles
n_cycles <- 10
annual_cost <- 5000
discount_rate <- 0.03
total_discounted_cost <- 0

for (cycle in 1:n_cycles) {
  discounted <- annual_cost / (1 + discount_rate)^cycle
  total_discounted_cost <- total_discounted_cost + discounted
  cat("Cycle", cycle, ": Discounted cost = ₹", round(discounted, 0), "\n")
}

cat("\nTotal discounted cost over", n_cycles, "cycles: ₹",
    round(total_discounted_cost, 0), "\n")
```

When you see a `for` loop, you know: **the model is repeating a calculation across time cycles.** You do not need to understand every line — just the logic.

### Pattern 3: Creating a plot

```{r}
#| label: pattern-plot
#| echo: true
#| fig-cap: "Discounted annual costs over 10 years"

library(ggplot2)

cycle_data <- data.frame(
  cycle = 1:10,
  discounted_cost = annual_cost / (1 + discount_rate)^(1:10)
)

ggplot(cycle_data, aes(x = cycle, y = discounted_cost)) +
  geom_col(fill = "steelblue") +
  labs(x = "Year", y = "Discounted Cost (₹)",
       title = "Effect of Discounting on Annual Costs") +
  scale_x_continuous(breaks = 1:10) +
  theme_minimal()
```

When you see `ggplot`, you know: **a chart is being created.** The `labs()` part sets the labels, `aes()` maps data to axes.

```{r}
#| label: fig-workflow
#| echo: true
#| fig-cap: "The workshop workflow: Read, Modify, Run"

grViz("
digraph workflow {
  graph [rankdir=LR, bgcolor='transparent', fontname='Helvetica']
  node [shape=box, style='filled,rounded', fontname='Helvetica', fontsize=12]

  A [label='READ\nthe code\n(understand logic)', fillcolor='#4e79a7', fontcolor='white']
  B [label='MODIFY\na parameter\n(change a number)', fillcolor='#59a14f', fontcolor='white']
  C [label='RUN\nthe chunk\n(click play)', fillcolor='#f28e2b', fontcolor='white']
  D [label='INTERPRET\nthe output\n(what changed?)', fillcolor='#e15759', fontcolor='white']

  A -> B -> C -> D
  D -> B [style=dashed, label='Try again', fontsize=9]
}
")
```

## Working with .qmd Files

All workshop materials are in **Quarto (.qmd)** files. These combine:

- **Text** (explanations, like this paragraph)
- **Code chunks** (the grey blocks with R code)
- **Outputs** (results, tables, plots)

To work with a .qmd file:

1. Open it in RStudio
2. Read the text explanations
3. Run code chunks one at a time (click the green play button ▶)
4. See the output appear below each chunk
5. Modify values and re-run to see what changes

You can also click **"Render"** to produce a complete HTML document with all code, outputs, and text formatted together.

## What You Need to Remember

Almost nothing from this session needs to be memorised. Instead, remember these three things:

1. **Values are stored with `<-`** and you can change them by editing the number
2. **Code runs top to bottom** — run chunks in order
3. **If something breaks, look at the error message** — it often tells you what went wrong (a missing package, a typo in a variable name, a file not found)

Everything else, you can look up as needed — or ask an AI tool, as we will discuss in Session 7.

## Try It Yourself

Before we move to Session 3, try modifying the code below. Change the costs and QALYs to see how the ICER changes:

```{r}
#| label: try-it
#| echo: true

# YOUR TASK: Change these values and re-run
cost_A <- 3000    # Try changing this
cost_B <- 1500    # Try changing this
qaly_A <- 0.82    # Try changing this
qaly_B <- 0.75    # Try changing this

# ICER calculation (do not modify below this line)
icer <- (cost_A - cost_B) / (qaly_A - qaly_B)
cat("ICER = ₹", round(icer, 0), "per QALY gained\n")

# Is it cost-effective at a WTP of ₹1,00,000 per QALY?
wtp <- 100000
if (icer < wtp) {
  cat("→ Cost-effective at a WTP threshold of ₹", format(wtp, big.mark = ","), "per QALY\n")
} else {
  cat("→ NOT cost-effective at a WTP threshold of ₹", format(wtp, big.mark = ","), "per QALY\n")
}
```

If you managed to run this and change the result — you are ready for the rest of the workshop.
