---
title: "R Orientation"
subtitle: "Getting comfortable with R and RStudio — just enough to follow along"
author: "R for HTA (Basics) — Workshop 2026"
institute: "RRC-HTA, AIIMS Bhopal | HTAIn, DHR"
format:
  revealjs:
    theme: [default, ../../custom-theme.scss]
    slide-number: true
    footer: "R for HTA (Basics) — RRC-HTA, AIIMS Bhopal"
    logo: ""
    transition: fade
    width: 1280
    height: 720
    code-fold: true
    code-overflow: wrap
    highlight-style: github
    fig-align: center
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

## What This Session Is (and Isn't)

This is **not** an R programming course.

. . .

You do not need to memorise anything here.

. . .

**Goal:** Build enough familiarity that when you see R code:

- You can recognise what it is doing
- You can change an input value
- You can run it and see output
- You are not intimidated by it

. . .

Think of this as learning to **read a language** well enough to follow a conversation — not to write poetry.

## The RStudio Interface

```{r}
#| label: fig-rstudio-layout
#| eval: true
#| fig-cap: "The four panels of RStudio"

library(DiagrammeR)

grViz("
digraph rstudio {
  graph [bgcolor='transparent', fontname='Helvetica']
  node [shape=box, style='filled', fontname='Helvetica', fontsize=11]

  source [label='SOURCE EDITOR\n(Write & edit code)\n.R and .qmd files',
          fillcolor='#4e79a7', fontcolor='white', width=3]
  env [label='ENVIRONMENT\n(Variables in memory)\nData you have created',
       fillcolor='#59a14f', fontcolor='white', width=3]
  console [label='CONSOLE\n(Run code here)\nSee text output',
           fillcolor='#f28e2b', fontcolor='white', width=3]
  output [label='FILES / PLOTS / HELP\n(See visualisations)\nBrowse files',
          fillcolor='#e15759', fontcolor='white', width=3]

  source -> console [style=invis]
  env -> output [style=invis]
}
")
```

## Running Your First Code

```{r}
#| label: hello-hta
#| echo: true

# This is a comment — R ignores everything after #
message("Welcome to R for HTA!")
```

If you see the message, your R installation is working.

## Variables and Assignment

In R, you store values using `<-` :

```{r}
#| label: variables
#| echo: true

# Store some values
drug_cost <- 5000       # Cost of the drug in ₹
admin_cost <- 800       # Administration cost in ₹
total_cost <- drug_cost + admin_cost

# Print the result
total_cost
```

When you see this: **a value is stored, and you can change that number.**

## Vectors: Collections of Values

A vector is a collection of values. This is how R handles data efficiently:

```{r}
#| label: vectors
#| echo: true

# Cost of five different treatments
treatment_costs <- c(1200, 3400, 5600, 2300, 4100)

# Operations on vectors
mean(treatment_costs)    # Average
sum(treatment_costs)     # Total
min(treatment_costs)     # Minimum
max(treatment_costs)     # Maximum
```

In HTA modelling: transition probabilities, costs per cycle, utility values — all vectors.

## Data Frames: Tables in R

If you are used to Excel spreadsheets, a **data frame** is the R equivalent:

```{r}
#| label: dataframe
#| echo: true

# Create a data frame — like an Excel table
interventions <- data.frame(
  strategy = c("Standard Care", "New Screening", "Enhanced"),
  cost     = c(1200, 2500, 4800),
  qalys    = c(0.78, 0.85, 0.91)
)

interventions
```

## Accessing Data Frames

```{r}
#| label: access-columns
#| echo: true

# Access the cost column
interventions$cost

# Calculate mean cost
mean(interventions$cost)
```

The `$` symbol means "get this column from that data frame".

## Plotting with ggplot2

```{r}
#| label: plot-interventions
#| echo: true
#| eval: false

library(ggplot2)

ggplot(interventions, aes(x = qalys, y = cost, label = strategy)) +
  geom_point(size = 4, colour = "#4e79a7") +
  geom_text(vjust = -1.2, size = 3.5) +
  labs(x = "QALYs", y = "Cost (₹)",
       title = "Cost vs QALYs for Three Interventions") +
  theme_minimal()
```

ggplot is your best friend for visualisations.

## Installing vs Loading Packages

R packages are like add-ons:

```{r}
#| label: packages
#| echo: true
#| eval: false

# Install a package (only need to do this once)
install.packages("ggplot2")

# Load a package (need to do this each session)
library(ggplot2)
```

**Analogy:** Installing is like downloading an app. Loading is like opening it.

## Pattern 1: Defining Model Parameters

```{r}
#| label: pattern-parameters
#| echo: true

# Transition probabilities (annual)
p_mild_to_moderate <- 0.15
p_moderate_to_severe <- 0.10

# Costs in ₹
c_mild <- 5000
c_moderate <- 25000
```

When you see this: **these are the model inputs — change these numbers to explore scenarios.**

## Pattern 2: A Loop Over Cycles

```{r}
#| label: pattern-loop
#| echo: true

n_cycles <- 10
annual_cost <- 5000
discount_rate <- 0.03
total_cost <- 0

for (cycle in 1:n_cycles) {
  discounted <- annual_cost / (1 + discount_rate)^cycle
  total_cost <- total_cost + discounted
}

cat("Total discounted cost:", round(total_cost, 0), "\n")
```

When you see a `for` loop: **the model repeats a calculation across time cycles.**

## Pattern 3: Creating a Plot

```{r}
#| label: pattern-plot
#| echo: true
#| eval: false

library(ggplot2)

# Prepare data
cycle_data <- data.frame(
  cycle = 1:10,
  discounted_cost = annual_cost / (1 + discount_rate)^(1:10)
)

# Create plot
ggplot(cycle_data, aes(x = cycle, y = discounted_cost)) +
  geom_col(fill = "steelblue") +
  labs(x = "Year", y = "Discounted Cost (₹)",
       title = "Effect of Discounting") +
  theme_minimal()
```

When you see `ggplot`: **a chart is being created.**

## The Read → Modify → Run → Interpret Workflow

```{r}
#| label: fig-workflow
#| eval: true
#| fig-cap: "The workshop workflow"

grViz("
digraph workflow {
  graph [rankdir=LR, bgcolor='transparent', fontname='Helvetica']
  node [shape=box, style='filled,rounded', fontname='Helvetica', fontsize=12]

  A [label='READ\nthe code\n(understand logic)', fillcolor='#4e79a7', fontcolor='white']
  B [label='MODIFY\na parameter\n(change a number)', fillcolor='#59a14f', fontcolor='white']
  C [label='RUN\nthe chunk\n(click play)', fillcolor='#f28e2b', fontcolor='white']
  D [label='INTERPRET\nthe output\n(what changed?)', fillcolor='#e15759', fontcolor='white']

  A -> B -> C -> D
  D -> B [style=dashed, label='Try again', fontsize=9]
}
")
```

## Working with .qmd Files

All workshop materials are in **Quarto (.qmd)** files:

- **Text** — explanations
- **Code chunks** — grey blocks with R code
- **Outputs** — results, tables, plots

. . .

**To work with a .qmd file:**

1. Open it in RStudio
2. Read the text explanations
3. Run code chunks one at a time (click the green play button ▶)
4. See the output appear below each chunk
5. Modify values and re-run

## Three Things to Remember

1. **Values are stored with `<-`**
   - Change them by editing the number

. . .

2. **Code runs top to bottom**
   - Run chunks in order

. . .

3. **If something breaks, read the error message**
   - It often tells you what went wrong

Everything else you can look up — or ask an AI tool.

## Try It Yourself

```{r}
#| label: try-it
#| echo: true

# YOUR TASK: Change these values and re-run
cost_A <- 3000
cost_B <- 1500
qaly_A <- 0.82
qaly_B <- 0.75

# Calculate ICER
icer <- (cost_A - cost_B) / (qaly_A - qaly_B)
cat("ICER = ₹", round(icer, 0), "per QALY gained\n")

# Is it cost-effective?
wtp <- 100000
if (icer < wtp) {
  cat("→ Cost-effective at WTP = ₹", format(wtp, big.mark=","), "\n")
} else {
  cat("→ NOT cost-effective\n")
}
```

If you ran this and changed the result — **you are ready for the rest of the workshop.**
