---
title: "Solution: DES vs BMS Decision Tree Exercise"
subtitle: "Completed answers with interpretation"
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
---

```{r}
#| label: setup
#| echo: false
#| message: false

library(ggplot2)
library(knitr)

# Model function (same as exercise)
run_des_bms_model <- function(
    cost_des_stent = 38933, cost_bms_stent = 10693,
    cost_procedure = 120000,
    cost_dapt_des = 8000, cost_dapt_bms = 3000,
    cost_followup = 5000,
    p_restenosis_des = 0.08, p_restenosis_bms = 0.20,
    p_mace_des_no_rest = 0.08, p_mace_bms_no_rest = 0.15,
    p_mace_des_rest = 0.15, p_mace_bms_rest = 0.25,
    p_mi_given_mace = 0.70,
    cost_repeat_revasc = 194000,
    cost_mi = 200000, cost_death = 50000,
    utility_well = 0.85, utility_restenosis = 0.78,
    utility_mi = 0.65,
    n_cohort = 1000, time_horizon = 5, discount_rate = 0.03
) {
  discount_factor <- sum(1 / (1 + discount_rate)^(0:(time_horizon - 1)))

  des_rest <- n_cohort * p_restenosis_des
  des_no_rest <- n_cohort * (1 - p_restenosis_des)
  des_mace <- des_rest * p_mace_des_rest + des_no_rest * p_mace_des_no_rest
  des_mi <- des_mace * p_mi_given_mace
  des_death <- des_mace * (1 - p_mi_given_mace)

  cost_des <- n_cohort * (cost_procedure + cost_des_stent) +
              n_cohort * cost_dapt_des +
              n_cohort * cost_followup * discount_factor +
              des_rest * cost_repeat_revasc +
              des_mi * cost_mi + des_death * cost_death

  qaly_des <- des_no_rest * (1 - p_mace_des_no_rest) * utility_well * time_horizon +
              des_no_rest * p_mace_des_no_rest * p_mi_given_mace *
                (utility_well * 2 + utility_mi * 3) +
              des_no_rest * p_mace_des_no_rest * (1 - p_mi_given_mace) *
                utility_well * (time_horizon / 2) +
              (des_rest * (1 - p_mace_des_rest)) *
                (utility_well * 0.5 + utility_restenosis * (time_horizon - 0.5)) +
              des_rest * p_mace_des_rest * p_mi_given_mace *
                (utility_well * 1 + utility_mi * 2) +
              des_rest * p_mace_des_rest * (1 - p_mi_given_mace) *
                utility_well * 1.5

  bms_rest <- n_cohort * p_restenosis_bms
  bms_no_rest <- n_cohort * (1 - p_restenosis_bms)
  bms_mace <- bms_rest * p_mace_bms_rest + bms_no_rest * p_mace_bms_no_rest
  bms_mi <- bms_mace * p_mi_given_mace
  bms_death <- bms_mace * (1 - p_mi_given_mace)

  cost_bms <- n_cohort * (cost_procedure + cost_bms_stent) +
              n_cohort * cost_dapt_bms +
              n_cohort * cost_followup * discount_factor +
              bms_rest * cost_repeat_revasc +
              bms_mi * cost_mi + bms_death * cost_death

  qaly_bms <- bms_no_rest * (1 - p_mace_bms_no_rest) * utility_well * time_horizon +
              bms_no_rest * p_mace_bms_no_rest * p_mi_given_mace *
                (utility_well * 2 + utility_mi * 3) +
              bms_no_rest * p_mace_bms_no_rest * (1 - p_mi_given_mace) *
                utility_well * (time_horizon / 2) +
              (bms_rest * (1 - p_mace_bms_rest)) *
                (utility_well * 0.5 + utility_restenosis * (time_horizon - 0.5)) +
              bms_rest * p_mace_bms_rest * p_mi_given_mace *
                (utility_well * 1 + utility_mi * 2) +
              bms_rest * p_mace_bms_rest * (1 - p_mi_given_mace) *
                utility_well * 1.5

  list(
    cost_des = cost_des, cost_bms = cost_bms,
    qaly_des = qaly_des, qaly_bms = qaly_bms,
    restenosis_des = des_rest, restenosis_bms = bms_rest,
    mace_des = des_mace, mace_bms = bms_mace,
    inc_cost = cost_des - cost_bms,
    inc_qaly = qaly_des - qaly_bms,
    icer = (cost_des - cost_bms) / (qaly_des - qaly_bms)
  )
}
```

## Task 1 Solution: Pre-Regulation Prices

```{r}
#| label: task1
#| echo: true

result_pre_reg <- run_des_bms_model(cost_des_stent = 120000, cost_bms_stent = 45000)
result_current <- run_des_bms_model()

cat("Pre-regulation (DES ₹1,20,000 / BMS ₹45,000):\n")
cat("  ICER: ₹", format(round(result_pre_reg$icer), big.mark = ","), "per QALY\n\n")
cat("Current NPPA ceiling (DES ₹38,933 / BMS ₹10,693):\n")
cat("  ICER: ₹", format(round(result_current$icer), big.mark = ","), "per QALY\n\n")
cat("Reduction in ICER:", round((1 - result_current$icer / result_pre_reg$icer) * 100), "%\n")
```

**Interpretation:** The NPPA price regulation dramatically improved the cost-effectiveness of DES. Before regulation, the large price gap between DES and BMS made the incremental cost high relative to the QALY gain, potentially pushing the ICER above affordability thresholds for India. After regulation, the price gap shrank substantially, making DES much more cost-effective. This is a powerful example of how pricing policy directly shapes HTA conclusions — the same clinical evidence, but a completely different economic answer. It also illustrates why HTA should not be a one-time exercise: when market conditions change, cost-effectiveness should be re-evaluated.

## Task 2 Solution: Varying Restenosis Rates

```{r}
#| label: task2
#| echo: true

result_a <- run_des_bms_model(p_restenosis_des = 0.03, p_restenosis_bms = 0.20)
result_b <- run_des_bms_model(p_restenosis_des = 0.08, p_restenosis_bms = 0.10)

cat("Scenario A — Large gap (DES 3% vs BMS 20%):\n")
cat("  ICER: ₹", format(round(result_a$icer), big.mark = ","), "per QALY\n")
cat("  Restenosis avoided:", result_a$restenosis_bms - result_a$restenosis_des, "\n\n")

cat("Scenario B — Narrow gap (DES 8% vs BMS 10%):\n")
cat("  ICER: ₹", format(round(result_b$icer), big.mark = ","), "per QALY\n")
cat("  Restenosis avoided:", result_b$restenosis_bms - result_b$restenosis_des, "\n")
```

**Interpretation:** It is the *difference* in restenosis rates that drives cost-effectiveness, not the absolute level. When the gap is large (Scenario A: 17 percentage points), DES provides substantial clinical benefit and the ICER is more favourable. When the gap narrows (Scenario B: only 2 percentage points), the incremental benefit shrinks but the incremental cost remains, pushing the ICER higher. This has practical implications: for patient subgroups where BMS performs relatively well (simple, short lesions in large vessels), the case for DES is weaker. For high-restenosis-risk patients (diabetics, long lesions, small vessels), DES becomes strongly cost-effective.

## Task 3 Solution: Threshold Analysis

```{r}
#| label: task3
#| echo: true
#| fig-cap: "Threshold analysis with annotations"

des_prices <- seq(10000, 150000, by = 5000)
icers <- numeric(length(des_prices))

for (i in seq_along(des_prices)) {
  res <- run_des_bms_model(cost_des_stent = des_prices[i])
  icers[i] <- res$icer
}

threshold_data <- data.frame(DES_Price = des_prices, ICER = icers)

# Find threshold price
wtp <- 170000
threshold_idx <- which(icers > wtp)[1]
threshold_price <- if (!is.na(threshold_idx)) des_prices[threshold_idx] else NA

ggplot(threshold_data, aes(x = DES_Price / 1000, y = ICER)) +
  geom_line(colour = "steelblue", linewidth = 1.2) +
  geom_hline(yintercept = wtp, linetype = "dashed", colour = "red") +
  geom_vline(xintercept = 38.933, linetype = "dotted", colour = "green4") +
  annotate("text", x = 120, y = wtp + 15000,
           label = "WTP = ₹1,70,000/QALY", colour = "red", size = 3) +
  annotate("text", x = 50, y = min(icers) + 10000,
           label = "NPPA ceiling\n₹38,933", colour = "green4", size = 3) +
  labs(x = "DES Stent Price (₹ thousands)", y = "ICER (₹ per QALY)",
       title = "Threshold Analysis: DES Price vs Cost-Effectiveness") +
  theme_minimal()

if (!is.na(threshold_price)) {
  cat("DES becomes NOT cost-effective at approximately ₹",
      format(threshold_price, big.mark = ","), "\n")
  cat("Current NPPA ceiling is ₹38,933 — well within the cost-effective range\n")
}
```

**Interpretation:** The current NPPA ceiling price for DES sits comfortably below the threshold at which DES would stop being cost-effective. There is substantial headroom — even if DES prices were to increase, the intervention would likely remain cost-effective. This type of threshold analysis is very useful for regulators: it tells the NPPA the maximum price at which DES provides value for money, helping to set evidence-informed price ceilings.

## Task 4 Solution: NNT Analysis

```{r}
#| label: task4
#| echo: true

result <- run_des_bms_model()

mace_rate_des <- result$mace_des / 1000
mace_rate_bms <- result$mace_bms / 1000
arr <- mace_rate_bms - mace_rate_des
nnt <- 1 / arr

cat("MACE rate with DES:", round(mace_rate_des * 100, 1), "%\n")
cat("MACE rate with BMS:", round(mace_rate_bms * 100, 1), "%\n")
cat("Absolute risk reduction:", round(arr * 100, 1), "percentage points\n")
cat("NNT:", round(nnt), "patients\n\n")

cost_per_mace_avoided <- result$inc_cost / (result$mace_bms - result$mace_des)
cat("Cost per MACE avoided: ₹", format(round(cost_per_mace_avoided), big.mark = ","), "\n")
```

**Interpretation:** The NNT tells clinicians and policymakers a concrete story: for every N patients who receive DES instead of BMS, one additional MACE event (heart attack or cardiac death) is prevented. This is often more intuitive than an ICER for clinical audiences. The cost per MACE avoided provides a complementary perspective — it tells the payer exactly how much additional investment is needed to prevent one serious cardiac event. Presenting results in multiple formats (ICER, NNT, cost per event avoided) is good HTA practice, as different decision-makers find different metrics compelling.

→ **Return to** the main module or **proceed to Day 2 sessions**.
