---
title: "Session 4: Therapeutic Decision Tree — DES vs BMS"
subtitle: "Drug-eluting stents vs bare metal stents for coronary artery disease in India"
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
---

## 1. The Clinical Question

Coronary artery disease (CAD) is a leading cause of mortality in India. Percutaneous coronary intervention (PCI) with stent placement is a standard treatment, and the choice between **drug-eluting stents (DES)** and **bare metal stents (BMS)** has major cost and outcome implications.

In 2017, India's NPPA capped stent prices — BMS at approximately ₹7,260 and DES at ₹29,600. As of 2025, revised ceilings stand at approximately ₹10,693 for BMS and ₹38,933 for DES. This dramatically changed the economics of stent choice in India.

**The HTA question:** At current Indian price ceilings, is DES cost-effective compared to BMS over a 5-year time horizon?

This is a **therapeutic decision tree** — comparing two treatment strategies with different costs and probabilities of downstream events.

## 2. Framing the Decision Problem — PICO

Before we touch any data, let us formalise the question using the **PICO framework**, extended for HTA.

```{r}
#| label: fig-pico
#| echo: false
#| fig-cap: "PICO framework extended for HTA decision modelling"

library(DiagrammeR)

grViz("
digraph pico {
  graph [rankdir=TB, bgcolor='transparent', fontname='Helvetica', nodesep=0.6, ranksep=0.5]
  node [fontname='Helvetica', fontsize=11, style='filled,rounded', shape=box]

  Q [label='Structured Decision Problem', fillcolor='#2c3e50', fontcolor='white', fontsize=13, width=3.5]

  P [label='P — Population\\nAdults with stable CAD\\nundergoing elective PCI in India', fillcolor='#3498db', fontcolor='white', width=3]
  I [label='I — Intervention\\nDrug-Eluting Stent (DES)\\n₹38,933 (NPPA 2025)', fillcolor='#27ae60', fontcolor='white', width=3]
  C [label='C — Comparator\\nBare Metal Stent (BMS)\\n₹10,693 (NPPA 2025)', fillcolor='#e67e22', fontcolor='white', width=3]
  O [label='O — Outcomes\\nCosts (₹), QALYs, ICER', fillcolor='#e74c3c', fontcolor='white', width=3]

  H [label='HTA Extensions', fillcolor='#8e44ad', fontcolor='white', fontsize=12, width=3.5]

  Per [label='Perspective\\nHealthcare payer', fillcolor='#9b59b6', fontcolor='white', width=2.5]
  TH [label='Time Horizon\\n5 years', fillcolor='#9b59b6', fontcolor='white', width=2.5]
  DR [label='Discounting\\n3% per annum', fillcolor='#9b59b6', fontcolor='white', width=2.5]
  WTP [label='Threshold\\n₹1.7 lakh/QALY\\n(1× GDP per capita)', fillcolor='#9b59b6', fontcolor='white', width=2.5]

  Q -> P
  Q -> I
  Q -> C
  Q -> O

  Q -> H [style=dashed]
  H -> Per
  H -> TH
  H -> DR
  H -> WTP
}
")
```

::: {.callout-tip}
## Why PICO matters for modelling
Every HTA model answers a *specific* question. The PICO forces us to be precise: which patients? which interventions? what are we measuring? The HTA extensions add the economic lens — whose costs count (perspective), how far we look (time horizon), how we value future outcomes (discounting), and when we say "worth it" (WTP threshold).

Notice that changing *any* element of the PICO changes the entire model. Younger patients? Different probabilities. Different country? Different costs and thresholds. This is why we build models in code — so we can adapt them.
:::


## 3. Conceptual Map — What Do We Need?

Now that we know *what* we're asking, let's brainstorm: **what information do we need to answer it?**

Before looking up a single number, think about the building blocks of a cost-effectiveness model.

```{mermaid}
%%| fig-cap: "Conceptual map: building blocks of a decision tree model"
mindmap
  root((DES vs BMS<br/>What do we need?))
    **Cohort**
      Who are we modelling?
      How many patients?
      How long do we follow them?
    **Probabilities**
      Does the procedure succeed?
      Does restenosis occur?
      Does a major event happen?
      What kind — MI or death?
    **Costs ₹**
      Stent itself
      PCI procedure
      Medications post-PCI
      Follow-up visits
      Repeat intervention if restenosis
      Managing MI or death
    **Utilities QALYs**
      Quality of life when well
      Quality after restenosis recovery
      Quality after surviving MI
      Death = 0
    **Framework**
      Time horizon
      Discount rate
      WTP threshold
      Perspective
```

::: {.callout-note}
## Think before you code
This is how HTA analysts work in practice. You sketch the model on a whiteboard *first*, identify what data you need, *then* go find the numbers. The mind map above is your shopping list — now let's fill in each category.
:::


## 4. Input Parameters — One Category at a Time

### 4a. The Cohort

Every model starts with: *who are we modelling, for how long, and what economic rules apply?*

```{r}
#| label: params-cohort
#| echo: true

# === COHORT & FRAMEWORK ===
n_cohort      <- 1000   # 1,000 patients undergoing PCI
time_horizon  <- 5      # years of follow-up
discount_rate <- 0.03   # 3% annual discount (years 2–5)
```

### 4b. Probabilities — What Can Happen?

These are the clinical evidence inputs. Each number comes from trials, registries, or meta-analyses.

```{r}
#| label: params-probabilities
#| echo: true

# === PROBABILITIES ===

# Procedural success
# Source: Indian PCI registry data
p_success_des <- 0.97
p_success_bms <- 0.96

# Restenosis within 1 year (requiring repeat revascularisation)
# Source: PMC comparative study (Indian data); international meta-analyses
p_restenosis_des <- 0.08   # DES: 5–10%
p_restenosis_bms <- 0.20   # BMS: 15–20%

# MACE at 5 years (MI or cardiac death)
# Among patients WITHOUT restenosis
p_mace_des_no_restenosis <- 0.08
p_mace_bms_no_restenosis <- 0.15

# Among patients WITH restenosis (after repeat revascularisation)
p_mace_des_restenosis <- 0.15
p_mace_bms_restenosis <- 0.25

# Among MACE events: what proportion is MI vs cardiac death?
p_mi_given_mace        <- 0.70   # 70% MI
p_cardiac_death_given_mace <- 0.30   # 30% cardiac death

# Proportion needing CABG vs repeat PCI among restenosis cases
p_cabg_if_restenosis <- 0.20
```

::: {.callout-tip}
## Where do these numbers come from?
In a real HTA submission, every probability needs a source citation, a justification for the chosen value, and ideally a range (for sensitivity analysis later). Notice how each parameter has a brief source comment — this is good practice.
:::

### 4c. Costs (₹) — What Does Each Event Cost?

```{r}
#| label: params-costs
#| echo: true

# === COSTS (Indian Rupees) ===

# Stent costs — NPPA revised ceiling prices 2025
cost_des_stent <- 38933
cost_bms_stent <- 10693

# Base PCI procedure cost (same for both arms)
# Source: PMJAY/CGHS package rates; hospital data
cost_procedure <- 120000

# Post-PCI medication costs (annual)
cost_dapt_des_annual <- 8000    # 12-month dual antiplatelet (DES)
cost_dapt_bms_annual <- 3000    # Shorter DAPT + single antiplatelet (BMS)
cost_followup_annual <- 5000    # Annual follow-up visits and tests

# Repeat revascularisation costs
cost_repeat_pci <- 180000       # Repeat PCI (more complex than index)
cost_cabg       <- 250000       # CABG if PCI not suitable

# MACE event costs
cost_mi_management <- 200000    # Acute MI management
cost_cardiac_death <- 50000     # Terminal care
```

### 4d. Utilities — How Does Each State *Feel*?

```{r}
#| label: params-utilities
#| echo: true

# === UTILITY WEIGHTS (0 = dead, 1 = perfect health) ===
# Source: Adapted from international literature (Sullivan et al.)
utility_well_post_pci        <- 0.85   # Stable post-PCI
utility_restenosis_recovered <- 0.78   # After repeat revascularisation
utility_post_mi              <- 0.65   # Post-MI, surviving
utility_death                <- 0.00   # Death
```

::: {.callout-warning}
## Indian Utility Data Gap
Indian-specific EQ-5D utility values for post-PCI patients are limited. The utility weights here are adapted from international literature. For a formal HTA submission to an Indian body, a local utility study or regionally validated values would strengthen the analysis. *Keep this caveat in mind — we'll revisit it in the thought exercises.*
:::


## 5. The Decision Tree — How Parameters Connect

Now we can see the structure. The diagram below maps directly onto the parameters you just defined — each branch uses a probability from Section 4b, and each terminal node will receive costs (4c) and utilities (4d).

```{r}
#| label: fig-des-bms-tree
#| echo: true
#| fig-cap: "Decision tree structure: DES vs BMS for coronary intervention"

library(DiagrammeR)

grViz("
digraph des_bms {
  graph [rankdir=LR, bgcolor='transparent', fontname='Helvetica', nodesep=0.4]
  node [fontname='Helvetica', fontsize=10]

  D [label='PCI STENT\nChoice', shape=square, style=filled, fillcolor='#4e79a7', fontcolor='white', width=1.2]

  DES [label='DES\n₹38,933', shape=box, style='filled,rounded', fillcolor='#4e79a7', fontcolor='white']
  BMS [label='BMS\n₹10,693', shape=box, style='filled,rounded', fillcolor='#e15759', fontcolor='white']

  # DES branch
  DS [label='Success\n(97%)', shape=circle, style=filled, fillcolor='#d4e6f1', fontsize=9]
  DF [label='Failure\n(3%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=9]
  DR [label='Restenosis\n(8%)', shape=circle, style=filled, fillcolor='#fdebd0', fontsize=9]
  DNR [label='No Restenosis\n(92%)', shape=circle, style=filled, fillcolor='#d5f5e3', fontsize=9]
  DM1 [label='MACE\n(15%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=9]
  DW1 [label='Well', shape=box, style='filled,rounded', fillcolor='#d5f5e3', fontsize=9]
  DM2 [label='MACE\n(8%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=9]
  DW2 [label='Well', shape=box, style='filled,rounded', fillcolor='#d5f5e3', fontsize=9]

  # BMS branch
  BS [label='Success\n(96%)', shape=circle, style=filled, fillcolor='#d4e6f1', fontsize=9]
  BF [label='Failure\n(4%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=9]
  BR [label='Restenosis\n(20%)', shape=circle, style=filled, fillcolor='#fdebd0', fontsize=9]
  BNR [label='No Restenosis\n(80%)', shape=circle, style=filled, fillcolor='#d5f5e3', fontsize=9]
  BM1 [label='MACE\n(25%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=9]
  BW1 [label='Well', shape=box, style='filled,rounded', fillcolor='#d5f5e3', fontsize=9]
  BM2 [label='MACE\n(15%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=9]
  BW2 [label='Well', shape=box, style='filled,rounded', fillcolor='#d5f5e3', fontsize=9]

  D -> DES
  D -> BMS

  DES -> DS
  DES -> DF
  DS -> DR
  DS -> DNR
  DR -> DM1
  DR -> DW1
  DNR -> DM2
  DNR -> DW2

  BMS -> BS
  BMS -> BF
  BS -> BR
  BS -> BNR
  BR -> BM1
  BR -> BW1
  BNR -> BM2
  BNR -> BW2
}
")
```

::: {.callout-note}
## Reading the tree
Each **square node** is a decision (we choose). Each **circle node** is a chance event (nature decides). Each **terminal box** is an outcome with associated cost and utility. The percentages on chance nodes come directly from the probability parameters you defined in Section 4b.
:::


## 6. Pathway Counts — Trace Patients Through the Tree

Before calculating any costs or QALYs, we count *how many patients end up where*. This is the foundation — the denominator that drives everything.

```{r}
#| label: pathway-counts
#| echo: true

library(knitr)

# ============================================================
# DES ARM — count patients at each node
# ============================================================
des_success       <- n_cohort * p_success_des
des_failure       <- n_cohort * (1 - p_success_des)     # → CABG

des_restenosis    <- des_success * p_restenosis_des
des_no_restenosis <- des_success * (1 - p_restenosis_des)

# MACE from restenosis pathway
des_mace_from_restenosis    <- des_restenosis * p_mace_des_restenosis
des_no_mace_from_restenosis <- des_restenosis * (1 - p_mace_des_restenosis)

# MACE from no-restenosis pathway
des_mace_no_restenosis    <- des_no_restenosis * p_mace_des_no_restenosis
des_no_mace_no_restenosis <- des_no_restenosis * (1 - p_mace_des_no_restenosis)

# Total MACE and breakdown
des_total_mace   <- des_mace_from_restenosis + des_mace_no_restenosis
des_mi           <- des_total_mace * p_mi_given_mace
des_cardiac_death <- des_total_mace * p_cardiac_death_given_mace

# ============================================================
# BMS ARM — count patients at each node
# ============================================================
bms_success       <- n_cohort * p_success_bms
bms_failure       <- n_cohort * (1 - p_success_bms)

bms_restenosis    <- bms_success * p_restenosis_bms
bms_no_restenosis <- bms_success * (1 - p_restenosis_bms)

bms_mace_from_restenosis    <- bms_restenosis * p_mace_bms_restenosis
bms_no_mace_from_restenosis <- bms_restenosis * (1 - p_mace_bms_restenosis)

bms_mace_no_restenosis    <- bms_no_restenosis * p_mace_bms_no_restenosis
bms_no_mace_no_restenosis <- bms_no_restenosis * (1 - p_mace_bms_no_restenosis)

bms_total_mace    <- bms_mace_from_restenosis + bms_mace_no_restenosis
bms_mi            <- bms_total_mace * p_mi_given_mace
bms_cardiac_death <- bms_total_mace * p_cardiac_death_given_mace

# --- Clean summary as a table ---
pathway_summary <- data.frame(
  Outcome = c("Successful PCI", "Procedural failure → CABG",
              "Restenosis (repeat intervention)", "No restenosis",
              "Total MACE", "— MI", "— Cardiac death",
              "Well (no complications)"),
  DES = c(des_success, des_failure,
          round(des_restenosis, 1), round(des_no_restenosis, 1),
          round(des_total_mace, 1), round(des_mi, 1), round(des_cardiac_death, 1),
          round(des_no_mace_no_restenosis, 1)),
  BMS = c(bms_success, bms_failure,
          round(bms_restenosis, 1), round(bms_no_restenosis, 1),
          round(bms_total_mace, 1), round(bms_mi, 1), round(bms_cardiac_death, 1),
          round(bms_no_mace_no_restenosis, 1))
)

kable(pathway_summary, col.names = c("Outcome", "DES (n)", "BMS (n)"),
      caption = "Patient flow per 1,000 patients at 5 years",
      align = c("l", "r", "r"))
```

```{r}
#| label: fig-patient-flow
#| echo: true
#| fig-cap: "Patient flow comparison: DES vs BMS per 1,000 patients"

library(ggplot2)

flow_data <- data.frame(
  Outcome = rep(c("Successful PCI\n(no complications)", "Restenosis\n(repeat intervention)",
                   "MI", "Cardiac Death"), 2),
  Stent = rep(c("DES", "BMS"), each = 4),
  Count = c(
    round(des_no_mace_no_restenosis), round(des_restenosis), round(des_mi), round(des_cardiac_death),
    round(bms_no_mace_no_restenosis), round(bms_restenosis), round(bms_mi), round(bms_cardiac_death)
  )
)

flow_data$Outcome <- factor(flow_data$Outcome,
  levels = c("Successful PCI\n(no complications)", "Restenosis\n(repeat intervention)",
             "MI", "Cardiac Death"))

ggplot(flow_data, aes(x = Outcome, y = Count, fill = Stent)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = Count), position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("DES" = "#4e79a7", "BMS" = "#e15759")) +
  labs(x = "", y = "Number of Patients (per 1,000)",
       title = "Patient Outcomes at 5 Years: DES vs BMS") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

::: {.callout-tip}
## Count first, calculate later
This is a deliberate discipline. By counting pathway populations *before* attaching costs or utilities, you can verify the tree logic independently. Do the numbers add up to 1,000? They should. If not, you have a bug in your tree.
:::


## 7. Cost Vectors — Attach Costs to Each Pathway

Now we multiply: **pathway count × unit cost** for each cost component. We build the total cost from pieces, so you can see exactly what drives the expense.

```{r}
#| label: cost-vectors
#| echo: true

# Discount factor for summing over years
discount_factor <- sum(1 / (1 + discount_rate)^(0:(time_horizon - 1)))

# ============================================================
# DES ARM — Cost components
# ============================================================

# 1. Initial PCI + stent (successful patients get stent; failures get CABG)
cost_initial_des <- des_success * (cost_procedure + cost_des_stent) +
                    des_failure * cost_cabg

# 2. Medications: Year 1 DAPT + annual follow-up (discounted)
cost_medication_des <- n_cohort * cost_dapt_des_annual +
                       n_cohort * cost_followup_annual * discount_factor

# 3. Restenosis management (weighted: 80% repeat PCI, 20% CABG)
cost_restenosis_des <- des_restenosis * (
  (1 - p_cabg_if_restenosis) * cost_repeat_pci +
  p_cabg_if_restenosis * cost_cabg
)

# 4. MACE management
cost_mace_des <- des_mi * cost_mi_management +
                 des_cardiac_death * cost_cardiac_death

# TOTAL
total_cost_des <- cost_initial_des + cost_medication_des +
                  cost_restenosis_des + cost_mace_des

# ============================================================
# BMS ARM — Cost components
# ============================================================
cost_initial_bms <- bms_success * (cost_procedure + cost_bms_stent) +
                    bms_failure * cost_cabg

cost_medication_bms <- n_cohort * cost_dapt_bms_annual +
                       n_cohort * cost_followup_annual * discount_factor

cost_restenosis_bms <- bms_restenosis * (
  (1 - p_cabg_if_restenosis) * cost_repeat_pci +
  p_cabg_if_restenosis * cost_cabg
)

cost_mace_bms <- bms_mi * cost_mi_management +
                 bms_cardiac_death * cost_cardiac_death

total_cost_bms <- cost_initial_bms + cost_medication_bms +
                  cost_restenosis_bms + cost_mace_bms

# --- Clean summary as a table ---
fmt <- function(x) paste0("₹", format(round(x), big.mark = ","))

cost_summary <- data.frame(
  Component = c("Initial PCI + Stent", "Medications (5 yr)",
                "Restenosis management", "MACE management",
                "**TOTAL (cohort)**", "**Per patient**"),
  DES = c(fmt(cost_initial_des), fmt(cost_medication_des),
          fmt(cost_restenosis_des), fmt(cost_mace_des),
          fmt(total_cost_des), fmt(total_cost_des / n_cohort)),
  BMS = c(fmt(cost_initial_bms), fmt(cost_medication_bms),
          fmt(cost_restenosis_bms), fmt(cost_mace_bms),
          fmt(total_cost_bms), fmt(total_cost_bms / n_cohort))
)

kable(cost_summary, col.names = c("Cost Component", "DES", "BMS"),
      caption = "5-year cost summary: DES vs BMS",
      align = c("l", "r", "r"))
```

```{r}
#| label: fig-cost-breakdown
#| echo: true
#| fig-cap: "5-year cost breakdown per patient: DES vs BMS"

library(tidyr)

cost_breakdown <- data.frame(
  Category = c("Initial PCI + Stent", "Medications (5yr)", "Restenosis Mgmt", "MACE Mgmt"),
  DES = c(cost_initial_des, cost_medication_des, cost_restenosis_des, cost_mace_des) / n_cohort,
  BMS = c(cost_initial_bms, cost_medication_bms, cost_restenosis_bms, cost_mace_bms) / n_cohort
)

cost_long <- pivot_longer(cost_breakdown, cols = c(DES, BMS),
                          names_to = "Stent", values_to = "Cost")

ggplot(cost_long, aes(x = Stent, y = Cost, fill = Category)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("Initial PCI + Stent" = "#4e79a7",
                                "Medications (5yr)" = "pink",
                                "Restenosis Mgmt" = "#f28e2b",
                                "MACE Mgmt" = "#e15759")) +
  labs(x = "Stent Type", y = "Cost per Patient (₹)",
       title = "5-Year Cost Breakdown: DES vs BMS",
       subtitle = "Indian NPPA ceiling prices (2025)") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

::: {.callout-note}
## The cost story
DES costs more upfront (₹28,240 more per stent), but look at the restenosis and MACE management bars — BMS generates far more downstream costs. Whether DES ends up cheaper *overall* depends on the balance between these components.
:::


## 8. QALY Vectors — Attach Utilities to Each Pathway

Same logic: **pathway count × utility weight × time**. Each patient group accumulates QALYs at a rate determined by their health state.

```{r}
#| label: qaly-vectors
#| echo: true

# ============================================================
# DES ARM — QALY components
# ============================================================

# Patients who are well (no restenosis, no MACE) — best outcome
qaly_des_well <- des_no_mace_no_restenosis * utility_well_post_pci * time_horizon

# Restenosis but recovered (no subsequent MACE)
# 6 months unwell pre-intervention, then recovered for remaining years
qaly_des_restenosis_ok <- des_no_mace_from_restenosis *
  (utility_well_post_pci * 0.5 + utility_restenosis_recovered * (time_horizon - 0.5))

# MI survivors
# ~2 years well before MI, then reduced utility for remaining years
qaly_des_mi <- des_mi *
  (utility_well_post_pci * 2 + utility_post_mi * (time_horizon - 2))

# Cardiac death (assumed at midpoint of follow-up)
qaly_des_death <- des_cardiac_death * utility_well_post_pci * (time_horizon / 2)

# Procedural failures (→ CABG, assumed similar to restenosis recovery)
qaly_des_failure <- des_failure * utility_restenosis_recovered * time_horizon

# TOTAL
total_qaly_des <- qaly_des_well + qaly_des_restenosis_ok +
                  qaly_des_mi + qaly_des_death + qaly_des_failure

# ============================================================
# BMS ARM — QALY components
# ============================================================
qaly_bms_well <- bms_no_mace_no_restenosis * utility_well_post_pci * time_horizon

qaly_bms_restenosis_ok <- bms_no_mace_from_restenosis *
  (utility_well_post_pci * 0.5 + utility_restenosis_recovered * (time_horizon - 0.5))

qaly_bms_mi <- bms_mi *
  (utility_well_post_pci * 2 + utility_post_mi * (time_horizon - 2))

qaly_bms_death <- bms_cardiac_death * utility_well_post_pci * (time_horizon / 2)

qaly_bms_failure <- bms_failure * utility_restenosis_recovered * time_horizon

total_qaly_bms <- qaly_bms_well + qaly_bms_restenosis_ok +
                  qaly_bms_mi + qaly_bms_death + qaly_bms_failure

# --- Clean summary as a table ---
qaly_summary <- data.frame(
  Component = c("Well (no complications)", "Restenosis recovered",
                "MI survivors", "Cardiac death", "Procedural failure",
                "**TOTAL (cohort)**", "**Per patient**"),
  DES = c(round(qaly_des_well, 1), round(qaly_des_restenosis_ok, 1),
          round(qaly_des_mi, 1), round(qaly_des_death, 1),
          round(qaly_des_failure, 1),
          round(total_qaly_des, 1), round(total_qaly_des / n_cohort, 3)),
  BMS = c(round(qaly_bms_well, 1), round(qaly_bms_restenosis_ok, 1),
          round(qaly_bms_mi, 1), round(qaly_bms_death, 1),
          round(qaly_bms_failure, 1),
          round(total_qaly_bms, 1), round(total_qaly_bms / n_cohort, 3))
)

kable(qaly_summary, col.names = c("Patient Group", "DES (QALYs)", "BMS (QALYs)"),
      caption = "5-year QALY summary: DES vs BMS",
      align = c("l", "r", "r"))
```


## 9. The ICER — Defining and Calculating

The **Incremental Cost-Effectiveness Ratio** is the core metric of HTA. Let's define it as a function — because we will reuse it many times.

```{r}
#| label: icer-function
#| echo: true

# === ICER FUNCTION ===
calc_icer <- function(cost_new, cost_old, qaly_new, qaly_old) {
  inc_cost <- cost_new - cost_old
  inc_qaly <- qaly_new - qaly_old
  icer     <- inc_cost / inc_qaly
  list(inc_cost = inc_cost, inc_qaly = inc_qaly, icer = icer)
}

# Apply to our results
result <- calc_icer(
  cost_new = total_cost_des, cost_old = total_cost_bms,
  qaly_new = total_qaly_des, qaly_old = total_qaly_bms
)

# --- Interpret against WTP thresholds ---
wtp_1gdp <- 170000    # ~1× GDP per capita India
wtp_3gdp <- 510000    # ~3× GDP per capita

interpretation <- if (result$icer < 0) {
  "DES is **DOMINANT** (less costly AND more effective)"
} else if (result$icer < wtp_1gdp) {
  "DES is **HIGHLY cost-effective** (ICER < 1× GDP per capita)"
} else if (result$icer < wtp_3gdp) {
  "DES is **cost-effective** (ICER < 3× GDP per capita)"
} else {
  "DES is **NOT cost-effective** at conventional WTP thresholds"
}

# Display as a clean summary table
icer_table <- data.frame(
  Metric = c("Incremental cost (DES − BMS)",
             "Incremental QALYs (DES − BMS)",
             "ICER",
             "Interpretation"),
  Value = c(paste0("₹", format(round(result$inc_cost), big.mark = ",")),
            round(result$inc_qaly, 2),
            paste0("₹", format(round(result$icer), big.mark = ","), " per QALY"),
            interpretation)
)

kable(icer_table, col.names = c("", ""),
      caption = "Incremental cost-effectiveness: DES vs BMS",
      align = c("l", "l"))
```

```{r}
#| label: fig-nnt
#| echo: true
#| fig-cap: "Number needed to treat: DES vs BMS"

# NNT — a clinical translation of the ICER
nnt_restenosis <- round(1 / (p_restenosis_bms - p_restenosis_des))
nnt_mace <- round(1 / ((bms_total_mace - des_total_mace) / n_cohort))

nnt_data <- data.frame(
  Outcome = c("Avoid 1\nRestenosis", "Avoid 1\nMACE Event"),
  NNT = c(nnt_restenosis, nnt_mace)
)

ggplot(nnt_data, aes(x = Outcome, y = NNT, fill = Outcome)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0("NNT = ", NNT)), vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("#4e79a7", "#e15759")) +
  labs(x = "", y = "Number Needed to Treat",
       title = "How Many Patients Need DES to Prevent One Event?",
       subtitle = "Compared to BMS over 5-year horizon") +
  theme_minimal() +
  expand_limits(y = c(0, max(nnt_data$NNT) * 1.2))
```


## 10. Visualising the Results

```{r}
#| label: fig-outcomes-comparison
#| echo: true
#| fig-cap: "Clinical outcomes per 1,000 patients"

outcomes <- data.frame(
  Outcome = rep(c("Restenosis", "MI", "Cardiac Death"), 2),
  Stent = rep(c("DES", "BMS"), each = 3),
  Count = c(round(des_restenosis), round(des_mi), round(des_cardiac_death),
            round(bms_restenosis), round(bms_mi), round(bms_cardiac_death))
)

ggplot(outcomes, aes(x = Outcome, y = Count, fill = Stent)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = Count), position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("DES" = "#4e79a7", "BMS" = "#e15759")) +
  labs(x = "", y = "Number of Patients (per 1,000)",
       title = "Clinical Outcomes: DES vs BMS at 5 Years") +
  theme_minimal() +
  theme(legend.position = "bottom")
```


## 11. What Should Keep You Up at Night

We have an ICER. The model says DES is cost-effective. **But how confident are you?**

Every number in this model is a *point estimate* — a single "best guess." In reality, each parameter carries uncertainty. The questions below expose why a single ICER is not enough, and why we need **sensitivity analysis** (our next sessions).

### But first — wrapping the model in a function

Look back at Sections 6–9. You wrote about 80 lines of R code to get from parameters to ICER. Now imagine you want to ask: "What if the restenosis rate were 12% instead of 20%?" Would you copy-paste those 80 lines and change one number? What about testing 10 different restenosis rates? 100?

This is where **functions** transform your workflow. A function wraps your entire model into a single reusable command. You change one input, call the function, and get the answer. No copy-paste, no errors from forgetting to update a line.

::: {.callout-tip}
## Functions = Reusable models
Remember from Session 2 (R Orientation): a function takes inputs, does something, and returns outputs. In HTA, your model *is* a function — it takes parameters and returns costs, QALYs, and an ICER. Wrapping it this way is not just good coding practice; it is *essential* for sensitivity analysis, where you will call this function hundreds or thousands of times with different inputs.
:::

```{r}
#| label: run-model-function
#| echo: true

# ============================================================
# THE MODEL AS A FUNCTION
# ============================================================
# Everything from Sections 6–9, wrapped so we can reuse it.
# Default values = our base case. Change any input to explore.

run_model <- function(p_rest_des = 0.08, p_rest_bms = 0.20,
                      c_des = 38933, c_bms = 10693,
                      u_well = 0.85, u_rest = 0.78, u_mi = 0.65) {

  des_s   <- n_cohort * p_success_des
  des_r   <- des_s * p_rest_des
  des_nr  <- des_s * (1 - p_rest_des)
  des_mac <- des_r * p_mace_des_restenosis + des_nr * p_mace_des_no_restenosis
  des_m   <- des_mac * p_mi_given_mace
  des_d   <- des_mac * p_cardiac_death_given_mace

  bms_s   <- n_cohort * p_success_bms
  bms_r   <- bms_s * p_rest_bms
  bms_nr  <- bms_s * (1 - p_rest_bms)
  bms_mac <- bms_r * p_mace_bms_restenosis + bms_nr * p_mace_bms_no_restenosis
  bms_m   <- bms_mac * p_mi_given_mace
  bms_d   <- bms_mac * p_cardiac_death_given_mace

  df <- sum(1 / (1 + discount_rate)^(0:(time_horizon - 1)))

  tc_des <- des_s * (cost_procedure + c_des) + (n_cohort - des_s) * cost_cabg +
            n_cohort * cost_dapt_des_annual + n_cohort * cost_followup_annual * df +
            des_r * ((1 - p_cabg_if_restenosis) * cost_repeat_pci + p_cabg_if_restenosis * cost_cabg) +
            des_m * cost_mi_management + des_d * cost_cardiac_death

  tc_bms <- bms_s * (cost_procedure + c_bms) + (n_cohort - bms_s) * cost_cabg +
            n_cohort * cost_dapt_bms_annual + n_cohort * cost_followup_annual * df +
            bms_r * ((1 - p_cabg_if_restenosis) * cost_repeat_pci + p_cabg_if_restenosis * cost_cabg) +
            bms_m * cost_mi_management + bms_d * cost_cardiac_death

  tq_des <- (des_nr - des_nr * p_mace_des_no_restenosis) * u_well * time_horizon +
            (des_r - des_r * p_mace_des_restenosis) * (u_well * 0.5 + u_rest * (time_horizon - 0.5)) +
            des_m * (u_well * 2 + u_mi * 3) +
            des_d * u_well * (time_horizon / 2) +
            (n_cohort - des_s) * u_rest * time_horizon

  tq_bms <- (bms_nr - bms_nr * p_mace_bms_no_restenosis) * u_well * time_horizon +
            (bms_r - bms_r * p_mace_bms_restenosis) * (u_well * 0.5 + u_rest * (time_horizon - 0.5)) +
            bms_m * (u_well * 2 + u_mi * 3) +
            bms_d * u_well * (time_horizon / 2) +
            (n_cohort - bms_s) * u_rest * time_horizon

  list(cost_des = tc_des, cost_bms = tc_bms,
       qaly_des = tq_des, qaly_bms = tq_bms,
       icer = (tc_des - tc_bms) / (tq_des - tq_bms))
}
```

Now we can explore "what if" questions with a single line of code. Watch how easy it becomes.

### Question 1: What if restenosis rates are different?

The restenosis rate differential is the biggest clinical driver. We used 8% for DES and 20% for BMS. But what if newer BMS perform better — say 12% restenosis?

```{r}
#| label: thought-q1
#| echo: true

# One line each — that's the power of a function
base <- run_model()                       # base case
alt  <- run_model(p_rest_bms = 0.12)      # what if BMS improves?

kable(data.frame(
  Scenario = c("Base case (DES 8% vs BMS 20%)",
               "BMS restenosis drops to 12%"),
  ICER = c(paste0("₹", format(round(base$icer), big.mark = ",")),
           paste0("₹", format(round(alt$icer), big.mark = ",")))
), col.names = c("Scenario", "ICER (₹/QALY)"),
  caption = "How restenosis rates change the conclusion",
  align = c("l", "r"))
```

::: {.callout-warning}
The ICER changes dramatically with one parameter. How confident are you in the 20% estimate? This is exactly what **Deterministic Sensitivity Analysis (DSA)** addresses — varying one parameter at a time to see which ones matter most.
:::

### Questions to carry forward

The restenosis example above is just the beginning. Here are questions we *cannot* answer with a single ICER — and each one points to a method we will learn in the next sessions:

**At what DES price does it stop being worth it?** NPPA ceilings can be revised. Costs rise with inflation. There must be a price beyond which DES is no longer cost-effective — but what is it? → *Threshold analysis / DSA*

**What if ALL uncertain parameters vary at once?** We used point estimates throughout. But every probability, cost, and utility has a range. What does the ICER look like when everything varies simultaneously? → *Probabilistic Sensitivity Analysis (PSA)*

**"How confident are you that DES is cost-effective?"** A policy maker won't accept a single number. They want a probability. You cannot answer this with one ICER — you need a distribution of ICERs. → *Cost-Effectiveness Acceptability Curve (CEAC)*

**What about Indian utility values?** We flagged that our utility weights come from international literature. If Indian patients experience different quality of life, the entire conclusion could shift. Is it worth investing in a local EQ-5D study? → *Value of Information (VOI) analysis*

::: {.callout-important}
## The bridge to next sessions
A single ICER is a *starting point*, not the answer. The real answer is: **how robust is this conclusion to the things we don't know?** That is what sensitivity analysis provides, and it is exactly what we will build next — using the `run_model()` function you just created.
:::


## Key References

- NPPA India. Revised ceiling prices for coronary stents (2025).
- Bangalore S et al. (2012). Short- and long-term outcomes with DES and BMS. *Circulation*.
- Kaul U et al. (2017). India and the coronary stent market. *Circulation*.
- PMC (2010). Comparative study of restenosis rates in BMS and DES — Indian data.
- NEJM (2016). Drug-eluting or bare-metal stents for coronary artery disease.
- Sullivan PW et al. (2011). Catalog of EQ-5D scores for the United Kingdom. *Medical Decision Making*.

→ **Next session:** We'll take this exact model and systematically vary every parameter — first one at a time (DSA), then all at once (PSA).
