---
title: "Session 4: Therapeutic Decision Tree — DES vs BMS"
subtitle: "Drug-eluting stents vs bare metal stents for coronary artery disease in India"
format:
  html:
    toc: true
    code-fold: show
    code-tools: true
---

## The Clinical Question

Coronary artery disease (CAD) is a leading cause of mortality in India. Percutaneous coronary intervention (PCI) with stent placement is a standard treatment, and the choice between **drug-eluting stents (DES)** and **bare metal stents (BMS)** has major cost and outcome implications.

In 2017, India's NPPA capped stent prices — BMS at approximately ₹7,260 and DES at ₹29,600. As of 2025, revised ceilings stand at approximately ₹10,693 for BMS and ₹38,933 for DES. This dramatically changed the economics of stent choice in India.

**The HTA question:** At current Indian price ceilings, is DES cost-effective compared to BMS over a 5-year time horizon?

This is a **therapeutic decision tree** — comparing two treatment strategies with different costs and probabilities of downstream events.

## The Decision Problem

We compare two strategies for a patient undergoing PCI for stable CAD:

1. **DES (Drug-Eluting Stent):** Higher initial cost, lower restenosis and MACE rates
2. **BMS (Bare Metal Stent):** Lower initial cost, higher restenosis and repeat revascularization rates

The decision tree branches on: procedural success → restenosis (requiring repeat revascularization) → major adverse cardiac events (MACE: MI or cardiac death).

## Model Parameters

```{r}
#| label: parameters
#| echo: true

# ============================================================
# MODEL PARAMETERS — DES vs BMS Decision Tree
# ============================================================

# --- Cohort ---
n_cohort <- 1000  # 1,000 patients undergoing PCI

# --- Stent Costs (₹) ---
# Source: NPPA revised ceiling prices 2025
cost_des_stent <- 38933     # DES ceiling price
cost_bms_stent <- 10693     # BMS ceiling price

# --- Procedure Costs (₹, excluding stent) ---
# Source: Estimated from Indian hospital data, PMJAY/CGHS package rates
# Includes catheterisation lab, consumables, hospital stay, medications
cost_procedure <- 120000    # Base PCI procedure cost (same for both)

# --- Post-PCI Medication Costs (annual, ₹) ---
# DES requires longer dual antiplatelet therapy (DAPT): 12 months
# BMS requires shorter DAPT: 1-3 months, then single antiplatelet
# Source: Indian cardiology practice guidelines
cost_dapt_des_annual <- 8000   # Clopidogrel + Aspirin for 12 months
cost_dapt_bms_annual <- 3000   # Shorter DAPT + single antiplatelet
cost_followup_annual <- 5000   # Annual follow-up visits, tests

# --- Procedural Success ---
# Source: Indian PCI registry data
p_success_des <- 0.97
p_success_bms <- 0.96

# --- Restenosis (within 1 year, requiring repeat revascularization) ---
# Source: PMC comparative study (Indian data); international meta-analyses
# BMS: 15-20% restenosis requiring intervention
# DES: 5-10% restenosis
p_restenosis_des <- 0.08
p_restenosis_bms <- 0.20

# --- Cost of Repeat Revascularization (₹) ---
# Repeat PCI or CABG
# Source: PMJAY package rates
cost_repeat_pci <- 180000    # Repeat PCI (more complex than index)
cost_cabg <- 250000          # CABG if PCI not suitable

# Proportion requiring CABG vs repeat PCI among restenosis cases
p_cabg_if_restenosis <- 0.20  # 20% need CABG, 80% get repeat PCI

# --- MACE at 5 years (MI, cardiac death) ---
# Among patients WITHOUT restenosis
# Source: Circulation meta-analysis; 5-year follow-up data
p_mace_des_no_restenosis <- 0.08    # 5-year MACE rate, DES, no restenosis
p_mace_bms_no_restenosis <- 0.15    # 5-year MACE rate, BMS, no restenosis

# Among patients WITH restenosis (after repeat revascularization)
p_mace_des_restenosis <- 0.15
p_mace_bms_restenosis <- 0.25

# --- MACE Outcomes ---
# Among those with MACE: proportion MI vs cardiac death
p_mi_given_mace <- 0.70        # 70% MI
p_cardiac_death_given_mace <- 0.30  # 30% cardiac death

# --- Costs of MACE Events (₹) ---
cost_mi_management <- 200000    # Acute MI management
cost_cardiac_death <- 50000     # Terminal care costs

# --- Utility Weights (5-year horizon) ---
# Source: Adapted from international literature (Sullivan et al.);
# No dedicated Indian EQ-5D data for post-PCI; disclaimer applies
utility_well_post_pci <- 0.85     # Stable post-PCI
utility_restenosis_recovered <- 0.78  # After repeat revascularization
utility_post_mi <- 0.65          # Post-MI, surviving
utility_death <- 0.00            # Death

# --- Discount Rate ---
discount_rate <- 0.03   # 3% annual, applied to years 2-5

# --- Time Horizon ---
time_horizon <- 5  # years
```

::: {.callout-note}
## Note on Indian Utility Data
Indian-specific EQ-5D utility values for post-PCI patients are limited. The utility weights used here are adapted from international literature. For a formal HTA submission to an Indian HTA body, a local utility study or regionally validated values would strengthen the analysis.
:::

## Building the Decision Tree

```{r}
#| label: fig-des-bms-tree
#| echo: true
#| fig-cap: "Decision tree structure: DES vs BMS for coronary intervention"

library(DiagrammeR)

grViz("
digraph des_bms {
  graph [rankdir=LR, bgcolor='transparent', fontname='Helvetica', nodesep=0.4]
  node [fontname='Helvetica', fontsize=10]

  D [label='PCI Stent\nChoice', shape=square, style=filled, fillcolor='#4e79a7', fontcolor='white', width=1.2]

  DES [label='DES\n₹38,933', shape=box, style='filled,rounded', fillcolor='#4e79a7', fontcolor='white']
  BMS [label='BMS\n₹10,693', shape=box, style='filled,rounded', fillcolor='#e15759', fontcolor='white']

  # DES branch
  DS [label='Success\n(97%)', shape=circle, style=filled, fillcolor='#d4e6f1', fontsize=9]
  DF [label='Failure\n(3%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=9]
  DR [label='Restenosis\n(8%)', shape=circle, style=filled, fillcolor='#fdebd0', fontsize=9]
  DNR [label='No Restenosis\n(92%)', shape=circle, style=filled, fillcolor='#d5f5e3', fontsize=9]
  DM1 [label='MACE\n(15%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=9]
  DW1 [label='Well', shape=box, style='filled,rounded', fillcolor='#d5f5e3', fontsize=9]
  DM2 [label='MACE\n(8%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=9]
  DW2 [label='Well', shape=box, style='filled,rounded', fillcolor='#d5f5e3', fontsize=9]

  # BMS branch
  BS [label='Success\n(96%)', shape=circle, style=filled, fillcolor='#d4e6f1', fontsize=9]
  BF [label='Failure\n(4%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=9]
  BR [label='Restenosis\n(20%)', shape=circle, style=filled, fillcolor='#fdebd0', fontsize=9]
  BNR [label='No Restenosis\n(80%)', shape=circle, style=filled, fillcolor='#d5f5e3', fontsize=9]
  BM1 [label='MACE\n(25%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=9]
  BW1 [label='Well', shape=box, style='filled,rounded', fillcolor='#d5f5e3', fontsize=9]
  BM2 [label='MACE\n(15%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=9]
  BW2 [label='Well', shape=box, style='filled,rounded', fillcolor='#d5f5e3', fontsize=9]

  D -> DES
  D -> BMS

  DES -> DS
  DES -> DF
  DS -> DR
  DS -> DNR
  DR -> DM1
  DR -> DW1
  DNR -> DM2
  DNR -> DW2

  BMS -> BS
  BMS -> BF
  BS -> BR
  BS -> BNR
  BR -> BM1
  BR -> BW1
  BNR -> BM2
  BNR -> BW2
}
")
```

### Step 1: Initial PCI and Stent Costs

```{r}
#| label: initial-costs
#| echo: true

# --- Total Initial PCI Cost per Patient ---
initial_cost_des <- cost_procedure + cost_des_stent
initial_cost_bms <- cost_procedure + cost_bms_stent

cat("Initial PCI cost per patient:\n")
cat("  DES: ₹", format(initial_cost_des, big.mark = ","), "\n")
cat("  BMS: ₹", format(initial_cost_bms, big.mark = ","), "\n")
cat("  Difference: ₹", format(initial_cost_des - initial_cost_bms, big.mark = ","), "\n")
```

### Step 2: Trace Through the Decision Tree

```{r}
#| label: tree-tracing
#| echo: true

# ============================================================
# DES ARM
# ============================================================

# Procedural outcomes
des_success <- n_cohort * p_success_des
des_failure <- n_cohort * (1 - p_success_des)  # Proceed to CABG

# Among successful DES
des_restenosis <- des_success * p_restenosis_des
des_no_restenosis <- des_success * (1 - p_restenosis_des)

# MACE outcomes
des_mace_from_restenosis <- des_restenosis * p_mace_des_restenosis
des_no_mace_from_restenosis <- des_restenosis * (1 - p_mace_des_restenosis)

des_mace_no_restenosis <- des_no_restenosis * p_mace_des_no_restenosis
des_no_mace_no_restenosis <- des_no_restenosis * (1 - p_mace_des_no_restenosis)

# MACE breakdown (MI vs death)
des_total_mace <- des_mace_from_restenosis + des_mace_no_restenosis
des_mi <- des_total_mace * p_mi_given_mace
des_cardiac_death <- des_total_mace * p_cardiac_death_given_mace

cat("=== DES Arm: Patient Flow ===\n")
cat("Successful PCI:", des_success, "\n")
cat("Procedural failure (→ CABG):", des_failure, "\n")
cat("Restenosis requiring intervention:", round(des_restenosis, 1), "\n")
cat("Total MACE at 5 years:", round(des_total_mace, 1), "\n")
cat("  - MI:", round(des_mi, 1), "\n")
cat("  - Cardiac death:", round(des_cardiac_death, 1), "\n")

# ============================================================
# BMS ARM
# ============================================================

bms_success <- n_cohort * p_success_bms
bms_failure <- n_cohort * (1 - p_success_bms)

bms_restenosis <- bms_success * p_restenosis_bms
bms_no_restenosis <- bms_success * (1 - p_restenosis_bms)

bms_mace_from_restenosis <- bms_restenosis * p_mace_bms_restenosis
bms_no_mace_from_restenosis <- bms_restenosis * (1 - p_mace_bms_restenosis)

bms_mace_no_restenosis <- bms_no_restenosis * p_mace_bms_no_restenosis
bms_no_mace_no_restenosis <- bms_no_restenosis * (1 - p_mace_bms_no_restenosis)

bms_total_mace <- bms_mace_from_restenosis + bms_mace_no_restenosis
bms_mi <- bms_total_mace * p_mi_given_mace
bms_cardiac_death <- bms_total_mace * p_cardiac_death_given_mace

cat("\n=== BMS Arm: Patient Flow ===\n")
cat("Successful PCI:", bms_success, "\n")
cat("Procedural failure (→ CABG):", bms_failure, "\n")
cat("Restenosis requiring intervention:", round(bms_restenosis, 1), "\n")
cat("Total MACE at 5 years:", round(bms_total_mace, 1), "\n")
cat("  - MI:", round(bms_mi, 1), "\n")
cat("  - Cardiac death:", round(bms_cardiac_death, 1), "\n")
```

```{r}
#| label: fig-patient-flow
#| echo: true
#| fig-cap: "Patient flow comparison: DES vs BMS per 1,000 patients"

library(ggplot2)

flow_data <- data.frame(
  Outcome = rep(c("Successful PCI\n(no complications)", "Restenosis\n(repeat intervention)",
                   "MI", "Cardiac Death"), 2),
  Stent = rep(c("DES", "BMS"), each = 4),
  Count = c(
    round(des_no_mace_no_restenosis), round(des_restenosis), round(des_mi), round(des_cardiac_death),
    round(bms_no_mace_no_restenosis), round(bms_restenosis), round(bms_mi), round(bms_cardiac_death)
  )
)

flow_data$Outcome <- factor(flow_data$Outcome,
  levels = c("Successful PCI\n(no complications)", "Restenosis\n(repeat intervention)",
             "MI", "Cardiac Death"))

ggplot(flow_data, aes(x = Outcome, y = Count, fill = Stent)) +
  geom_col(position = "dodge") +
  geom_text(aes(label = Count), position = position_dodge(width = 0.9), vjust = -0.5, size = 3) +
  scale_fill_manual(values = c("DES" = "#4e79a7", "BMS" = "#e15759")) +
  labs(x = "", y = "Number of Patients (per 1,000)",
       title = "Patient Outcomes at 5 Years: DES vs BMS") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Step 3: Calculate Total Costs

```{r}
#| label: total-costs
#| echo: true

# --- DES: Total Costs ---
# Initial PCI
cost_initial_des <- des_success * initial_cost_des + des_failure * cost_cabg

# Annual medication and follow-up (5 years, discounted)
discount_factor <- sum(1 / (1 + discount_rate)^(0:(time_horizon - 1)))
cost_annual_des <- n_cohort * (cost_dapt_des_annual + cost_followup_annual)
# Year 1 DAPT is higher; subsequent years are follow-up only
cost_medication_des <- n_cohort * cost_dapt_des_annual +  # Year 1
                       n_cohort * cost_followup_annual * discount_factor  # All years

# Restenosis management
cost_restenosis_des <- des_restenosis * (
  (1 - p_cabg_if_restenosis) * cost_repeat_pci +
  p_cabg_if_restenosis * cost_cabg
)

# MACE management
cost_mace_des <- des_mi * cost_mi_management + des_cardiac_death * cost_cardiac_death

# Total
total_cost_des <- cost_initial_des + cost_medication_des +
                  cost_restenosis_des + cost_mace_des

# --- BMS: Total Costs ---
cost_initial_bms <- bms_success * initial_cost_bms + bms_failure * cost_cabg

cost_medication_bms <- n_cohort * cost_dapt_bms_annual +
                       n_cohort * cost_followup_annual * discount_factor

cost_restenosis_bms <- bms_restenosis * (
  (1 - p_cabg_if_restenosis) * cost_repeat_pci +
  p_cabg_if_restenosis * cost_cabg
)

cost_mace_bms <- bms_mi * cost_mi_management + bms_cardiac_death * cost_cardiac_death

total_cost_bms <- cost_initial_bms + cost_medication_bms +
                  cost_restenosis_bms + cost_mace_bms

cat("=== Cost Summary (cohort of", n_cohort, "patients) ===\n\n")
cat("DES:\n")
cat("  Initial PCI: ₹", format(round(cost_initial_des), big.mark = ","), "\n")
cat("  Medications: ₹", format(round(cost_medication_des), big.mark = ","), "\n")
cat("  Restenosis mgmt: ₹", format(round(cost_restenosis_des), big.mark = ","), "\n")
cat("  MACE mgmt: ₹", format(round(cost_mace_des), big.mark = ","), "\n")
cat("  TOTAL: ₹", format(round(total_cost_des), big.mark = ","), "\n\n")

cat("BMS:\n")
cat("  Initial PCI: ₹", format(round(cost_initial_bms), big.mark = ","), "\n")
cat("  Medications: ₹", format(round(cost_medication_bms), big.mark = ","), "\n")
cat("  Restenosis mgmt: ₹", format(round(cost_restenosis_bms), big.mark = ","), "\n")
cat("  MACE mgmt: ₹", format(round(cost_mace_bms), big.mark = ","), "\n")
cat("  TOTAL: ₹", format(round(total_cost_bms), big.mark = ","), "\n")
```

### Step 4: Calculate QALYs

```{r}
#| label: qalys
#| echo: true

# --- DES: QALYs over 5 years ---
# Patients who are well (no restenosis, no MACE)
qaly_des_well <- des_no_mace_no_restenosis * utility_well_post_pci * time_horizon

# Patients with restenosis but recovered (no subsequent MACE)
qaly_des_restenosis_ok <- des_no_mace_from_restenosis *
                          (utility_well_post_pci * 0.5 +  # 6 months pre-intervention
                           utility_restenosis_recovered * (time_horizon - 0.5))

# Patients with MI (surviving)
qaly_des_mi <- des_mi * (utility_well_post_pci * 2 +  # ~2 years before MI
                          utility_post_mi * (time_horizon - 2))

# Patients with cardiac death (~midpoint)
qaly_des_death <- des_cardiac_death * utility_well_post_pci * (time_horizon / 2)

# Procedural failures (CABG — assumed similar utility trajectory)
qaly_des_failure <- des_failure * utility_restenosis_recovered * time_horizon

total_qaly_des <- qaly_des_well + qaly_des_restenosis_ok +
                  qaly_des_mi + qaly_des_death + qaly_des_failure

# --- BMS: QALYs ---
qaly_bms_well <- bms_no_mace_no_restenosis * utility_well_post_pci * time_horizon

qaly_bms_restenosis_ok <- bms_no_mace_from_restenosis *
                          (utility_well_post_pci * 0.5 +
                           utility_restenosis_recovered * (time_horizon - 0.5))

qaly_bms_mi <- bms_mi * (utility_well_post_pci * 2 +
                          utility_post_mi * (time_horizon - 2))

qaly_bms_death <- bms_cardiac_death * utility_well_post_pci * (time_horizon / 2)

qaly_bms_failure <- bms_failure * utility_restenosis_recovered * time_horizon

total_qaly_bms <- qaly_bms_well + qaly_bms_restenosis_ok +
                  qaly_bms_mi + qaly_bms_death + qaly_bms_failure

cat("=== QALY Summary ===\n")
cat("DES total QALYs:", round(total_qaly_des, 1), "\n")
cat("DES QALYs per patient:", round(total_qaly_des / n_cohort, 3), "\n\n")
cat("BMS total QALYs:", round(total_qaly_bms, 1), "\n")
cat("BMS QALYs per patient:", round(total_qaly_bms / n_cohort, 3), "\n")
```

### Step 5: ICER Calculation

```{r}
#| label: icer
#| echo: true

# --- Incremental Analysis: DES vs BMS ---
inc_cost <- total_cost_des - total_cost_bms
inc_qaly <- total_qaly_des - total_qaly_bms
icer <- inc_cost / inc_qaly

cat("=== Incremental Cost-Effectiveness: DES vs BMS ===\n\n")
cat("Incremental cost: ₹", format(round(inc_cost), big.mark = ","), "\n")
cat("Incremental QALYs:", round(inc_qaly, 2), "\n")
cat("ICER: ₹", format(round(icer), big.mark = ","), "per QALY gained\n\n")

# Interpret against WTP thresholds
wtp_1gdp <- 170000   # ~1x GDP per capita India (approximate)
wtp_3gdp <- 510000   # ~3x GDP per capita

if (icer < 0) {
  cat("→ DES is DOMINANT (less costly and more effective)\n")
} else if (icer < wtp_1gdp) {
  cat("→ DES is HIGHLY cost-effective (ICER < 1x GDP per capita)\n")
} else if (icer < wtp_3gdp) {
  cat("→ DES is cost-effective (ICER < 3x GDP per capita)\n")
} else {
  cat("→ DES is NOT cost-effective at conventional WTP thresholds\n")
}
```

```{r}
#| label: fig-nnt-visual
#| echo: true
#| fig-cap: "Number needed to treat: DES vs BMS"

# NNT for avoiding adverse events
nnt_restenosis <- round(1 / (p_restenosis_bms - p_restenosis_des))
nnt_mace <- round(1 / ((bms_total_mace - des_total_mace) / n_cohort))

nnt_data <- data.frame(
  Outcome = c("Avoid 1\nRestenosis", "Avoid 1\nMACE Event"),
  NNT = c(nnt_restenosis, nnt_mace)
)

ggplot(nnt_data, aes(x = Outcome, y = NNT, fill = Outcome)) +
  geom_col(show.legend = FALSE) +
  geom_text(aes(label = paste0("NNT = ", NNT)), vjust = -0.5, size = 4, fontface = "bold") +
  scale_fill_manual(values = c("#4e79a7", "#e15759")) +
  labs(x = "", y = "Number Needed to Treat",
       title = "How Many Patients Need DES to Prevent One Event?",
       subtitle = "Compared to BMS over 5-year horizon") +
  theme_minimal() +
  expand_limits(y = c(0, max(nnt_data$NNT) * 1.2))
```

## Visualising the Results

```{r}
#| label: cost-breakdown-plot
#| echo: true
#| fig-cap: "Cost breakdown comparison: DES vs BMS"

library(ggplot2)
library(tidyr)

cost_breakdown <- data.frame(
  Category = c("Initial PCI", "Medications (5yr)", "Restenosis Mgmt", "MACE Mgmt"),
  DES = c(cost_initial_des, cost_medication_des, cost_restenosis_des, cost_mace_des) / n_cohort,
  BMS = c(cost_initial_bms, cost_medication_bms, cost_restenosis_bms, cost_mace_bms) / n_cohort
)

cost_long <- pivot_longer(cost_breakdown, cols = c(DES, BMS),
                          names_to = "Stent", values_to = "Cost")

ggplot(cost_long, aes(x = Stent, y = Cost, fill = Category)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("Initial PCI" = "#4e79a7",
                                "Medications (5yr)" = "#59a14f",
                                "Restenosis Mgmt" = "#f28e2b",
                                "MACE Mgmt" = "#e15759")) +
  labs(
    x = "Stent Type", y = "Cost per Patient (₹)",
    title = "5-Year Cost Breakdown: DES vs BMS",
    subtitle = "Indian NPPA ceiling prices (2025)"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r}
#| label: outcomes-comparison
#| echo: true
#| fig-cap: "Clinical outcomes per 1,000 patients"

outcomes <- data.frame(
  Outcome = rep(c("Restenosis", "MI", "Cardiac Death"), 2),
  Stent = rep(c("DES", "BMS"), each = 3),
  Count = c(round(des_restenosis), round(des_mi), round(des_cardiac_death),
            round(bms_restenosis), round(bms_mi), round(bms_cardiac_death))
)

ggplot(outcomes, aes(x = Outcome, y = Count, fill = Stent)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c("DES" = "#4e79a7", "BMS" = "#e15759")) +
  labs(
    x = "", y = "Number of Patients (per 1,000)",
    title = "Clinical Outcomes: DES vs BMS at 5 Years"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## The Key Insight

Notice what this analysis reveals: DES costs more upfront (₹28,240 more per stent), but **saves money downstream** through fewer repeat revascularisations and fewer MACE events. Whether DES ends up cost-effective, cost-saving, or too expensive depends heavily on the restenosis rate differential and the cost of managing complications.

This is exactly the kind of analysis where R shines — you can instantly see how changing one parameter (like the stent price, or the restenosis rate) shifts the conclusion. In Excel, this would require careful re-wiring of formulas across multiple sheets.

## Key References

- NPPA India. Revised ceiling prices for coronary stents (2025).
- Bangalore S et al. (2012). Short- and long-term outcomes with DES and BMS. *Circulation*.
- Kaul U et al. (2017). India and the coronary stent market. *Circulation*.
- PMC (2010). Comparative study of restenosis rates in BMS and DES — Indian data.
- NEJM (2016). Drug-eluting or bare-metal stents for coronary artery disease.
- JAHA (2020). DES vs BMS for LAD/LM revascularisation.

→ **Next:** Open `des-bms-exercise.qmd` to explore how parameter changes affect the results.
