---
title: "DES vs BMS Therapeutic Decision Tree"
subtitle: "Drug-eluting vs bare metal stents for coronary artery disease"
author: "R for HTA (Basics) — Workshop 2026"
institute: "RRC-HTA, AIIMS Bhopal | HTAIn, DHR"
format:
  revealjs:
    theme: [default, ../../custom-theme.scss]
    slide-number: true
    footer: "R for HTA (Basics) — RRC-HTA, AIIMS Bhopal"
    logo: ""
    transition: fade
    width: 1280
    height: 720
    code-fold: true
    code-overflow: wrap
    highlight-style: github
    fig-align: center
execute:
  echo: true
  eval: false
  warning: false
  message: false
---

## The Clinical Question

Coronary artery disease (CAD) is a leading cause of mortality in India.

. . .

Percutaneous coronary intervention (PCI) with stent placement is standard.

. . .

**The choice: Drug-eluting stents (DES) vs bare metal stents (BMS)** has major cost and outcome implications.

## The Indian Context

**NPPA Price Ceilings (2025):**

- BMS: ~₹10,693
- DES: ~₹38,933

. . .

**The HTA Question**

At current Indian price ceilings, is **DES cost-effective compared to BMS** over 5 years?

## DES vs BMS: The Tradeoff

**DES (Drug-Eluting Stent)**
- Higher initial cost
- Lower restenosis rate (fewer repeat procedures)
- Lower MACE rate (fewer MI/deaths)

. . .

**BMS (Bare Metal Stent)**
- Lower initial cost
- Higher restenosis rate (15–20% need repeat intervention)
- Higher MACE rate

. . .

**The question:** Do DES's downstream savings outweigh the upfront cost?

## Key Parameters

```{r}
#| label: fig-params-des-bms
#| echo: false
#| eval: true

library(knitr)
params <- data.frame(
  Parameter = c("DES stent cost", "BMS stent cost", "Procedure cost",
                 "DES restenosis", "BMS restenosis", "DES MACE (5yr)", "BMS MACE (5yr)"),
  Value = c("₹38,933", "₹10,693", "₹120,000 (both)",
            "8%", "20%", "8%", "15%")
)
kable(params, col.names = c("Parameter", "Value"))
```

All **sourced from Indian data and NPPA 2025 guidelines**.

## The Decision Tree Structure

```{r}
#| label: fig-des-bms-tree
#| eval: true
#| fig-cap: "DES vs BMS decision tree for PCI"

library(DiagrammeR)

grViz("
digraph des_bms {
  graph [rankdir=LR, bgcolor='transparent', fontname='Helvetica', nodesep=0.3]
  node [fontname='Helvetica', fontsize=9]

  D [label='PCI Stent\nChoice', shape=square, style=filled, fillcolor='#4e79a7', fontcolor='white']

  DES [label='DES\n₹38,933', shape=box, style='filled,rounded', fillcolor='#4e79a7', fontcolor='white']
  BMS [label='BMS\n₹10,693', shape=box, style='filled,rounded', fillcolor='#e15759', fontcolor='white']

  DS [label='Success\n(97%)', shape=circle, style=filled, fillcolor='#d4e6f1', fontsize=8]
  DR [label='Restenosis\n(8%)', shape=circle, style=filled, fillcolor='#fdebd0', fontsize=8]
  DNR [label='No Restenosis\n(92%)', shape=circle, style=filled, fillcolor='#d5f5e3', fontsize=8]
  DM [label='MACE\n(8%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=8]
  DW [label='Well', shape=box, style='filled,rounded', fillcolor='#d5f5e3', fontsize=8]

  BS [label='Success\n(96%)', shape=circle, style=filled, fillcolor='#d4e6f1', fontsize=8]
  BR [label='Restenosis\n(20%)', shape=circle, style=filled, fillcolor='#fdebd0', fontsize=8]
  BNR [label='No Restenosis\n(80%)', shape=circle, style=filled, fillcolor='#d5f5e3', fontsize=8]
  BM [label='MACE\n(15%)', shape=box, style='filled,rounded', fillcolor='#f5b7b1', fontsize=8]
  BW [label='Well', shape=box, style='filled,rounded', fillcolor='#d5f5e3', fontsize=8]

  D -> DES
  D -> BMS

  DES -> DS
  DS -> DR
  DS -> DNR
  DR -> DM
  DR -> DW
  DNR -> DM
  DNR -> DW

  BMS -> BS
  BS -> BR
  BS -> BNR
  BR -> BM
  BR -> BW
  BNR -> BM
  BNR -> BW
}
")
```

## Initial PCI Cost

```{r}
#| label: initial-costs
#| echo: true
#| eval: false

# Total initial PCI cost per patient
initial_cost_des <- cost_procedure + cost_des_stent  # ₹120,000 + ₹38,933
initial_cost_bms <- cost_procedure + cost_bms_stent  # ₹120,000 + ₹10,693

# Cost difference
upfront_difference <- initial_cost_des - initial_cost_bms
# ₹28,240 per stent
```

**Upfront cost:** DES is ~₹28,240 more expensive per patient.

The question: **Does this pay for itself downstream?**

## Tree Tracing: DES Arm

```{r}
#| label: tree-tracing-des
#| echo: true
#| eval: false

# Per 1,000 patients
des_success <- 1000 * 0.97           # 970 patients
des_failure <- 1000 * 0.03           # 30 patients (CABG)

des_restenosis <- des_success * 0.08    # 77 restenosis
des_no_restenosis <- des_success * 0.92 # 893 no restenosis

# MACE among those with restenosis
des_mace_restenosis <- 77 * 0.15     # 12 MI/cardiac death

# MACE among those without restenosis
des_mace_no_restenosis <- 893 * 0.08 # 71 MI/cardiac death

des_total_mace <- 12 + 71            # 83 total MACE events
```

## Tree Tracing: BMS Arm

```{r}
#| label: tree-tracing-bms
#| echo: true
#| eval: false

# Per 1,000 patients
bms_success <- 1000 * 0.96           # 960 patients
bms_failure <- 1000 * 0.04           # 40 patients

bms_restenosis <- bms_success * 0.20    # 192 restenosis (much higher!)
bms_no_restenosis <- bms_success * 0.80 # 768 no restenosis

# MACE outcomes
bms_mace_restenosis <- 192 * 0.25    # 48 MI/cardiac death
bms_mace_no_restenosis <- 768 * 0.15 # 115 MI/cardiac death

bms_total_mace <- 48 + 115           # 163 total MACE events
```

## Patient Flow Comparison

Over 5 years per 1,000 patients:

**DES:**
- Successful with no complications: ~893
- Restenosis requiring intervention: ~77
- MACE (MI or cardiac death): ~83

. . .

**BMS:**
- Successful with no complications: ~768
- Restenosis requiring intervention: ~192 (2.5x higher!)
- MACE: ~163 (2x higher!)

## Cost Calculation: DES

```{r}
#| label: costs-des
#| echo: true
#| eval: false

# Initial PCI
cost_initial_des <- des_success * initial_cost_des +
                    des_failure * cost_cabg

# Medications (5 years)
cost_medication_des <- n_cohort * cost_dapt_des_annual +  # Year 1
                      n_cohort * cost_followup_annual * 4  # Years 2-5

# Restenosis management (repeat PCI or CABG)
cost_restenosis_des <- des_restenosis * (
  0.80 * cost_repeat_pci +     # 80% repeat PCI
  0.20 * cost_cabg              # 20% CABG
)

# MACE management
cost_mace_des <- des_mi * cost_mi_management +
                 des_cardiac_death * cost_cardiac_death

total_cost_des <- cost_initial_des + cost_medication_des +
                  cost_restenosis_des + cost_mace_des
```

## Cost Calculation: BMS

```{r}
#| label: costs-bms
#| echo: true
#| eval: false

# Initial PCI (lower stent cost)
cost_initial_bms <- bms_success * initial_cost_bms +
                    bms_failure * cost_cabg

# Medications (shorter DAPT)
cost_medication_bms <- n_cohort * cost_dapt_bms_annual +
                       n_cohort * cost_followup_annual * 4

# Restenosis management (higher rate!)
cost_restenosis_bms <- bms_restenosis * (
  0.80 * cost_repeat_pci +
  0.20 * cost_cabg
)

# MACE management (higher rate!)
cost_mace_bms <- bms_mi * cost_mi_management +
                 bms_cardiac_death * cost_cardiac_death

total_cost_bms <- cost_initial_bms + cost_medication_bms +
                  cost_restenosis_bms + cost_mace_bms
```

## QALY Calculation

```{r}
#| label: qalys
#| echo: true
#| eval: false

# DES: QALYs over 5 years
qaly_des_well <- 893 * 0.85 * 5         # No complications
qaly_des_restenosis <- 77 * (0.78 * 4)  # After intervention
qaly_des_mace <- 83 * (0.65 * 3)        # Post-MI survival
qaly_des_death <- (83 * 0.3) * 0         # Death contributes 0 QALYs

total_qaly_des <- qaly_des_well + qaly_des_restenosis + qaly_des_mace

# BMS: QALYs (similar structure, with higher MACE)
# ...similar calculations but with higher adverse event rates
```

Higher MACE in BMS arm → Lower QALYs in BMS.

## ICER Result

```{r}
#| label: icer
#| echo: true
#| eval: false

inc_cost <- total_cost_des - total_cost_bms
inc_qaly <- total_qaly_des - total_qaly_bms
icer <- inc_cost / inc_qaly

# 4-quadrant interpretation (always check signs!)
if (inc_cost < 0 & inc_qaly > 0) {
  cat("DES is DOMINANT (cheaper AND better)")
} else if (inc_cost > 0 & inc_qaly < 0) {
  cat("DES is DOMINATED")
} else if (inc_cost > 0 & inc_qaly > 0) {
  if (icer < 170000) cat("Cost-effective") else cat("Not CE")
} else {
  cat("Trade-off")
}
```

## Net Monetary Benefit (NMB)

Recall from Session 3: **NMB = WTP × QALYs − Cost**

. . .

```{r}
#| echo: true
#| eval: false

nmb_des <- wtp * total_qaly_des - total_cost_des
nmb_bms <- wtp * total_qaly_bms - total_cost_bms
inc_nmb <- nmb_des - nmb_bms
# Positive ΔNMB → adopt DES
```

. . .

If DES is **dominant** (negative ΔCost, positive ΔQALYs), ΔNMB is always positive — no ambiguity.

## Number Needed to Treat (NNT)

**How many patients need DES to prevent one adverse event?**

```{r}
#| label: nnt
#| echo: true
#| eval: false

# Prevent one restenosis event
nnt_restenosis <- 1 / (0.20 - 0.08)  # NNT = 8.3

# Prevent one MACE event
nnt_mace <- 1 / ((163 - 83) / 1000)  # NNT = 12.5

cat("NNT to prevent 1 restenosis:", round(nnt_restenosis, 1), "\n")
cat("NNT to prevent 1 MACE:", round(nnt_mace, 1), "\n")
```

**For every 8-13 DES stents placed instead of BMS, you prevent one adverse event.**

## The Key Insight

DES costs more upfront (₹28,240 per stent).

. . .

**BUT:**
- Fewer repeat revascularisations
- Fewer MI events
- Fewer cardiac deaths

. . .

**Downstream savings may outweigh upfront cost** — depending on:
- Restenosis rate differential
- Cost of managing complications
- Strength of evidence from Indian data

## Therapeutic Decision Trees

This model structure applies to **any treatment choice**:

- **Drug A vs Drug B** (cost vs efficacy)
- **Surgery vs Medical** (invasive vs conservative)
- **Screening intensity** (more tests vs fewer tests)

. . .

Each branches on:
1. Procedural success/failure
2. Complication rates
3. Long-term outcomes (MACE, relapse, etc.)

## Why R Excels Here

**In Excel:** If NPPA revises DES price from ₹38,933 to ₹35,000, you manually update cells and pray you didn't break formulas.

. . .

**In R:** Change one number at the top → entire model re-calculates in milliseconds.

. . .

This is essential for **rapid scenario analysis** in HTA policy work.

## Key Takeaway

A therapeutic decision tree is **another reusable template**:

- **Treatment choice** (two or more strategies)
- **Probability branches** (success, complication, outcome rates)
- **Cost & QALY tracking** through each pathway
- **ICER with 4-quadrant interpretation** + **NMB**

. . .

Just like diagnostics, but for **treatment and intervention** decisions.

## Bonus: rdecision Package

The same DES vs BMS model rebuilt with `rdecision`:

- Object-oriented tree with `DecisionNode`, `ChanceNode`, `LeafNode`
- Built-in **tornado DSA** and **PSA** (1,000 iterations)
- **CE plane + CEAC** for decision uncertainty

. . .

→ See the **Session 4 Bonus** page on the workshop website.
